---
title: "R Notebook"
output: html_notebook
---
MiniBulk lineage barcode processing pipeline (Marthe: Scripts MiniBulkRNA_DRAG_PreProcess & MiniBulkRNA_DRAG_Figure):

Due to the library design, read 1 contained the UMI sequence and read 2 the lineage barcode. We used awk to extract these sequences from paired reads and to generate hybrid fused reads combining both identifiers.

Barcode sequences and count matrices for each PCR reaction were obtained using the CellBarcode package (ref). Barcode extraction was performed with the following regular expression:
AACGCAGAGT(.{8})GGG.*CAGTAGGTCTACTCTCGAGCTTGCTACGGT([ACGT]+)TTCGATGACCTCGAGGAGGT 
where:
  Capture group 1 corresponds to the 8 bp UMI sequence
  Capture group 2 corresponds to the lineage barcode

Barcode filtering was carried out through four successive steps to minimize PCR and sequencing artifacts. Initially, 64,264 unique sequences were detected before filtering.

QC1 - PCR efficiency: For each sequencing run, the distribution of barcode read counts per library was examined. The first local minimum preceding the main peak of the distribution was identified and used as a run-specific threshold to distinguish true signal from background noise. Libraries with read counts below this threshold were considered low quality and excluded from further analyses. Using this approach, 363 libraries (85% of all libraries) were retained for downstream analyses, corresponding to a total of 63,794 barcodes remaining.

QC2 - Detection threshold: We used technical replicates to calculate the evolution of spearman correlation between replicates in function of a minimum count detection threshold, to identify the threshold of detection for this experiment. Here it was 5 counts. Barcodes supported by fewer than 5 reads per library (3.668 barcodes remaining).

Technical replicates: When technical replicates were available, only barcode/UMI pairs detected in both replicates were retained. Median read counts were  used for subsequent analyses (3,293 barcodes remaining).

QC3 - PCR error removal: Within each library, barcodes differing by a single nucleotide substitution (computed using Levenshtein distance with weights: deletion = 0.1, insertion = 0.1, transposition = 0.1, substitution = 1) were filtered. The most abundant sequence (in terms of UMI number) was kept, or one was chosen at random in case of equal UMI numbers  (1,588 barcodes remaining).

QC4 - high-probability barcode removal: As for the single-cell dataset described earlier, the probability of generation (Pgen) for each lineage barcode was computed using the Igor pipeline. A Pgen threshold was determined based on the fraction of barcodes shared between mice: the 10-4.5 was retained as it allow to have ~94% of unique sequence, resulting in 549 high confidence barcodes retained. 


# Intro



```{r}
# -------------------------------------------------------------- 
# sample display order
# --------------------------------------------------------------
x_order <- c("9891-G4Gl_Early","9891-D4Gl_Early","9892-G4Gl_Early","9892-D4Gl_Early",
             "9873-G4Gl_Early","9873-D4Gl_Early","9873-G3Gl_Early","9873-D3Gl_Early",
             "9826-G4Gl_Early","9826-D4Gl_Early","9784-G4Gl_Early","9784-D4Gl_Early",
             "9757-G4Gl_Early","9757-D4Gl_Early","9757-G3Gl_Early","9757-D3Gl_Early",
             
             "9660-G4Gl_Early", "9660-D4Gl_Early", "9601-G4Gl_Early", "9601-D4Gl_Early", "9568-G4Gl_Early",
             "9568-D2Gl_Early", "9052Jx", "9052Tm", "8943Jx", "8943Tm", "9341Gl_Late", "8988Jx", "8988Tm", 
             "9424Jx", "8962Jx", "8962Tm", "9423Jx", "8903Jx")

x_order_ <- c("9891-G4_Gl_Early","9891-D4_Gl_Early","9892-G4_Gl_Early","9892-D4_Gl_Early",
              "9660-G4_Gl_Early", "9660-D4_Gl_Early",
              
             "9873-G4_Gl_Early","9873-D4_Gl_Early","9873-G3_Gl_Early","9873-D3_Gl_Early",
             "9601-G4_Gl_Early", "9601-D4_Gl_Early", 
             
             "9826-G4_Gl_Early","9826-D4_Gl_Early",
              "9568-G4_Gl_Early", "9568-D2_Gl_Early",
             "9784-G4_Gl_Early","9784-D4_Gl_Early",
             "9757-G4_Gl_Early","9757-D4_Gl_Early","9757-G3_Gl_Early","9757-D3_Gl_Early",
             
              "9052_Jx", "9052_Tm", "8943_Jx", "8943_Tm", "9341_Gl_Late", "8988_Jx", "8988_Tm", 
             "9424_Jx", "8962_Jx", "8962_Tm", "9423_Jx", "8903_Jx")

# -------------------------------------------------------------- 
# Color definition 
# --------------------------------------------------------------

cols_fate <- c("basal"  = "#236ec4", 
               "LP"     = "#a01858",      
               "HS"     = "#f9c40c",  
               "Alv"    = "#F7DCAD")


cols_Group <- c("EARLY" = "#C0D69F", 
                "EARLY_2m" = "#C0D69F", 
                "EARLY_3m" = "#C0D69F", 
                "LATE"  = "#61980E",
                "Tm"    = "#841811")

cols_MonoGland <- c("8415_G2-3" = "#375708",  
                    "8718_D2-3"     = "#61980E",
                    "8964_G4"       = "#C0D69F")
```


LIBRARIES


```{r}
library(CellBarcode)


library(data.table)
library(tidyverse)
library(dplyr)
library(readxl)

library(Seurat)
library(patchwork)
library(DoubletFinder)

library(ggplot2)
library(gridExtra)
library(ggExtra)
library(viridis)
library(Matrix)
library(ggsci)
library(cowplot)
library(platetools)
library(scales)
library(gridExtra)
library(qs)
library(RColorBrewer)
library(ggpubr)

library(stringdist)
library(Biostrings)
library(msa)
library(seqinr)
library(ggmsa)
```

LIBRARIES


```{r}
# Output directory for figures and results
outputDir <- "C:/Users/Marthe Laisne/OneDrive - INSTITUT CURIE/m_laisne/Publications/2026_Laisne/Scripts/Output/MiniBulk_DRAG/"
```


# PreProcess - Do not run

To construct the XXX object

```{r}
# Path to raw FASTQ files (paired-end sequencing)
inputDir <- "C:/Users/Marthe Laisne/Desktop/FlashSeq_MultiOneCell/Data/MiniBulk_BCM1/Fastq_HybridR1-R2/"

fq_files <- dir(inputDir, full=TRUE)
samplename = gsub(inputDir, "", fq_files) %>% gsub("\\.HybridR1R2\\.fastq", "",.)


# Metadata describing seq samples (mouse, condition, replicate, etc.)
sampleplan_joint <- read.delim("C:/Users/Marthe Laisne/OneDrive - INSTITUT CURIE/m_laisne/Publications/2026_Laisne/Scripts/Output/MiniBulk_DRAG/metadata_minibulk.txt")

rownames(sampleplan_joint) <- sampleplan_joint$samplename

```


##Extract BCM

- Paired-end 2x150 bp reads
- Read1 contains UMI
- Read2 contains lineage barcode (VDJ-derived construct)

Reference construct (ideal, non-mutated VDJ region):
 V: *acctcctcgaggtcatcgaa*gtatcaagtccagtag 
 D: tctactatcgttacgac
 J-linker:gtagctactaccgtagcaag*ctcgagagtagacCTACTG*


length: 86nt

```{r}
# ================================================================
# BARCODE EXTRACTION (CellBarcode pipeline)
# ================================================================

# Extract UMI and barcode using regular expression:
# Capture group 1 = UMI (8 bp)
# Capture group 2 = lineage barcode sequence

bc_obj_fusion <- bc_extract(
  fq_files,  # fastq file
  pattern = "AACGCAGAGT(.{8})GGG.*CAGTAGGTCTACTCTCGAG([ACGT]+)TTCGATGACCTCGAGGAGGT", 
  pattern_type = c("UMI" = 1,"barcode" = 2), 
  sample_name = samplename,
  metadata = sampleplan_joint
)
# Initial depth correction (no filtering)
bc_obj_fusion <- bc_cure_depth(bc_obj_fusion, depth =0 , isUpdate = FALSE)

# QC filtering of barcode/UMI pairs:
# - depth filtering (10)
# - UMI deduplication (unique UMI assumption)
# - optional "fishing" of low-depth UMIs (disabled here)
bc_obj_fusion_cure <- bc_cure_depth(bc_obj_fusion, depth =10 , isUpdate = FALSE) %>%
  bc_cure_umi(depth = 1,  isUniqueUMI = TRUE, doFish = FALSE)
```




##Generate dowstream table objects


```{r}
# Clean barcode object and convert to data.frame
AllBCM1_R2umi <- bc_cleanBc(bc_obj_fusion_cure, isList=FALSE)

# Extract sample name prefix (remove everything after "/")
AllBCM1_R2umi$sample_name_short <- gsub("/.*", 
                                        "", 
                                        AllBCM1_R2umi$sample_nmae)

# Standardize column names
colnames(AllBCM1_R2umi) <- c("sample_l", "umi_seq" , "barcode_seq" ,"count", "samplename")

# Add sample metadata (mouse, condition, etc.)
AllBCM1_R2umi <- left_join(AllBCM1_R2umi, 
                           sampleplan_joint,
                           by = "samplename")

# Create a combined biological sample identifier
AllBCM1_R2umi$SampleBio <-  paste0(AllBCM1_R2umi$mouse, 
                                   "_", 
                                   AllBCM1_R2umi$TypeSample)


# --------------------------------------------------------------
# Build aggregated tables
# --------------------------------------------------------------

# 1) Sum of read counts per barcode per biological sample
BCM1Summary_Counts <- AllBCM1_R2umi %>% 
  group_by(SampleBio, barcode_seq) %>%
  summarise(counts = sum(count), .groups = "drop"
  ) 

# 2) Number of distinct UMIs per barcode per biological sample
BCM1Summary_UMI <- AllBCM1_R2umi %>% 
  group_by(SampleBio, barcode_seq) %>%
  summarise(n_umi_distinct = n_distinct(umi_seq), .groups = "drop"
  )

# --------------------------------------------------------------
# Annotate sample identity
# --------------------------------------------------------------

# Extract mouse ID and sample type from composite ID
BCM1Summary_UMI$mouse <- gsub( "_.*", "", BCM1Summary_UMI$SampleBio)
BCM1Summary_UMI$TypeSample <- gsub( ".*_", "", BCM1Summary_UMI$SampleBio)

# Simplify sample groups into major biological classes
BCM1Summary_UMI <- BCM1Summary_UMI %>% mutate(TypeSample_sh = case_when(TypeSample %in% c("Gl_Late", "Jx") ~"LATE",
                                                          TypeSample == "Early" ~ "EARLY",
                                                          TRUE ~ TypeSample))

# Log-transform UMI counts (pseudocount to avoid log(0))
BCM1Summary_UMI <- BCM1Summary_UMI %>%
  mutate(n_umi_log10 = log10(n_umi_distinct + 1)) 

```

## Add Pgen from IGOR pipeline

```{r}
# ============================================================
# Load data
# ============================================================

ubulk_indexed_sequences <- read.csv(
  file.path("C:/Users/Marthe Laisne/Desktop/FlashSeq_MultiOneCell/Data/Igor Barcodes Pgen/ubulk_indexed_sequences.csv"),
  sep = ";")

pgen_ubulk <- read.csv(
  file.path("C:/Users/Marthe Laisne/Desktop/FlashSeq_MultiOneCell/Data/Igor Barcodes Pgen/pgen_ubulk(1).txt"),
  header = FALSE,  sep = ";")

colnames(pgen_ubulk) <- c("Seq_ID", "Pgen")

Key_pgen_ubulk <- read.delim(
  file.path("C:/Users/Marthe Laisne/Desktop/FlashSeq_MultiOneCell/Results/BCM_Files for modelisation/260113_Files/Laisne_Bulk.txt"))

# Map sequence indices to actual barcode sequences
pgen_ubulk <- pgen_ubulk %>%
  left_join(ubulk_indexed_sequences,
            by = c("Seq_ID" = "seq_index"))

# Add full barcode annotation
pgen_ubulk <- pgen_ubulk %>%
  left_join(Key_pgen_ubulk,
            by = c("sequence" = "long_barcode")) %>%
  distinct()

# Compute reverse complement of barcode sequence
library(Biostrings)

pgen_ubulk$small_barcode_Rv <- sapply(
  pgen_ubulk$small_barcode,
  function(x) as.character(reverseComplement(DNAString(x)))
)

# ============================================================
# Add Pgen annotation to BCM datasets
# ============================================================

AllBCM1_R2umi <- AllBCM1_R2umi %>%
  left_join(
    pgen_ubulk %>% select(small_barcode_Rv, Pgen),
    by = c("barcode_seq" = "small_barcode_Rv")
  )


head(AllBCM1_R2umi, 15)
```



Save
```{r}
Data_dir <-  "C:/Users/Marthe Laisne/OneDrive - INSTITUT CURIE/m_laisne/Publications/2026_Laisne/Data/"

qs::qsave(AllBCM1_R2umi, paste0(Data_dir,"MiniBulk_DRAG_raw.qs"))

```


#QCs

```{r}
Data_dir <-  "C:/Users/Marthe Laisne/OneDrive - INSTITUT CURIE/m_laisne/Publications/2026_Laisne/Data/"

# Load BCM dataset
AllBCM1_R2umi <- qs::qread(paste0(Data_dir, "DRAG_raw.qs"))

sampleplan_joint <- AllBCM1_R2umi %>%
  select(samplename, plaque, well, mouse, TypeSample, Ncell, Run, 
         Replicat, mouseV2, SampleBio) %>% 
  unique()

Pgen_table <- AllBCM1_R2umi %>%
  select(barcode_seq , Pgen) %>% 
  unique()
```

## QC1 - Remove low-quality libraries based on sequencing depth
```{r}
# Compute total read counts per library (well)
BCM1Summary_Counts <- AllBCM1_R2umi %>% 
  group_by(samplename) %>%
  summarise(counts = sum(count) ) %>% 
  left_join(., sampleplan_joint,  by = "samplename") %>% 
  ungroup()

# Compute total sequencing depth for each sequencing run
Total_reads <- BCM1Summary_Counts %>% 
  ungroup() %>%
  group_by(Run) %>%
  summarise(Total_reads_Run = sum(counts)) %>% 
  ungroup()

# Add run-level sequencing depth to each library
BCM1Summary_Counts <- BCM1Summary_Counts %>% 
  left_join(., Total_reads,  by = "Run") 
```



 Determine run-specific sequencing depth threshold:
 Identify the first local minimum preceding the main density peak
This threshold separates low-depth libraries from the main population.

```{r}
# Fonction pour trouver la "cassure" de la densité
find_density_break <- function(x) {
  d <- density(x, na.rm = TRUE)
  y <- d$y
  xvals <- d$x
  
  #  Identify the first local minimum preceding the main density peak
  peak_index <- which.max(y)
  before_peak <- seq(2, peak_index - 1)
  
  if (length(before_peak) == 0) return(NA)
  
  #Loca minimum
  minima <- before_peak[
    which(diff(sign(diff(y[before_peak]))) == 2) + 1
  ]
  
  if (length(minima) > 0) {
    return(xvals[minima[length(minima)]]) # last minimum before peak
  } else {
    return(0)
  }
}

# Compute threshold independently for each sequencing run
density_breaks <- BCM1Summary_Counts %>%
  group_by(Run) %>%
  summarise(
    break_point = find_density_break(counts),
    .groups = "drop"
  )

density_breaks


# Visualize read count distributions and selected thresholds
BCM1Summary_Counts %>%
  ggplot(aes(x = counts, color = Run)) +
  geom_density() +
  geom_vline(data = density_breaks, aes(xintercept = break_point, color = Run),
             linetype = "dashed") +
  scale_x_log10() +
  theme_classic()
```

```{r}
# --------------------------------------------------------------
# Assign QC status to each run
# --------------------------------------------------------------

BCM1Summary_Counts <- BCM1Summary_Counts %>%
  left_join(., density_breaks,  by = "Run" ) %>%
  mutate(QC1_pass = case_when(counts > break_point ~ "Pass",
                              TRUE ~ "Failed"))

```


```{r}
# --------------------------------------------------------------
# Visualize retained and discarded libraries
# --------------------------------------------------------------

ggplot(data = BCM1Summary_Counts, 
       aes(x = well, y = counts, fill = QC1_pass))+
  geom_point(shape = 21, size = 3,alpha = 0.6)+
  scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
              labels = trans_format("log10", math_format(10^.x)))+
  theme_classic()+ 
  geom_hline(data = density_breaks, aes(yintercept = break_point, color = Run) )+  
  scale_fill_manual(values = sample(colorRampPalette(brewer.pal(8, "Spectral"))(2)))

```
```{r}
# --------------------------------------------------------------
# Keep only QC-passed libraries
# --------------------------------------------------------------

samples_filteredIn<- BCM1Summary_Counts$samplename[which(BCM1Summary_Counts$QC1_pass  == "Pass" )]

AllBCM1_R2umi_QC1 <- subset(AllBCM1_R2umi, sample_l %in% samples_filteredIn)

sample_plan_in <- subset(sampleplan_joint, samplename %in% samples_filteredIn)

# --------------------------------------------------------------
# Recompute barcode summary tables after QC1 filtering
# --------------------------------------------------------------

# Total read counts per barcode
BCM1Summary_Counts <- AllBCM1_R2umi_QC1 %>% 
  group_by(SampleBio, barcode_seq) %>%
  summarise(counts = sum(count), .groups = "drop"
  )

# Number of distinct UMIs per barcode
BCM1Summary_UMI <- AllBCM1_R2umi_QC1 %>% 
  group_by(SampleBio, barcode_seq) %>%
  summarise(n_umi_distinct = n_distinct(umi_seq), .groups = "drop"
  )

# Recover sample annotations
BCM1Summary_UMI$mouse <- gsub( "_.*", "", BCM1Summary_UMI$SampleBio)
BCM1Summary_UMI$TypeSample <- gsub( ".*_", "", BCM1Summary_UMI$SampleBio)
BCM1Summary_UMI <- BCM1Summary_UMI %>% mutate(TypeSample_sh = case_when(TypeSample %in% c("Gl_Late", "Jx") ~"LATE",
                                                          TypeSample == "Early" ~ "EARLY",
                                                          TRUE ~ TypeSample))

```












## QC2 - Count-based filtering using technical replicates

```{r}
# Names of technical replicate libraries used to determine the
# optimal read count threshold for barcode retention.
rep_names_1 <- c("L564T07", "L564T08", "L564T09",  "L564T11", "L564T12", 
                 "L563T28", 
                 "L564T35", "L564T36",  "L564T39",
                 "L563T42", "L563T51",
                 "L564T21", "L564T22", "L564T23", "L564T24", "L564T25", "L564T27", 
                paste0("L590C0", 1:9) ,  paste0("L590C", 10:41)     ) 

rep_names_2 <- c("L564T74","L564T75", "L564T76",  "L564T78", "L564T79",
                 "L563T69",
                 "L564T89", "L564T90",   "L564T93", 
                 "L563T71", "L563T73", 
                 "L564T81","L564T82", "L564T83", "L564T84", "L564T85",  "L564T87",
                  paste0("L590C", 55:95) )


# --------------------------------------------------------------
# Compute barcode read counts for each technical replicate
# --------------------------------------------------------------

BCM1Summary_Counts <- AllBCM1_R2umi_QC1 %>% 
  group_by(samplename, barcode_seq, Replicat) %>%
  summarise(counts = sum(count), .groups = "drop") %>% 
  left_join(., sample_plan_in , by = "samplename")

# Keep only samples with technical replicates
subset_Replicats_counts <- subset(BCM1Summary_Counts, samplename %in% c(rep_names_1, rep_names_2))


# --------------------------------------------------------------
# Build replicate comparison table
# --------------------------------------------------------------

dt <- subset_Replicats_counts[,c("barcode_seq", "counts", "mouse", "TypeSample", "Ncell", "Replicat.x")] %>%
  pivot_wider(names_from =  "Replicat.x", values_from = "counts") %>% as.data.frame()

colnames(dt) <- c("barcode_seq", "mouse", "TypeSample", "Ncell", "r1" ,"r2")


# Convert list columns into numeric read counts
dt$r1 <- lapply(dt$r1, function(X) {
  if (is.null(X)) {
    return(0)
  } else if (length(X) == 1) {
    return(X)
  } else {
    return(sum(X))
  }
}) %>% unlist()
dt$r2 <- lapply(dt$r2, function(X) {
  if (is.null(X)) {
    return(0)} else if (length(X) == 1) {
    return(X)} else {
    return(sum(X))}
}) %>% unlist()

```


```{r}
# --------------------------------------------------------------
# Identify the optimal count threshold
# --------------------------------------------------------------

# Compute replicate correlation after progressively increasing
# the minimum barcode count threshold.

thresholds <- seq(1, 100, length.out = 100)

corr_curve <- sapply(thresholds, function(t) {
  subset <- dt %>% dplyr::filter(r1 >= t & r2 >= t )
  if (nrow(subset) > 10) cor(subset$r1, subset$r2, method = "spearman") else NA
})

# Select the threshold maximizing replicate concordance.
thr_count <-  thresholds[which(corr_curve == max(corr_curve))]

```


```{r}
# --------------------------------------------------------------
# Apply count threshold to all samples
# --------------------------------------------------------------

# Keep only barcodes with read counts above the selected threshold
# within each sequencing reaction.

idx <- by(BCM1Summary_Counts$counts, BCM1Summary_Counts$samplename, function(X){ which(X> thr_count)})

white_list_BCM <- BCM1Summary_Counts %>%
  group_by(samplename) %>%
  dplyr::filter(row_number() %in% idx[[as.character(unique(samplename))]])  # Sélectionne les bonnes lignes

white_list_BCM <- split(white_list_BCM$barcode_seq, white_list_BCM$samplename)
white_list_BCM <- lapply(white_list_BCM, unique)

AllBCM1_R2umi_QC2 <- AllBCM1_R2umi_QC1[order(AllBCM1_R2umi_QC1$samplename),]

AllBCM1_R2umi_QC2 <- AllBCM1_R2umi_QC2 %>%
  group_by(samplename) %>%
  dplyr::filter(barcode_seq %in% white_list_BCM[[as.character(unique(samplename))]])

AllBCM1_R2umi_QC2$barcode_seq %>% unique %>% length
```

```{r}
# --------------------------------------------------------------
# Recompute barcode summary tables after QC2
# --------------------------------------------------------------

# Barcode read counts
BCM1Summary_Counts<- AllBCM1_R2umi_QC2 %>% 
  group_by(SampleBio, barcode_seq) %>%
  summarise(counts = sum(count), .groups = "drop"
  ) 

# Barcode UMI counts
BCM1Summary_UMI <- AllBCM1_R2umi_QC2 %>% 
  group_by(SampleBio, barcode_seq) %>%
  summarise(n_umi_distinct = n_distinct(umi_seq), .groups = "drop"
  )

# --------------------------------------------------------------
# Restore sample annotations
# --------------------------------------------------------------

BCM1Summary_UMI$mouse <- gsub( "_.*", "", BCM1Summary_UMI$SampleBio)
BCM1Summary_UMI$TypeSample <- gsub( ".*_", "", BCM1Summary_UMI$SampleBio)

BCM1Summary_UMI <- BCM1Summary_UMI %>% mutate(TypeSample_sh = case_when(TypeSample %in% c("Gl_Late", "Jx") ~"LATE",
                                                          TypeSample == "Early" ~ "EARLY",
                                                          TRUE ~ TypeSample))
```


Extended Fig 2k. 

```{r}
# --------------------------------------------------------------
# Visualize replicate concordance after count filtering
# --------------------------------------------------------------

subset_Replicats_counts <- subset(AllBCM1_R2umi_QC2, samplename %in% c(rep_names_1, rep_names_2))

#GGPLOT
dt2 <- subset_Replicats_counts[,c("barcode_seq", "count", "mouse", "TypeSample", "Ncell", "Replicat")] %>%
  pivot_wider(names_from =  "Replicat", values_from = "count") %>% as.data.frame()

colnames(dt2) <- c("barcode_seq", "mouse", "TypeSample", "Ncell", "r1" ,"r2")
dt2$r1 <- lapply(dt2$r1, function(X) {if (is.null(X)) {return(NA) } else if (length(X) == 1) {  return(X)} else { return(sum(X))}}) %>% unlist()

dt2$r2 <- lapply(dt2$r2, function(X) { if (is.null(X)) { return(NA)} else if (length(X) == 1) { return(X)} else { return(sum(X))}}) %>% unlist()



sample_row <- sample(1:nrow(dt), 2000)


ggplot()+
  geom_point(data = dt[sample_row,],
             aes(x = r1, y = r2),
             shape = 21, size = 3,alpha = 0.6, fill = "grey")+
    geom_point(data = dt2,
               aes(x = r1, y = r2),
             shape = 21, size = 3,alpha = 0.6, fill =cols_Group[4])+
  geom_smooth(data = dt2,
              aes(x = r1, y = r2),
              method = "lm", se=FALSE, color="black", formula = y ~ x) +
   stat_cor(label.y = 3) +
   stat_regline_equation( label.y = 3.5)+
  scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
              labels = trans_format("log10", math_format(10^.x)), limits = c(1, 10^4))+
  scale_x_log10(breaks = trans_breaks("log10", function(x) 10^x),
              labels = trans_format("log10", math_format(10^.x)), limits = c(1, 10^4))+
  theme_classic()+ 
  geom_hline(yintercept = thr_count)+ geom_vline(xintercept = thr_count)+
  scale_fill_manual(values = sample(colorRampPalette(brewer.pal(8, "Spectral"))(15)))

```


Extended Fig 2k. 


```{r}
plot(thresholds, corr_curve, type = "l", log = "x",
     xlab = "Minimum counts", ylab = "Correlation (r1 vs r2)")
abline(h = 0.9, col = "red", lty = 2)
```






## Step -  Merge technical replicates

```{r}
# Technical replicates originating from the same PCR well are merged.
# For barcodes detected in both replicates, read counts are averaged.
# Barcodes detected in only one replicate are discarded.

# Add a unique identifier for each PCR reaction
AllBCM1_R2umi_QC3 <- AllBCM1_R2umi_QC2 %>%
  mutate(tech_id = paste(plaque, well, sep = "_"))

# Identify PCR reactions sequenced in duplicate
replicated_ids <- AllBCM1_R2umi_QC3 %>%
  ungroup() %>%
  dplyr::filter(!is.na(Replicat)) %>%
  dplyr::count(tech_id, Replicat) %>%
  dplyr::count(tech_id) %>%
  dplyr::filter(n == 2) %>%
  pull(tech_id)


# Keep the sample identifier corresponding to duplicated reactions
replicated_sample <- AllBCM1_R2umi_QC3 %>%
  dplyr::filter(tech_id %in% replicated_ids) %>%
  group_by(tech_id) %>%
  summarise(sample_l = dplyr::last(sample_l), .groups = "drop") %>%
  pull(sample_l)

# 1.Average read counts for barcodes shared by both replicates
replicats_avg <- AllBCM1_R2umi_QC3 %>%
  dplyr::filter(tech_id %in% replicated_ids) %>%
  dplyr::group_by(tech_id, umi_seq, barcode_seq) %>%
  dplyr::filter(n_distinct(Replicat) == 2) %>%
  dplyr::summarise(count = mean(count), .groups = "drop")

# 2.Retain reactions without technical replicates
no_replicats <- AllBCM1_R2umi_QC3 %>%
  dplyr::filter(is.na(Replicat) | !(tech_id %in% replicated_ids)) %>%
  dplyr::select(tech_id, umi_seq, barcode_seq, count)

# 3.Merge averaged and non-replicated reactions
final_counts <- bind_rows(replicats_avg, no_replicats)


# 4. Restaure metadata
meta <- AllBCM1_R2umi_QC3 %>%
  dplyr::select(tech_id, plaque, well, sample_l, samplename, mouse, TypeSample, Ncell, RAW) %>%
  dplyr::distinct()

AllBCM1_R2umi_QC3 <- final_counts %>%
  dplyr::left_join(meta, by = "tech_id") %>%
  dplyr::distinct()

colnames(AllBCM1_R2umi_QC3) <- c("tech_id","umi_seq" ,"barcode_seq" ,
                                       "count", "samplename" ,"plaque" ,
                                       "well", "sample_l" ,"samplename.y",
                                       "mouse" ,"TypeSample","Ncell" ,"RAW" )

# Restore missing sample names after merging
AllBCM1_R2umi_QC3 <- AllBCM1_R2umi_QC3 %>%
  mutate(samplename = case_when(is.na(samplename)==T ~ samplename.y,
                                TRUE ~ samplename))

# Remove duplicated technical replicate entries
AllBCM1_R2umi_QC3 <- AllBCM1_R2umi_QC3 %>%
  subset(!samplename %in% replicated_sample)

head(AllBCM1_R2umi_QC3)
```




```{r}
# --------------------------------------------------------------
# Recompute barcode summary tables after merging replicates
# --------------------------------------------------------------

AllBCM1_R2umi_QC3$SampleBio <- paste0(
  AllBCM1_R2umi_QC3$mouse,
  "_",
  AllBCM1_R2umi_QC3$TypeSample)

# Total read counts per barcode
BCM1Summary_Counts <- AllBCM1_R2umi_QC3 %>%
  group_by(SampleBio, barcode_seq) %>%
  summarise(
    counts = sum(count),    .groups = "drop"  )

# Number of distinct UMIs per barcode
BCM1Summary_UMI <- AllBCM1_R2umi_QC3 %>%
  group_by(SampleBio, barcode_seq) %>%
  summarise(
    n_umi_distinct = n_distinct(umi_seq),    .groups = "drop"  )


#Add sampleBio anno
BCM1Summary_UMI$mouse <- gsub( "_.*", "", BCM1Summary_UMI$SampleBio)
BCM1Summary_UMI$TypeSample <- gsub( ".*_", "", BCM1Summary_UMI$SampleBio)
BCM1Summary_UMI <- BCM1Summary_UMI %>% mutate(TypeSample_sh = case_when(TypeSample %in% c("Gl_Late", "Jx") ~"LATE",
                                                          TypeSample == "Early" ~ "EARLY",
                                                          TRUE ~ TypeSample))
```



##QC3 - Remove PCR errors
Filter barcodes based on sequence similarity (Levenshtein distance)
```{r}
# --------------------------------------------------------------
# Prepare barcode summary for sequence similarity filtering
# --------------------------------------------------------------

# Create a sample identifier without separator
AllBCM1_R2umi_QC3$SampleBio <- paste0(AllBCM1_R2umi_QC3$mouse, AllBCM1_R2umi_QC3$TypeSample)


# Count the number of distinct UMI supporting each barcode
BCM1Summary_UMI <- AllBCM1_R2umi_QC3 %>% 
  group_by(SampleBio, barcode_seq) %>%
  summarise(n_umi_distinct = n_distinct(umi_seq), .groups = "drop"  )

#Annotate Sample type
BCM1Summary_UMI$type <- gsub(".*_", "", BCM1Summary_UMI$SampleBio)

BCM1Summary_UMI <- BCM1Summary_UMI %>%
  mutate(type = case_when(type %in% c("8903Jx", "8943Jx", "8962Jx", "8988Jx", "9052Jx", "9341Gl_Late", "9423Jx", "9424Jx") ~"Late",
                          type %in% c("8943Tm", "8962Tm", "8988Tm", "9052Tm") ~"Tumor",
                          type %in% c("8988lympho", "9052lympho") ~ "lympho",
                          TRUE ~type))

# Extract mouse identifier
BCM1Summary_UMI$mouse <- gsub("_.*", "", BCM1Summary_UMI$SampleBio)

# Annotate samples according to mouse Age
BCM1Summary_UMI <- BCM1Summary_UMI %>%
  mutate(BCM_LVmergedbzl_Figure = case_when(mouse %in% 
                                              c("9052lympho","9052Tm", "9052Jx")~"5.9m",
                                            mouse %in% c("8943Tm", "8943Jx") ~ "6.3m",
                                            mouse == "9341Jx" ~ "6.6m",
                                            mouse %in% 
                                              c( "8988Tm", "8988Jx" , "8988lympho")~ "6.9m",
                                            mouse == "9424Jx" ~ "7.3m",
                                            mouse %in% c("8962Tm", "8962Jx") ~ "7.5m",
                                            mouse == "9423Jx" ~ "7.6m",
                                            mouse == "8903Jx" ~ "7.7m",
                                            mouse %in% c("9568-D2Gl", "9568-G4Gl") ~ "3.3m",
                                            mouse %in% c("9601-D4Gl", "9601-G4Gl") ~ "2.6m",
                                            mouse %in% c("9660-D4Gl", "9660-G4Gl") ~ "2.0m",
                                            mouse %in% c("9568-D2Gl", "9568-G4Gl") ~ "3.3m",
                                            mouse %in% c("9891-D4Gl", "9891-G4Gl",
                                                         "9892-D4Gl", "9892-G4Gl") ~ "1.5m",
                                            mouse %in% c("9873-D4Gl", "9873-G4Gl",
                                                         "9873-D3Gl", "9873-G3Gl") ~ "2.2m",
                                            mouse %in% c("9826-D4Gl", "9826-G4Gl") ~ "3.2m",
                                            mouse %in% c("9784-D4Gl", "9784-G4Gl") ~ "3.8m",
                                            mouse %in% c("9757-D4Gl", "9757-G4Gl",
                                                         "9757-D3Gl", "9757-G3Gl") ~ "4.3m"))
```


```{r}
# --------------------------------------------------------------
# Function to merge highly similar barcode sequences
# --------------------------------------------------------------

# Maximum weighted Levenshtein distance allowed
thr_Lv <- 0.1

fun_InterCell <- function(vecto_seq, n_umi){
  res <- rep(NA, length(vecto_seq))
  
    # Multiple sequence alignment
  dnaSet <- DNAStringSet(vecto_seq)
  align.msa = msa(dnaSet,method="ClustalOmega")
  alignnment <- msaConvert(align.msa, "seqinr::alignment")
  
  #Restaure aorder according to vecto_seq
  alignnment$seq <- alignnment$seq[match(vecto_seq, gsub("-", "", alignnment$seq))]
    
  for(k in 1:length(vecto_seq)){

    # Compute weighted Levenshtein distances
    lv = stringdistmatrix( a = alignnment$seq[k], 
                           b = alignnment$seq[-k], 
                           useNames = T, method='lv', 
                           weight = c(d = 1, i = 1, s = 0.1, t = 1))  %>%
      reshape2::melt()
    
    # Keep barcode if no similar sequence exists
    if( sum(lv$value == 0 | lv$value > thr_Lv, na.rm = T) ==  
        nrow(na.omit(lv))){ 
      
      res[k] <- vecto_seq[k]
      
    }else{
       # Merge to the highest-UMI barcode among similar sequences
      id_proches <- which(lv$value <= thr_Lv) #identify close sequences
      seq_proches <- lv$Var2[id_proches] 
      
      seq_proche_inVecto <- c(k, 
                              which(vecto_seq %in% gsub("-", "",seq_proches) == TRUE) )
      
      id_max <-  which(n_umi == max(n_umi[seq_proche_inVecto] ))
      seq_max <- vecto_seq[id_max[1]]
      res[k] <- seq_max
      }
  }
  return(res)
}

```



The sequence similarity filter was applied independently to each PCR reaction. Similar barcode sequences were merged to the dominant sequence and their supporting UMI counts were summed.

```{r}
# --------------------------------------------------------------
# Merge barcode variants within each PCR reaction
# --------------------------------------------------------------

# Compute UMI number for each barcode in each PCR reaction
BCM1_UMI_filt_LV <- AllBCM1_R2umi_QC3 %>% 
  group_by(samplename , barcode_seq) %>%
  summarise(n_umi_distinct = n_distinct(umi_seq), .groups = "drop") %>% 
  left_join(., sampleplan_joint,  by = "samplename")

# Apply the function to all samples / well: identify the representative barcodes & keep them
AllBCM1_R2umi_QC4 <- BCM1_UMI_filt_LV %>%
  ungroup() %>%
  group_by(samplename) %>% # Par PCR reaction ie. par puit
  mutate(BCM_LVmerged = fun_InterCell(vecto_seq = barcode_seq, 
                                        n_umi = n_umi_distinct)) %>%
  ungroup() 

#Flag merged barcodes
AllBCM1_R2umi_QC4 <- AllBCM1_R2umi_QC4 %>%
  mutate(QC4_LV = case_when(barcode_seq == BCM_LVmerged ~"Pass",
                            TRUE ~"Fail"))
  

# Collapse merged barcodes by summing their UMI support
AllBCM1_R2umi_QC4$SampleBio <- paste0(AllBCM1_R2umi_QC4$mouse, AllBCM1_R2umi_QC4$TypeSample)

BCM1_QC4_MergeOnLV <-  AllBCM1_R2umi_QC4 %>%
  subset(QC4_LV == "Pass") %>%
  group_by(SampleBio, BCM_LVmerged) %>%
  summarise(n_umi_distinct = sum(n_umi_distinct), .groups = "drop"
  )

# Number of unique lineage barcodes after sequence merging
BCM1_QC4_MergeOnLV$BCM_LVmerged %>% unique %>% length
```

```{r}
#if no Lv
BCM1_QC4 <-  AllBCM1_R2umi_QC4 %>%
  group_by(SampleBio, barcode_seq) %>%
  summarise(n_umi_distinct = sum(n_umi_distinct), .groups = "drop"
  )

# Number of unique lineage barcodes after sequence merging
BCM1_QC4$barcode_seq %>% unique %>% length
```


```{r}
# --------------------------------------------------------------
# Annotate merged lineage barcode table
# --------------------------------------------------------------

# Extract mouse identifier and sample type
BCM1_QC4_MergeOnLV <- BCM1_QC4_MergeOnLV %>%
  mutate(
    mouse = str_extract(SampleBio, "^\\d+"),
    type  = str_remove(SampleBio, "^\\d+")
  )

BCM1_QC4_MergeOnLV$SampleBio <- paste0(BCM1_QC4_MergeOnLV$mouse, "_", BCM1_QC4_MergeOnLV$type)

# Assign age info to each mouse
BCM1_QC4_MergeOnLV <- BCM1_QC4_MergeOnLV %>%
  mutate(BCM_LVmergedbzl_Figure = case_when(mouse == "9052"~"5.9m",
                                            mouse == "8943" ~ "6.3m",
                                            mouse == "9341" ~ "6.6m",
                                            mouse == "8988" ~ "6.9m",
                                            mouse == "9424" ~ "7.3m",
                                            mouse == "8962" ~ "7.5m",
                                            mouse == "9423" ~ "7.6m",
                                            mouse == "8903" ~ "7.7m",
                                            mouse %in% c("9568-D2", "9568-G4") ~ "3.3m",
                                            mouse %in% c("9601-D4", "9601-G4") ~ "2.6m",
                                            mouse %in% c("9660-D4", "9660-G4") ~ "2.0m",
                                            
                                            mouse %in% c("9892-D4", "9892-G4",
                                                         "9891-D4", "9891-G4") ~ "1.5m",
                                            mouse %in% c("9873-D4", "9873-G4",
                                                         "9873-D3", "9873-G3") ~ "2.2m",
                                            mouse %in% c("9826-D4", "9826-G4") ~ "3.2m",
                                            mouse %in% c("9784-D4", "9784-G4") ~ "3.8m",
                                            mouse %in% c("9757-D4", "9757-G4",
                                                         "9757-D3", "9757-G3") ~ "4.3m"))

# Create simplified sample categories
BCM1_QC4_MergeOnLV <- BCM1_QC4_MergeOnLV %>% mutate(TypeSample_sh = case_when(type %in% c("Late", "Jx") ~"LATE",
                                                          type == "Early" ~ "EARLY",
                                                          TRUE ~ type))


# Compute average barcode size for each biological group
mean_umi <- subset(BCM1_QC4_MergeOnLV, TypeSample_sh %in% c("EARLY", "LATE", "Tm")) %>%
  group_by(TypeSample_sh) %>%
  summarise(mean_umi = mean(n_umi_distinct+1, na.rm = TRUE))
```

```{r}
# --------------------------------------------------------------
# Extended Fig. 2l - Barcode filtering by sequence similarity
# --------------------------------------------------------------

# Summarize barcode abundance before LV merging
dt  <-  AllBCM1_R2umi_QC4 %>% 
  ungroup() %>%
  group_by(SampleBio, barcode_seq, QC4_LV) %>%
  summarise(n_umi_distinct = sum(n_umi_distinct), .groups = "drop" ) %>% 
  left_join(., sampleplan_joint[, c("SampleBio", "mouse", "TypeSample")], by = "SampleBio")%>%
  distinct() %>%
  mutate(TypeSample = case_when(TypeSample %in% c("Jx", "Gl_Early", "Gl_Late") ~ "Pretumoral",
                                TRUE ~ TypeSample))


dt$type <- gsub(".*_", "", dt$SampleBio)

# Visualize retained and merged barcodes
ggplot(data = subset(dt, type != "lympho"),
       aes(x = SampleBio, y = n_umi_distinct, fill = QC4_LV))+
  geom_point(shape = 21, position = position_jitterdodge(), 
              alpha = 0.6)+
  scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
              labels = trans_format("log10", math_format(10^.x)), limits = c(1, 2*10^5))+
  theme_classic()+ 
  scale_fill_manual(values = c("grey", "#61980E"))
```







```{r}
# --------------------------------------------------------------
# Normalize barcode abundance per biological sample
# --------------------------------------------------------------

# Build sample identifiers
sample_plan_in$SampleBio_t <- paste0(sample_plan_in$mouse, "_",sample_plan_in$TypeSample)
# Exclude duplicated technical replicates
replicat_out <- which(sample_plan_in$Replicat == 2)

# Exclude duplicated technical replicates
sampleSize = by(as.numeric(sample_plan_in$Ncell[-replicat_out]),
                sample_plan_in$SampleBio_t[-replicat_out], sum, na.omit = T)

# Normalize UMI counts by sample size
BCM1_QC4_MergeOnLV <- subset(BCM1_QC4_MergeOnLV, n_umi_distinct > 0) %>%
  group_by(SampleBio) %>%
  mutate(N_UMI_norm = n_umi_distinct / unlist(sampleSize[SampleBio]))

# Rescale normalized values between 0 and 1
BCM1_QC4_MergeOnLV <- BCM1_QC4_MergeOnLV %>%
  group_by(SampleBio) %>%
  mutate(N_UMI_norm_scale = rescale(N_UMI_norm , to = c(0, 1))) %>%
  ungroup()


# Count the number of unique lineage barcodes per sample
barcode_counts <- subset(BCM1_QC4_MergeOnLV, TypeSample_sh %in% c("EARLY", "LATE", "Tm")) %>%
  group_by(SampleBio) %>%
  summarise(n_barcodes = n_distinct(BCM_LVmerged))
barcode_counts <- left_join(barcode_counts,
                             distinct(subset(BCM1_QC4_MergeOnLV, TypeSample_sh %in% c("EARLY", "LATE", "Tm"))[, c("SampleBio", "TypeSample_sh")]),
                             by = "SampleBio")
```
##QC4 - Pgen
Pgen was computed on this dataset using IGor pipeline. 

```{r}
BCM1_QC4_MergeOnLV <- BCM1_QC4_MergeOnLV %>%
  left_join(Pgen_table, by = c( "BCM_LVmerged" =  "barcode_seq") )

AllBCM1_R2umi_QC3 <- AllBCM1_R2umi_QC3 %>%
  left_join(Pgen_table, by = c( "barcode_seq" =  "barcode_seq") )


AllBCM1_R2umi_QC4 <- AllBCM1_R2umi_QC4 %>%
  left_join(Pgen_table, by = c( "BCM_LVmerged" =  "barcode_seq") )
```







```{r}
# Load QC-filtered BCM datasets
qs::qsave(BCM1_QC4_MergeOnLV, paste0(outputDir, "MiniBulk_BCM1_QC3-4.qs"))
qs::qsave(AllBCM1_R2umi_QC3, paste0(outputDir, "AllBCM1_R2umi_QC3.qs"))
qs::qsave(AllBCM1_R2umi_QC4, paste0(outputDir, "AllBCM1_R2umi_QC4.qs"))

# Load experimental design / sampling plan
qs::qsave(sample_plan_in, paste0(outputDir, "sample_plan_keptQC3-4.qs"))
```



### Prepare MiniBulk barcode metadata

```{r}

# Load QC-filtered BCM datasets
BCM1_QC4_MergeOnLV <- qs::qread(paste0(outputDir, "MiniBulk_BCM1_QC3-4.qs"))
AllBCM1_R2umi_QC3 <- qs::qread(paste0(outputDir, "AllBCM1_R2umi_QC3.qs"))
AllBCM1_R2umi_QC4 <- qs::qread(paste0(outputDir, "AllBCM1_R2umi_QC4.qs"))

# Load experimental design / sampling plan
sample_plan_in <- qs::qread(paste0(outputDir, "sample_plan_keptQC3-4.qs"))


# Exclude replicate from downstream size estimation
replicat_out <- which(sample_plan_in$Replicat == 2)

# Compute number of cells per sample (excluding replicate ofc)
sampleSize = by(as.numeric(sample_plan_in$Ncell[-replicat_out]),
                sample_plan_in$SampleBio[-replicat_out], sum, na.omit = T)

#Add Nb_cells informations
BCM1_QC4_MergeOnLV$Nb_cells <- sampleSize[BCM1_QC4_MergeOnLV$SampleBio]

```

```{r}
# ============================================================
# Identify barcode sharing across mice
# ============================================================

# Extract simplified mouse identifier
BCM1_QC4_MergeOnLV$mouseV2 <- gsub("-.*", "", BCM1_QC4_MergeOnLV$mouse)

# Identify barcodes appearing in multiple mice (blacklist)
List_bcm <- by(
  BCM1_QC4_MergeOnLV$BCM_LVmerged,
  BCM1_QC4_MergeOnLV$mouseV2,
  unique
)

List_bcm <- unlist(List_bcm)
Blacklist <- List_bcm[duplicated(List_bcm)]

# Label barcodes as shared or unique across mice
BCM1_QC4_MergeOnLV <- BCM1_QC4_MergeOnLV %>%
  mutate(
    Barcode_sharing_mice = case_when(
      BCM_LVmerged %in% Blacklist ~ "shared",
      TRUE ~ "unique"    )  )

# Count number of mice per barcode
BCM1_QC4_MergeOnLV <- BCM1_QC4_MergeOnLV %>%
  group_by(BCM_LVmerged) %>%
  mutate(Barcode_sharing_mice_N = n_distinct(mouseV2))
```

```{r}
# Estimate number of clones per sample
Clone_Number <- by(
  subset(BCM1_QC4_MergeOnLV, n_umi_distinct > 0)$BCM_LVmerged,
  subset(BCM1_QC4_MergeOnLV, n_umi_distinct > 0)$SampleBio,
  function(X) length(unique(X))
)

# Reorder samples for plotting consistency
BCM1_QC4_MergeOnLV$SampleBio <- factor(
  BCM1_QC4_MergeOnLV$SampleBio,
  levels = x_order_
)

# Merge with cell number per sample
BCM1_QC4_MergeOnLV <- BCM1_QC4_MergeOnLV %>%
  group_by(SampleBio) %>%
  mutate(
    UMI_tot_Sample = sum(n_umi_distinct),
    UMI_fraction = n_umi_distinct / UMI_tot_Sample,
    CloneSizeNorm = UMI_fraction * Nb_cells
  ) %>%
  ungroup()

# Define sample categories (early / late / tumor-like)
BCM1_QC4_MergeOnLV <- BCM1_QC4_MergeOnLV %>%
  mutate(
    TypeSample_sh_v2 = case_when(
      BCM_LVmergedbzl_Figure %in% c("1.5m", "2.0m", "2.2m") ~ "EARLY_2m",
      BCM_LVmergedbzl_Figure %in% c("2.6m","3.2m","3.3m","3.8m","4.3m") ~ "EARLY_3m",
      TypeSample_sh == "LATE" ~ "LATE",
      TRUE ~ "others"
    )
  )
```

```{r}
# ============================================================
# Compute sample-level summaries
# ============================================================

# Number of cells per sample
df_cells <- sample_plan_in[-replicat_out, ] %>%
  group_by(SampleBio) %>%
  summarise(Nb_cells = sum(as.numeric(Ncell), na.rm = TRUE)) %>%
  ungroup()

# Number of barcodes per sample
df_barcodes <- BCM1_QC4_MergeOnLV %>%
  filter(
    TypeSample_sh %in% c("EARLY", "LATE", "Tm"),
    n_umi_distinct > 0
  ) %>%
  group_by(SampleBio) %>%
  summarise(Nb_Barcodes = n_distinct(BCM_LVmerged)) %>%
  ungroup() %>%
  left_join(
    BCM1_QC4_MergeOnLV[, c("SampleBio", "TypeSample_sh_v2")],
    by = "SampleBio"
  ) %>%
  distinct()

# Merge cell and barcode summaries
df_summary <- full_join(df_cells, df_barcodes, by = "SampleBio") %>%
  mutate(Ratio = Nb_cells / Nb_Barcodes) %>%
  arrange(SampleBio)

df_summary
```
```{r}
by(df_summary$Nb_Barcodes, df_summary$TypeSample_sh_v2, sum)
```





## Save
```{r}
qs::qsave(BCM1_QC4_MergeOnLV_Nmice_Pgen, paste0(outputDir,
                                                "BCM1_QC4_MergeOnLV_Nmice_Pgen.qs"))

qs::qsave(AllBCM1_R2umi_QC4_Pgen, paste0(outputDir,
                                                "AllBCM1_R2umi_QC4_Pgen.qs"))
```
















