---
title: "R Notebook"
output: html_notebook
---

#Intro

```{r}
library(CellBarcode)


library(data.table)
library(tidyverse)
library(dplyr)
library(readxl)

library(Seurat)
library(patchwork)
library(DoubletFinder)

library(ggplot2)
library(gridExtra)
library(ggExtra)
library(viridis)
library(Matrix)
library(ggsci)
library(cowplot)
library(platetools)
library(scales)
library(gridExtra)
library(qs)
library(RColorBrewer)
library(ggridges)
library(rstatix)
library(ComplexHeatmap)
library(circlize)


library(stringdist)
library(Biostrings)
library(msa)
library(seqinr)
library(ggmsa)

# -------------------------------------------------------------- 
# Color definition 
# --------------------------------------------------------------

cols_fate <- c("basal"  = "#236ec4", 
               "LP"     = "#a01858",      
               "HS"     = "#f9c40c",  
               "Alv"    = "#F7DCAD")


cols_Group <- c("EARLY" = "#C0D69F", 
                "EARLY_2m" = "#C0D69F", 
                "EARLY_3m" = "#C0D69F", 
                "LATE"  = "#61980E",
                "Tm"    = "#841811")

cols_MonoGland <- c("8415_G2-3" = "#375708",  
                    "8718_D2-3"     = "#61980E",
                    "8964_G4"       = "#C0D69F")
```

```{r}
# ============================================================
# Define sample ordering for visualization
# ============================================================

x_order <- c(
  "9891-G4Gl_Early","9891-D4Gl_Early","9892-G4Gl_Early","9892-D4Gl_Early",
  "9873-G4Gl_Early","9873-D4Gl_Early","9873-G3Gl_Early","9873-D3Gl_Early",
  "9826-G4Gl_Early","9826-D4Gl_Early","9784-G4Gl_Early","9784-D4Gl_Early",
  "9757-G4Gl_Early","9757-D4Gl_Early","9757-G3Gl_Early","9757-D3Gl_Early",
  "9660-G4Gl_Early","9660-D4Gl_Early","9601-G4Gl_Early","9601-D4Gl_Early",
  "9568-G4Gl_Early","9568-D2Gl_Early",
  "9052Jx","9052Tm","8943Jx","8943Tm",
  "9341Gl_Late","8988Jx","8988Tm",
  "9424Jx","8962Jx","8962Tm","9423Jx","8903Jx"
)

x_order_ <- c(
  "9891-G4_Gl_Early","9891-D4_Gl_Early","9892-G4_Gl_Early","9892-D4_Gl_Early",
  "9660-G4_Gl_Early","9660-D4_Gl_Early",
  "9873-G4_Gl_Early","9873-D4_Gl_Early","9873-G3_Gl_Early","9873-D3_Gl_Early",
  "9601-G4_Gl_Early","9601-D4_Gl_Early",
  "9826-G4_Gl_Early","9826-D4_Gl_Early",
  "9568-G4_Gl_Early","9568-D2_Gl_Early",
  "9784-G4_Gl_Early","9784-D4_Gl_Early",
  "9757-G4_Gl_Early","9757-D4_Gl_Early","9757-G3_Gl_Early","9757-D3_Gl_Early",
  "9052_Jx","9052_Tm","8943_Jx","8943_Tm",
  "9341_Gl_Late","8988_Jx","8988_Tm",
  "9424_Jx","8962_Jx","8962_Tm","9423_Jx","8903_Jx"
)
```
```{r}
Data_dir <- "C:/Users/Marthe Laisne/OneDrive - INSTITUT CURIE/m_laisne/Publications/2026_Laisne/Scripts/Output/MiniBulk_DRAG/"

# Load QC-filtered BCM datasets 
# 1. Summary table per sample after merging on LV distance
BCM1_QC4_MergeOnLV_Nmice_Pgen <- qs::qread(paste0(Data_dir, "BCM1_QC4_MergeOnLV_Nmice_Pgen.qs"))

#2. All barcodes, before merging on LV distance
AllBCM1_R2umi_QC4_Pgen <- qs::qread(paste0(Data_dir, "AllBCM1_R2umi_QC4_Pgen.qs"))
```












# Pgen-based analyses and visualization 



## Pgen vs barcode sharing


Scatter plot : 
```{r}

BCM1_QC4_MergeOnLV_Nmice_Pgen %>%
  mutate(Pgen = case_when( is.finite(Pgen) == F ~10**-12,
                           TRUE ~ Pgen)) %>%
  dplyr::filter(!is.na(Pgen), is.finite(Pgen)) %>%   # Replace missing or infinite Pgen values with a minimal value
   group_by(BCM_LVmerged) %>%
  summarise(
    Pgen = dplyr::first(Pgen),
    Barcode_sharing_mice_N = max(Barcode_sharing_mice_N),
    .groups = "drop"
  ) %>%
  mutate(Barcode_sharing_mice_Factor = case_when(Barcode_sharing_mice_N == 1  ~ "unique",
                                                 Barcode_sharing_mice_N > 1 & Barcode_sharing_mice_N < 10 ~ "2-9",
                                                 Barcode_sharing_mice_N>=10 ~  ">10/17")) %>%
  mutate(Barcode_sharing_mice_Factor = factor(Barcode_sharing_mice_Factor,
                                              levels = c("unique", "2-9",
                                                         ">10/17"))) %>%
  
  ggplot(aes(x = Pgen, y = Barcode_sharing_mice_N, 
             fill = Barcode_sharing_mice_Factor)) +
  geom_point(shape = 21, size = 3,
             position = position_jitter(width = 0.1, height = 0.1)) +
  scale_x_log10()+
  geom_vline(xintercept = 10**-4.5)+
  
  labs(
    x = "Pgen", y = "N mice",
    fill = "Barcode sharing (# mice)",
    title = "Pgen distribution by barcode sharing") +

  theme_classic()
```

2. Pgen distribution histogram

```{r}
BCM1_QC4_MergeOnLV_Nmice_Pgen %>%
 mutate(Pgen = case_when( is.finite(Pgen) == F ~10**-12,
                           TRUE ~ Pgen))%>%
  group_by(BCM_LVmerged) %>%
    summarise(
    Pgen = dplyr::first(Pgen),
    Barcode_sharing_mice_N = max(Barcode_sharing_mice_N),
    .groups = "drop"
  ) %>%
  mutate(Barcode_sharing_mice_Factor = 
           case_when(Barcode_sharing_mice_N == 1  ~ "unique",
                     Barcode_sharing_mice_N > 1 & Barcode_sharing_mice_N < 10 ~ "2-9",
                     Barcode_sharing_mice_N>=10 ~  ">10/17")) %>%
  mutate(Barcode_sharing_mice_Factor = 
           factor(Barcode_sharing_mice_Factor,
            levels = c("unique", "2-9", ">10/17"))) %>%
  
  ggplot(aes(x = -log10(Pgen), fill = Barcode_sharing_mice_Factor)) +
  geom_histogram(alpha = 0.5, aes(y = after_stat(count))) +
   geom_vline(xintercept = 4.5)+

  labs(
    x = "Pgen",
    fill = "Barcode sharing (# mice)",
    title = "Pgen distribution by barcode sharing") +

  theme_classic()

```


Threshold scan: relationship between Pgen and barcode sharing : 
```{r}

df <- BCM1_QC4_MergeOnLV_Nmice_Pgen %>%
  mutate(Pgen = case_when( is.finite(Pgen) == F ~10**-12,
                           TRUE ~ Pgen))%>%
   group_by(BCM_LVmerged) %>%
    summarise(
    Pgen = dplyr::first(Pgen),
    Barcode_sharing_mice_N = max(Barcode_sharing_mice_N),
    .groups = "drop"
  )

# Explore multiple Pgen cutoffs (10^0 to 10^-11)
thr_grid <- seq(0, 11, by = 0.5)

res <- lapply(thr_grid, function(x){

  thr <- 10^-x

  sub <- df %>% dplyr::filter(Pgen <= thr)

  if(nrow(sub) == 0){
    return(data.frame(x = x, frac = NA))
  }
  # Fraction of barcodes shared across ≥2 mice
  frac_shared <- mean(sub$Barcode_sharing_mice_N > 1)

  data.frame(
    x = x,
    frac = frac_shared
  )

}) %>% bind_rows()

# Number of barcodes retained at each threshold
res$n <- sapply(thr_grid, function(x){
  sum(df$Pgen <= 10^-x)
})

# plot

ggplot(res, aes(x = x, y = frac)) +
  geom_point(size = 3) +
  geom_text(aes(label = n), vjust = -1, size = 3)+
  labs(
    x = expression("Pgen threshold 10"^-x),
    y = "Fraction barcode shared between mice"
  ) +
  theme_classic()
```

```{r}
# Relationship between barcode generation probability and abundance
BCM1_QC4_MergeOnLV_Nmice_Pgen %>%
  
  ggplot(aes(x= -log10(Pgen), y = n_umi_distinct)) +
  geom_point(shape = 21, alpha = 0.5)+
  theme_classic() + 
  scale_y_log10() +
  
BCM1_QC4_MergeOnLV_Nmice_Pgen %>%
  
  ggplot(aes(x= -log10(Pgen), y = n_umi_distinct)) +
  geom_point(shape = 21, alpha = 0.5)+
  theme_classic() 

#Pgen discretization vs abundance
BCM1_QC4_MergeOnLV_Nmice_Pgen %>%
  
    # Bin Pgen into log10 intervals for visualization
  mutate(Pgen_disc = case_when(Pgen < 10**-12 |is.na(Pgen) ~ "<-12",
                               Pgen > 10**-12 & Pgen < 10**-11 ~ "-12, -11",
                                Pgen > 10**-11 & Pgen < 10**-10 ~ "-11, -10",
                                Pgen > 10**-10 & Pgen < 10**-9 ~ "-10, -9",
                                Pgen > 10**-9 & Pgen < 10**-8 ~ "-9, -8",
                                Pgen > 10**-8 & Pgen < 10**-7 ~ "-8, -7",
                                Pgen > 10**-7 & Pgen < 10**-6 ~ "-7, -6",
                                Pgen > 10**-6 & Pgen < 10**-5 ~ "-6, -5",
                                Pgen > 10**-5 & Pgen < 10**-4 ~ "-5, -4",
                                Pgen > 10**-4 & Pgen < 10**-3 ~ "-4, -3",
                                Pgen > 10**-3 & Pgen < 10**-2 ~ "-3, -2",
                                Pgen > 10**-2 & Pgen < 10**-1 ~ "-2, -1",
                                Pgen > 10**-1 & Pgen < 10**-0 ~ "-1, 0")) %>%
  
  ggplot(aes(x= Pgen_disc, y = n_umi_distinct)) +
  geom_boxplot()+
  theme_classic() +
  scale_y_log10()+
  scale_x_discrete(limits = c("<-12",    "-12, -11","-11, -10",    "-10, -9",  "-9, -8",     "-8, -7",        "-7, -6",  "-6, -5",  "-5, -4",         "-4, -3",    "-3, -2"))


```

## Pgen control
###Clone number
```{r}
# Pgen thresholds
thr_vec <- 10^seq(0, -7, by = -0.25)

# Compute clone number for each threshold
dt_thr <- purrr::map_dfr(thr_vec, function(thr){

  BCM1_QC4_MergeOnLV_Nmice_Pgen %>%
    filter(
      TypeSample_sh_v2 %in% c("EARLY_2m","EARLY_3m","LATE","Tm"),
      Pgen <= thr          # adapter si nécessaire
    ) %>%
    group_by(SampleBio, TypeSample_sh_v2) %>%
    summarise(
      CloneNumber = n_distinct(BCM_LVmerged),
      .groups = "drop"
    ) %>%
    mutate(
      Thr = thr,
      logThr = -log10(thr)    )

})

stats <- purrr::map_dfr(thr_vec, function(thr){

  dat <- BCM1_QC4_MergeOnLV_Nmice_Pgen %>%
    filter(
      TypeSample_sh_v2 %in% c("EARLY_2m","EARLY_3m","LATE"),
      Pgen <= thr    ) %>%

    group_by(SampleBio, TypeSample_sh_v2) %>%
    summarise(
      CloneNumber = n_distinct(BCM_LVmerged),
      .groups = "drop"
    )     %>%

  dunn_test(
    CloneNumber ~ TypeSample_sh_v2,
    p.adjust.method = "BH"
  ) %>%
    mutate(
      Thr = thr,
      logThr = -log10(thr)    )

})

stats




dt_sum <- dt_thr %>%
  group_by(TypeSample_sh_v2, Thr, logThr) %>%
  summarise(
    median = median(CloneNumber),
    mean = mean(CloneNumber),
    q25 = quantile(CloneNumber, 0.25),
    q75 = quantile(CloneNumber, 0.75),
    .groups = "drop"  )

stats <- stats %>%
  mutate(
    y.position = max(dt_sum$median_umi) * 1.05
  )


P_clones_thr <- ggplot(dt_sum,
                       aes(logThr, mean)) +

  geom_line(size = 1.2, aes( color = TypeSample_sh_v2,     fill = TypeSample_sh_v2)) +
  geom_point(size = 2, aes( color = TypeSample_sh_v2,     fill = TypeSample_sh_v2)) +

  labs(
    x = expression(-log[10](Pgen~threshold)),
    y = "Mean number of clones"  ) +
  theme_classic()

P_clones_thr


P_clones_thr +
  ggpubr::stat_pvalue_manual(    stats,
    x = "logThr",    y.position = "y.position",    hide.ns = T  )
```
### Clone size
```{r}

library(dplyr)
library(broom)


# Pgen thresholds
thr_vec <- 10^seq(0, -7, by = -0.5)

dt_thr <- purrr::map_dfr(thr_vec, function(thr){

  BCM1_QC4_MergeOnLV_Nmice_Pgen %>%
    filter(
      TypeSample_sh_v2 %in% c("EARLY_2m","EARLY_3m","LATE"),
      Pgen <= thr          ) %>%
    group_by(TypeSample_sh_v2) %>%
    summarise(
      median_umi = median(n_umi_distinct, na.rm = TRUE),
      q25 = quantile(n_umi_distinct, 0.25, na.rm = TRUE),
      q75 = quantile(n_umi_distinct, 0.75, na.rm = TRUE),
      .groups = "drop"    ) %>%
    mutate(
      Thr = thr,
      logThr = -log10(thr)    )

})

stats <- purrr::map_dfr(thr_vec, function(thr){
  dat <- BCM1_QC4_MergeOnLV_Nmice_Pgen %>%
    filter(
      TypeSample_sh_v2 %in% c("EARLY_2m","EARLY_3m","LATE"),
      Pgen <= thr    )

  fit <- aov(n_umi_distinct ~ TypeSample_sh_v2, data = dat)

  broom::tidy(fit) %>%
    filter(term == "TypeSample_sh_v2") %>%
    transmute(
      Thr = thr,
      logThr = -log10(thr),
      p = p.value
    )
})

stats <- stats %>%
  mutate(
    y.position = max(dt_thr$median_umi, na.rm = TRUE) * 1.05,
    label = paste0(signif(p, 2))
  )

P_UMI_thr <- ggplot(dt_thr,
                    aes(logThr, median_umi)) +
  geom_line(linewidth = 1.2, aes(color = TypeSample_sh_v2,fill = TypeSample_sh_v2)) +
  geom_point(size = 2, aes(color = TypeSample_sh_v2,fill = TypeSample_sh_v2)) +
  theme_classic() +
  labs(
    x = expression(-log[10](Pgen~threshold)),
    y = "Median clone size"  )

P_clones_thr |  P_UMI_thr +
  geom_text(
    data = stats,
    aes(x = logThr, y = y.position, label = label),
    inherit.aes = FALSE,
    size = 3
  )
```


Not used



# #### VISUALIZATION ####



```{r}
# Ensure sample order is consistent across plots
BCM1_QC4_MergeOnLV_Nmice_Pgen$SampleBio <- factor(BCM1_QC4_MergeOnLV_Nmice_Pgen$SampleBio, levels = (x_order_))


BCM1_QC4_MergeOnLV_Nmice_Pgen <- BCM1_QC4_MergeOnLV_Nmice_Pgen %>%
    # refine sample grouping
  mutate(TypeSample_sh_v2 = case_when(BCM_LVmergedbzl_Figure %in% c("1.5m",  "2.0m", "2.2m" ) ~             "EARLY_2m",
                                      BCM_LVmergedbzl_Figure %in% c("2.6m","3.2m", "3.3m", "3.8m", "4.3m") ~"EARLY_3m",
                                      TypeSample_sh == "LATE" ~                                             "LATE",
                                      TRUE ~ "others"))

############################################
# QC Summaries
############################################

# Number of barcodes per sample and condition
BCM1_QC4_MergeOnLV_Nmice_Pgen %>%
  ungroup()%>%
  group_by(SampleBio, TypeSample_sh_v2 ) %>%
  summarize(barcodes = length((BCM_LVmerged)))


# Number of Minibuk per condition
by(BCM1_QC4_MergeOnLV_Nmice_Pgen$BCM_LVmerged, 
   BCM1_QC4_MergeOnLV_Nmice_Pgen$TypeSample_sh_v2, 
   function(X){length(X)})

# Number of cells
by(BCM1_QC4_MergeOnLV_Nmice_Pgen$Nb_cells, 
   BCM1_QC4_MergeOnLV_Nmice_Pgen$TypeSample_sh_v2, 
   FUN = function(X){sum(unique(X))})

```

## Fig 3c. - Bubble with density

```{r}
BCM1_QC4_MergeOnLV_Nmice_Pgen %>%
  subset(SampleBio %in% c( "9891-G4_Gl_Early", "9891-D4_Gl_Early" ,  "8988_Jx", "8988_Tm"    )) %>%
  
  ggplot( aes(x = n_umi_distinct, fill = TypeSample_sh))+
  geom_density(alpha = 0.5, position = "identity", adjust = 0.8) +
  geom_jitter( aes(y = 0, size = n_umi_distinct),
    height = 0.02, width = 0, alpha = 0.6,  shape = 21, color = "black") +
  scale_x_log10() +
  facet_grid(SampleBio ~ .) +
  scale_fill_manual(values =cols_Group) +
  theme_classic() +
  theme(  strip.text.y = element_text(angle = 0),
    axis.text.y = element_text(size = 9)) +
  labs(x = "N UMI distinct", y = "Density", fill = "TypeSample") %>%
  
  print()
```




## Ext Fig 5c. Density + jitter per sample
```{r}
BCM1_QC4_MergeOnLV_Nmice_Pgen %>%
  subset( TypeSample_sh_v2 %in% c("EARLY_2m","EARLY_3m", "LATE")  ) %>%
  
  ggplot( aes(x = n_umi_distinct, y = SampleBio, fill = TypeSample_sh_v2, group = SampleBio))+
  geom_density_ridges(alpha = 0.6, scale = 1.2, rel_min_height = 0.01) +
  scale_x_log10(limits = c(0.1, 100000)) +
  scale_fill_manual(values = c(cols_Group, "grey")) +
  theme_classic() +

  theme(  strip.text.y = element_text(angle = 0),
  axis.text.y = element_text(size = 9)) +
  labs(x = "N UMI distinct", y = "Density", fill = "TypeSample") %>%
  
  print()
```

## Fig 3d. Clone number

```{r}
# Number of clone by mammary gland
dt <- BCM1_QC4_MergeOnLV_Nmice_Pgen %>%
  filter(TypeSample_sh_v2 %in% c("EARLY_2m", "EARLY_3m", "LATE", "Tm") ) %>% #, Barcode_sharing_mice_N <= 1) %>%
  group_by(SampleBio, TypeSample_sh_v2) %>%
  summarise(CloneNumber = n_distinct(BCM_LVmerged), .groups = "drop") %>%
  left_join(BCM1_QC4_MergeOnLV_Nmice_Pgen[,c("SampleBio", "BCM_LVmergedbzl_Figure")], by = "SampleBio") %>%
  distinct() %>%
  mutate(Age = gsub("m", "", BCM_LVmergedbzl_Figure) %>% as.numeric)

head(dt)


# summary by bilogical group
dt_sum <- dt %>%
  group_by(TypeSample_sh_v2) %>%
  summarise(
    m = mean(CloneNumber),
    s = sd(CloneNumber),
    .groups = "drop" )

# Plot
P_clones <- ggplot(dt_sum, aes(x = TypeSample_sh_v2, y = m, fill = TypeSample_sh_v2)) +
  geom_bar(stat = "identity", color = "black") +
  geom_point(
    data = dt,
    aes(x = TypeSample_sh_v2, y = CloneNumber, fill = TypeSample_sh_v2),
    alpha = 0.6, shape = 21, position = position_jitter(width = 0.2, height = 0), size = 3
  ) +
  geom_errorbar(aes(ymin = m , ymax = m + s), width = 0.2) +
  scale_fill_manual(values = cols_Group) +
  theme_classic() +
  labs(x = "TypeSample", y = "NCloneNumber", fill = "TypeSample")



P_clones
```

```{r}
# Statistics
aov(CloneNumber ~ TypeSample_sh_v2, data = dt) %>% summary()
TukeyHSD(aov(CloneNumber ~ TypeSample_sh_v2, data = dt))

```


## Fig 3e. Clone size

```{r}

df_plot <- BCM1_QC4_MergeOnLV_Nmice_Pgen %>%
  subset(TypeSample_sh_v2 %in% c("EARLY_2m","EARLY_3m", "LATE")) %>%
  group_by(TypeSample_sh_v2, TypeSample_sh) %>%
  summarise(
    mean_umi = mean(n_umi_distinct, na.rm = TRUE),
    sd_umi   = sd(n_umi_distinct, na.rm = TRUE),
    .groups = "drop"
  )


p1 <-  ggplot(df_plot, aes(x = mean_umi, y = TypeSample_sh_v2, fill = TypeSample_sh)) +
  geom_col(alpha = 0.7, color = "black") +
  geom_jitter(data = BCM1_QC4_MergeOnLV_Nmice_Pgen %>%
  subset( TypeSample_sh_v2 %in% c("EARLY_2m","EARLY_3m", "LATE")  ),
  aes(x = n_umi_distinct, y = TypeSample_sh_v2, fill = TypeSample_sh),
  height = 0.02, width = 0, alpha = 0.6,  shape = 21, color = "black") +
  geom_errorbarh(aes(xmin = mean_umi,
                     xmax = mean_umi + sd_umi),
                 height = 0.2) +
  scale_fill_manual(values = cols_Group) +
  scale_x_log10(breaks = scales::trans_breaks("log10", function(x) 10^x),
              labels = scales::trans_format("log10", scales::math_format(10^.x)))+
  theme_classic() +
  theme(strip.text.y = element_text(angle = 0),
        axis.text.y = element_text(size = 9)) +
  coord_flip() +
  labs(x = "N UMI distinct (mean ± SD)", y = "", fill = "TypeSample")


p2 <- BCM1_QC4_MergeOnLV_Nmice_Pgen %>%
  subset( TypeSample_sh_v2 %in% c("EARLY_2m","EARLY_3m", "LATE")  ) %>%
  
  ggplot( aes(x = n_umi_distinct, y = TypeSample_sh_v2, fill = TypeSample_sh))+
  geom_jitter( height = 0.2, width = 0, alpha = 0.6,  shape = 21, color = "black") +
  geom_boxplot(width = 0.4, outliers = F, alpha = 0.5)+
  scale_x_log10(limits = c(0.1, 100000)) +
  scale_fill_manual(values =cols_Group) +
  theme_classic() +
  theme(  strip.text.y = element_text(angle = 0),
  axis.text.y = element_text(size = 9)) +
  coord_flip()+
  labs(x = "N UMI distinct", y = "", fill = "TypeSample")

p1 + p2

```

Statistical comparisons
```{r}
aov((n_umi_distinct) ~ TypeSample_sh_v2, 
    data = BCM1_QC4_MergeOnLV_Nmice_Pgen %>%
  subset( TypeSample_sh_v2 %in% c("EARLY_2m","EARLY_3m", "LATE")  ) ) %>% summary()

TukeyHSD(aov((n_umi_distinct) ~ TypeSample_sh_v2, 
             data = BCM1_QC4_MergeOnLV_Nmice_Pgen %>%
  subset( TypeSample_sh_v2 %in% c("EARLY_2m","EARLY_3m", "LATE") ) ))

```






## Ext Fig2n. Correlograme
```{r}
# Build clone × sample matrix
mat <- AllBCM1_R2umi_QC4_Pgen %>%
  filter(QC4_LV == "Pass") %>% 
  dplyr::select(samplename, BCM_LVmerged, n_umi_distinct) %>%
  tidyr::pivot_wider(names_from = samplename, values_from = n_umi_distinct, values_fill = 0) 


cor_mat <- cor(mat %>% select(-BCM_LVmerged), method = "pearson", use = "pairwise.complete.obs")

# Annotation metadata for heatmap
annot <- AllBCM1_R2umi_QC4_Pgen %>%
  filter(QC4_LV == "Pass" ) %>%
  distinct(samplename, SampleBio, Ncell, TypeSample, mouseV2) %>%
  column_to_rownames("samplename") %>%
  mutate(Ncell = as.numeric(Ncell))

ha <- HeatmapAnnotation(

  TypeSample = annot$TypeSample,
  Mouse = annot$mouseV2,
    SampleBio = annot$SampleBio,
  Ncell = annot$Ncell,
  col = list(
    TypeSample = structure(
       sample(colorRampPalette(brewer.pal(8, "Spectral"))(length(unique(annot$TypeSample)))),
      names = unique(annot$TypeSample)),
    Mouse = structure(
       sample(colorRampPalette(brewer.pal(8, "Greens"))(length(unique(annot$mouseV2)))),
      names = unique(annot$mouseV2)),
    SampleBio = structure(
       sample(colorRampPalette(brewer.pal(8, "Greys"))(length(unique(annot$SampleBio)))),
      names = unique(annot$SampleBio)),
     
    Ncell = colorRamp2(c(1, 10, 20,40, 50,80, 100,  200), colorRampPalette(brewer.pal(8, "YlGnBu"))(8))))

All <- Heatmap(cor_mat,
        name = "Pearson r",
        top_annotation = ha,
        col = colorRamp2(  seq(0, 1, length.out = 10),  c("white", brewer.pal(9, "Reds"))),
        clustering_distance_rows = "pearson",
        clustering_method_rows = "ward.D2",
        clustering_distance_columns  = "pearson",
        clustering_method_columns  = "ward.D2",
        show_row_names = F,
        show_column_names = F, use_raster = TRUE)

######### Zoom one sample
# ------------------------------------------------
Sample_tested <- "9568-G4_Gl_Early" #"9757-G3_Gl_Early"
mat <- AllBCM1_R2umi_QC4_Pgen %>%
  ungroup() %>%
  filter(SampleBio %in% Sample_tested) %>%
  filter(QC4_LV == "Pass") %>% 
  filter(TypeSample %in% c("Gl_Early", "Gl_Late", "Jx", "Tm")) %>% 
  dplyr::select(samplename, BCM_LVmerged, n_umi_distinct) %>%
  tidyr::pivot_wider(names_from = samplename, values_from = n_umi_distinct, values_fill = 0) 

mat_cor <- mat %>% 
  dplyr::select(-BCM_LVmerged)

cor_mat <- cor(mat_cor, method = "pearson", use = "pairwise.complete.obs")

annot <- AllBCM1_R2umi_QC4_Pgen %>%
  ungroup() %>%
    filter(SampleBio %in% Sample_tested) %>%
    filter(TypeSample %in% c("Gl_Early", "Gl_Late", "Jx", "Tm")) %>% 
    #subset(Pgen <= Thr_Pgen) %>%
    filter(SampleBio %in% Sample_tested) %>%
    subset(QC4_LV == "Pass") %>% 
  distinct(samplename, SampleBio, Ncell, TypeSample) %>%
  column_to_rownames("samplename") %>%
  mutate(Ncell = as.numeric(Ncell))

ha <- HeatmapAnnotation(
  SampleBio = annot$SampleBio,
  Ncell = annot$Ncell,
  TypeSample = annot$TypeSample,
  col = list(
    SampleBio = structure(
      rand_color(length(unique(annot$SampleBio))),
      names = unique(annot$SampleBio)),
     TypeSample = structure(
      rand_color(length(unique(annot$TypeSample))),
      names = unique(annot$TypeSample)),
    Ncell = colorRamp2(c(0,100), c("lightblue", "darkblue"))))

One <- Heatmap(cor_mat,
        name = "Pearson r",
        top_annotation = ha,
        col = colorRamp2(  seq(0, 1, length.out = 10),  c("white", brewer.pal(9, "Reds"))),
        clustering_distance_rows = "pearson",
        clustering_method_rows = "ward.D2",
        clustering_distance_columns  = "pearson",
        clustering_method_columns  = "ward.D2",
        show_row_names = F,
        show_column_names = F)


#Final output
All %>% print()
One %>% print()
```


