---
title: "R Notebook"
output: html_notebook
---

```{r}
library(data.table)
library(tidyverse)
library(dplyr)
library(readxl)

library(Seurat)
library(patchwork)
library(DoubletFinder)

library(ggplot2)
library(gridExtra)
library(ggExtra)
library(viridis)
library(Matrix)
library("ggsci")
library(cowplot)
library(platetools)
library(scales)
library("gridExtra")
library(ggpubr)
library(RColorBrewer)
library(tidyr)
library(dplyr)
library(ggforce)

#To add filename
read_plus <- function(flnm) {
    read.csv(flnm, header = TRUE, na.strings=c("", " ")) %>% 
        mutate(filename = flnm)
}

# Def Paths
Input_Dir <- "C:/Users/Marthe Laisne/Desktop/FlashSeq_MultiOneCell/Data/MiniBulk_FullTranscriptome/N320/"
Output_Dir <- "C:/Users/Marthe Laisne/OneDrive - INSTITUT CURIE/m_laisne/Publications/2026_Laisne/Scripts/Output/MiniBulkRNA/"
```


# Pre-processing of the transcriptome dataset:
Raw FLASH sequencing files were processed using the scRNA-SmartSeq3 Nextflow pipeline developed by the Curie Institute (https://github.com/bioinfo-pf-curie/scRNA-SmartSeq3). Briefly, UMIs were extracted from tagged reads using umi-tools (v1.1.1), 3' linkers and poly(A) tails were trimmed with cutadapt (v3.3), and reads were aligned to the mouse reference genomes (mm10) using STAR (v2.7.8a). UMI-count matrices were used for all downstream analyses.


## Metadata construction with lineage barcode addition
```{r}
#Load Data
Input_Data= list.dirs(Input_Dir)[-1]
Input_Data <- Input_Data[grep("*10XlikeMatrix_umi", Input_Data)]

rna.data <- Read10X(data.dir =Input_Data)
rna_seurat_umi <- CreateSeuratObject(counts = rna.data,
                             min.cells = 1,
                             min.features = 100,
                             project = "mouse_MiniBulk")

rna_seurat_umi@meta.data$SampleName <- colnames(rna_seurat_umi)
```


Expression matrices were processed in R (v4.4.1) using the Seurat (v5.1) package.

Lineage barcode information, including barcode identity, relative barcode frequencies and corresponding Pgen values, was added to the objects for downstream analyses.

```{r}
#Load BCM1 information (post QC)
Res_dir <- "C:/Users/Marthe Laisne/Desktop/FlashSeq_MultiOneCell/Results/Fig3_MiniBulk_Igor/"

AllBCM1_R2umi_QC4 <- qs::qread(paste0(Res_dir, "AllBCM1_R2umi_QC4_Pgen.qs"))
head(AllBCM1_R2umi_QC4)
```
```{r}
#Match plate information
AllBCM1_R2umi_QC4 <- AllBCM1_R2umi_QC4 %>%
  mutate(plaque = case_when(plaque %in% c("1","2") ~ paste0("LatePlq", plaque),
                            plaque == "Early1"~"EarlyPlq1",
                            plaque == "Early2"~"EarlyPlq2",
                            plaque == "Early3"~"EarlyPlq3",

                            TRUE ~ plaque))


Summary_BCM1_perwell <-  AllBCM1_R2umi_QC4 %>%
  subset(QC4_LV == "Pass") %>%
  subset(Pgen  <= 10**-4.5) %>%
  group_by(samplename, BCM_LVmerged) %>%
  summarise(n_umi_distinct = sum(n_umi_distinct), .groups = "drop"
  ) %>%
  left_join(AllBCM1_R2umi_QC4 %>% 
              dplyr::select(samplename, plaque, well, mouseV2,
TypeSample, Ncell, Run, SampleBio, Barcode_sharing_mice_N, BCM_LVmerged)) %>%
  distinct()

Summary_BCM1_perwell %>% head
```
```{r}
#Match well information: For eg, "A1" will be replaced by "A01", "C12" remain unchanged

Summary_BCM1_perwell <- Summary_BCM1_perwell %>%
  mutate( well = str_replace(well, 
                            "^([A-Z])(\\d)$", 
                             "\\10\\2") )

Summary_BCM1_perwell %>% head
```



```{r}
summary_grouped <- Summary_BCM1_perwell %>%
  
#Group by plate and well
  group_by(plaque, well) %>%
  
# Build a summary table containing information for all barcodes detected by MiniBulk (aka well)
  summarise({
    
# Aggregate UMI counts by barcode
    df <- tibble(
      BCM_LVmerged,
      n_umi_distinct,
      Barcode_sharing_mice_N
    ) %>%
      group_by(BCM_LVmerged) %>%
      dplyr::summarise(

        n_umi_distinct = sum(n_umi_distinct),  # Total number of UMIs associated with each barcode
        Barcode_sharing_mice_N = dplyr::first(Barcode_sharing_mice_N),  # Keep barcode sharing information
        .groups = "drop"
      ) %>%

 # Compute the fraction of total UMIs represented by each Clone
      mutate(
        fraction = n_umi_distinct / sum(n_umi_distinct)
      )

    tibble(
      BCM_LVmerged_list = list(df$BCM_LVmerged),
      n_umi_list = list(df$n_umi_distinct),
      sharing_list = list(df$Barcode_sharing_mice_N),
      fraction_list = list(df$fraction),

      n_barcode = nrow(df), # Number of distinct barcodes detected in the well
      max_barcode_fraction = max(df$fraction),  # Fraction of UMIs corresponding to the dominant Clone
      BCM_major = df$BCM_LVmerged[which.max(df$n_umi_distinct)] # Identity of the dominant Clone
    )

  }, .groups = "drop")

summary_grouped %>% tail; summary_grouped$BCM_LVmerged_list[1]
```
Load metadata

```{r}
meta <- read_excel("C:/Users/Marthe Laisne/OneDrive - INSTITUT CURIE/m_laisne/Publications/2026_Laisne/Tables/MetaData_MiniBulk.xlsx")
head(meta)
```

Merge with 
```{r}
merged_meta <- meta %>%
  left_join(summary_grouped,
            by = c( "Plate_Source" = "plaque" ,
                   "Well_Source" = "well" ))

head(merged_meta, 15)
```


Load data


```{r}
Input_Data= list.dirs(Input_Dir)[-1]
Input_Data <- Input_Data[grep("*10XlikeMatrix_umi", Input_Data)]

rna.data <- Read10X(data.dir =Input_Data)
rna_seurat_umi <- CreateSeuratObject(counts = rna.data,
                             min.cells = 1,
                             min.features = 100,
                             project = "mouse_MiniBulk")

rna_seurat_umi@meta.data$SampleName <- colnames(rna_seurat_umi)
```


Merge samples: 
```{r}
tmp <- left_join(rna_seurat_umi@meta.data,
                            merged_meta,
                 by = c("SampleName" = "SampleName"))

rownames(tmp) <- tmp$SampleName

rna_seurat_umi@meta.data <- tmp

rna_seurat_umi@meta.data$BioSample <- paste0(rna_seurat_umi@meta.data$Mouse, "_", rna_seurat_umi@meta.data$Condition)
```



##QC

Samples were retained if they had over 1,000 UMIs. The top 5% of cells with the highest coverage and gene counts were excluded as outliers. Cells with >10% mitochondrial reads and >25% ribosomal reads were also removed. 


```{r}
            x1 = data.frame(nGene = rna_seurat_umi$nCount_RNA)
            x1$BioSample = rna_seurat_umi@meta.data$BioSample
            x1$Stage = rna_seurat_umi@meta.data$Stage   
            colnames(x1) = c("nGene", "BioSample", "Stage")

min.features.thr <- log10(10**5)
selected_cells <- which(x1$nGene >  10^min.features.thr) %>% length
selected_cells

p <- ggplot(x1, aes(x = BioSample, y=nGene , fill=BioSample)) + 
  geom_violin(color = "black", position = position_dodge(width = 0.7)) + 
  geom_boxplot(position = position_dodge(width = 0.7), width = 0.1, alpha = 0.9, fill = "white")+
  geom_point(position = position_jitterdodge(jitter.width = 0.1,dodge.width = 0.7),  size = 4)+
  geom_hline(yintercept = 10**3)+
 
  theme_classic() + 
  scale_shape_manual(values = c(18, 19))+
  xlab("log10(Number of RNA+1)") +
  scale_color_manual(values = brewer.pal(12, "Paired"))+
  scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
              labels = trans_format("log10", math_format(10^.x)))

print(p)
```


```{r}

rna_seurat_umi[["percent.mito"]] <- PercentageFeatureSet(rna_seurat_umi, pattern = "^mt-")

rna_seurat_umi[["percent.ribo"]] <- PercentageFeatureSet(rna_seurat_umi, pattern = "^Rp[sl]")

            x1$Mito = rna_seurat_umi@meta.data$percent.mito
            x1$Ribo = rna_seurat_umi@meta.data$percent.ribo


pmito <- ggplot(x1, aes(x = Mito, y=nGene, fill=BioSample)) + 
  geom_point(position = position_jitterdodge(jitter.width = 0.1,dodge.width = 0.7),  size = 4)+
  geom_vline(xintercept = 10)+
 
  theme_classic() + 
  xlab("Fraction mitochondrial gene (%)") +
  scale_color_manual(values = brewer.pal(12, "Paired"))+
  scale_shape_manual(values = c(18, 19))+
  scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
              labels = trans_format("log10", math_format(10^.x)))+
  scale_x_continuous(limits = c(0, 100))

pRibo <- ggplot(x1, aes(x = Ribo, y=nGene, fill=BioSample)) + 
  geom_point(position = position_jitterdodge(jitter.width = 0.1,dodge.width = 0.7),  size = 4)+
  geom_vline(xintercept = 25)+
 
  theme_classic() + 
  xlab("Fraction Ribosomal gene (%)") +
  scale_color_manual(values = brewer.pal(12, "Paired"))+
    scale_shape_manual(values = c(18, 19))+
  scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
              labels = trans_format("log10", math_format(10^.x)))+
  scale_x_continuous(limits = c(0, 100))


pmito+pRibo
```
```{r}
rna_seurat_umi <- subset(rna_seurat_umi,  nCount_RNA > 10**3 & percent.ribo < 25 & percent.mito < 10 )
```





Filtered gene-cell barcode matrices were normalized individually using the "SCTransform" method (ref).

# Pre-tumoral PCA space

```{r}
#Subset Pre-tumoral minibulk only (remove Tumor samples)
rna_seurat_umi_el <- subset(rna_seurat_umi, 
                            CellType == "MixedGFPpos")

rna_seurat_umi_el$BioSample <- paste0(rna_seurat_umi_el$Mouse,
                                      "_", rna_seurat_umi_el$Gland)


#Annotation 
rna_seurat_umi_el@meta.data <- rna_seurat_umi_el@meta.data %>%
  mutate(Juxta_Not = case_when(Plate_Source %in% c("LatePlq1", "LatePlq2") & 
                                 CellType != "Tum" & 
                                 BioSample !=  c("9341_G4") ~ "Jx",
                               CellType == "Tum" ~ "Tm",
                               TRUE ~ "Gl"))
```

```{r}
#Normalize & RD
rna_seurat_umi_el <- NormalizeData(rna_seurat_umi_el)
rna_seurat_umi_el <- FindVariableFeatures(rna_seurat_umi_el)
rna_seurat_umi_el <- ScaleData(rna_seurat_umi_el)
rna_seurat_umi_el <- RunPCA(rna_seurat_umi_el, 
                            features = VariableFeatures(rna_seurat_umi_el), 
                            npcs = 50)

```

#Subsampling PCA and projection
```{r}
meta <- rna_seurat_umi_el@meta.data # Metadata extraction

# -------------------------------
# 1. Remove outlier mini-bulk samples
# -------------------------------
# These samples are excluded based on known quality or biological issues (number of cells)

id_rm <- which(rna_seurat_umi_el$SampleName %in% c("p1-E15-9826-G4-1c-EarlyPlq3-MiniBulk",                                                  "p1-O11-9891-G4-23c-EarlyPlq2-MiniBulk",                                                   "p1-P11-9873-D3-723c-EarlyPlq2-MiniBulk",                                                    "p1-H05-9341-Gl-100c-LatePlq1-MiniBulk"))

# Keep remaining cells
cells_keep <- setdiff(colnames(rna_seurat_umi_el),
                      rna_seurat_umi_el$SampleName[id_rm])

meta_sub <- meta[cells_keep, ]

# -------------------------------
# 2. Balanced subsampling across mice
# -------------------------------

# Minimum number of cells per mouse (used for balancing)
n_min <- median(table(meta_sub$Mouse))

# subsample 
set.seed(1)
cells_sub <- c()
for (i in unique(meta_sub$Mouse)) {
  
  # number of cells available for this mouse
  Sample_i <- nrow(subset(meta_sub, Mouse == i) )
  
    # random subsampling per mouse, capped by n_min
  cells_sub <- c(cells_sub,
                subset(meta_sub, Mouse == i) %>% rownames()%>%
                  sample(.,size = min(Sample_i, n_min))
  )
}

# Sanity checks
table(meta_sub[cells_sub, "Mouse"])
table(meta_sub[cells_sub, "BioSample"])


# -------------------------------
# 3. PCA on balanced subsample
# -------------------------------

mat_sub <- GetAssayData(rna_seurat_umi_el, slot = "data")[
  VariableFeatures(rna_seurat_umi_el),
  cells_sub] %>% t()



# Centering (store parameters for projection)
center <- colMeans(mat_sub)

# Scale without variance normalization (preserve gene scale)
mat_sub_scaled <- scale(mat_sub, center = center, scale = F)#scale_sd)

# PCA computation
pca <- prcomp(mat_sub_scaled, center = F, scale. = F, rank. = 50)


# -------------------------------
# 4. Projection of all cells onto PCA space
# -------------------------------

mat_all <- GetAssayData(rna_seurat_umi_el, slot = "data")[
  VariableFeatures(rna_seurat_umi_el),
  cells_keep
] %>% t()


# Apply identical centering transformation
mat_all_scaled <- scale(mat_all, center = center, scale = F)# scale_sd)


# Projection onto PCA loadings
# projection = X * rotation
emb_all <- mat_all_scaled %*% pca$rotation


# -------------------------------
# 5. Reintegrate excluded cells as NA embeddings
# -------------------------------

cells_all <- colnames(rna_seurat_umi_el)

emb_full <- matrix(
  NA,
  nrow = length(cells_all),
  ncol = ncol(emb_all)
)

rownames(emb_full) <- cells_all
colnames(emb_full) <- colnames(emb_all)

emb_full[rownames(emb_all), ] <- emb_all


# -------------------------------
# 6. Store "balanced" PCA as a Seurat reduction object
# -------------------------------

rna_seurat_umi_el[["balanced_pca"]] <- CreateDimReducObject(
  embeddings = emb_full,
  loadings = pca$rotation,
  key = "BPC_",
  assay = DefaultAssay(rna_seurat_umi_el)
)

# Visualization
DimPlot(rna_seurat_umi_el, reduction = "balanced_pca", group.by = "BioSample")
DimPlot(rna_seurat_umi_el, reduction = "balanced_pca", group.by = "Plate_Source")
FeaturePlot(rna_seurat_umi_el, reduction = "balanced_pca", "CellNumber",max.cutoff = 100)

```
```{r}
# -------------------------------
# 7. Variance explained by PCs
# -------------------------------

var_explained <- (pca$sdev^2) / sum(pca$sdev^2)
percent_var <- var_explained * 100
percent_var[1:10]
```








##Correlation analysis between PCA space and age
```{r}
# -------------------------------
# 1. Avoid pseudo-replication by averaging per biological sample
# -------------------------------

pca_df_avg <- pca_df %>%
  group_by(BioSample) %>%
  summarise(
    # Average PCA coordinates across cells within each biosample
     across(starts_with("BPC_"), mean, na.rm = TRUE),
    Age = mean(Age, na.rm = TRUE), #Age 
    Juxta_Not = unique(Juxta_Not) # Group identity
  )

# -------------------------------
# 2. Pearson correlation between PCs and age
# -------------------------------

cor_results <- lapply(1:10, function(i) {
 # pc <- paste0("WPCA_", i)
   pc <- paste0("BPC_", i)
  test <- cor.test(pca_df_avg[[pc]], pca_df_avg$Age, method = "pearson")
  
  data.frame(
    PC = pc,
    r = test$estimate,
    pval = test$p.value
  )
}) %>% bind_rows()

# Multiple testing correction (BH)
cor_results$padj <- p.adjust(cor_results$pval, method = "BH")

# Significant correlations
cor_results %>% 
  filter(padj < 0.05 & abs(r) > 0.4)
```
```{r}
# -------------------------------
# 3. Scatter plots for each PC vs Age
# -------------------------------


plots <- lapply(1:10, function(i) {
 # pc <- paste0("WPCA_", i)
   pc <- paste0("BPC_", i)
  
  ggplot(pca_df_avg, aes_string(x = "Age", y = pc, fill = "BioSample")) +
    geom_point(alpha = 0.6,  size = 2, 
               aes(shape = Juxta_Not)) +
    geom_smooth(method = "lm", se = TRUE) +
      scale_shape_manual(values = c(21, 24))+
    stat_regline_equation(label.x = min(pca_df_avg$pc),
                        label.y = max(pca_df_avg$Age)) +
  stat_cor(label.x = min(pca_df_avg$pc),
           label.y = max(pca_df_avg$Age) * 0.1,
           method = "pearson") +
    ggtitle(paste0(pc, " vs Age")) +
    theme_classic()+
    scale_fill_manual(values = sample(colorRampPalette(brewer.pal(8, "Spectral"))(30)))+
    theme(legend.position = "none")
})


wrap_plots(plots, ncol = 10)

# -------------------------------
# 3.Global correlation summary plot:
# -------------------------------
#Extended Fig4a
ggplot(cor_results, aes(x = PC, y = r)) +
  geom_point(size = 3) +
  geom_hline(yintercept = c(0.5, 0 ,-0.5), linetype = "dashed") +
  theme_classic() +
  scale_x_discrete(limits = c(paste0("BPC_", 1:10)))+
  ylab("Pearson correlation (r)") +
  ggtitle("Correlation of PCs with Age")
```
##F Linear model: definition of pseudotime axis

```{r}
# -------------------------------
# 4. Linear regression defining a pseudotime axis from PCA space
# -------------------------------
fit <- lm(Age ~ BPC_1 +  BPC_3 #+ BPC_4
          , data = pca_df)

summary(fit)
```


```{r}
# -------------------------------
# 5. Project regression-based pseudotime onto all cells
# -------------------------------
cells_all <- colnames(rna_seurat_umi_el)
fit_full <- matrix(  NA,  nrow = length(cells_all),  ncol = 1)
rownames(fit_full) <- cells_all
colnames(fit_full) <- "pseudotime"


# Predict pseudotime from PCA-derived model
fit_full[names(predict(fit)), ] <- predict(fit)

pca_df$pseudotime <- fit_full
```


```{r}
#Store pseudotime in Seurat object
rna_seurat_umi_el@meta.data$pseudotime <- fit_full
```


Model coefficients
```{r}
#age = β0 + β1*WPC1 + β2*WPC5
coef(fit)
```

```{r}
# Group according to pre-tumoral pseudotime values
rna_seurat_umi_el@meta.data <- rna_seurat_umi_el@meta.data %>% 
  mutate(Group_PT = case_when(pseudotime < 2 ~"E1", 
                              pseudotime > 5.5 ~"L", 
                              TRUE ~"E2"))

```



# Dispersion metrics (annotated)
##Shannon index computation (clonal diversity, by Minibulk)

```{r}
#Extract PCA coordinates
pca <- as.data.frame(
  Embeddings(rna_seurat_umi_el, reduction = "balanced_pca")[, c(1,5)]
)

pca$BioSample <- rna_seurat_umi_el$BioSample

# Merge PCA coordinates with metadata
mm_meta <- rna_seurat_umi_el@meta.data
mm_meta$BioSample = rownames(mm_meta)

meta_plot <- mm_meta |>
  left_join(pca, by = "BioSample")

# -------------------------------
# 2. Compute Shannon diversity index per mini-bulk
# -------------------------------
# Based on barcode frequency distribution (fraction_list)

meta_plot <- meta_plot %>%
  rowwise() %>%
  mutate(
    shannon_barcode = if(!is.null(fraction_list) && length(fraction_list) > 0) {
      p <- unlist(fraction_list)
      p <- p[p > 0] # avoid log(0)
      -sum(p * log(p)) # Shannon entropy
    } else {
      NA_real_
    }
  ) %>%  ungroup()
```

```{r}
# Shannon vs number of cells per mini-bulk
ggplot(meta_plot, aes(x = CellNumber, y = shannon_barcode))+
  geom_jitter(size =3, hape = 21, alpha = 0.5)+
  scale_x_continuous(limits = c(0, 100))+
  theme_classic()
```
```{r}
# Shannon vs age
ggplot(meta_plot, aes(x = Age, y = shannon_barcode))+
  geom_jitter(size =3, shape = 21, alpha = 0.5)+
   geom_smooth(method = "lm", se = FALSE, color = "black") +
  stat_regline_equation(label.x = min(meta_plot$Age),
                        label.y = max(meta_plot$shannon_barcode)) +
  stat_cor(label.x = min(meta_plot$Age),
           label.y = max(meta_plot$shannon_barcode) * 0.9,
           method = "pearson")+
  theme_classic()
```


Group comparison of clonal diversity
```{r}
ggplot(meta_plot, aes(x = Group_PT, fill = Group_PT, y = shannon_barcode))+
  geom_boxplot(outliers = F)+
  geom_point(size =3, alpha = 0.5,
             position = position_jitter(width = 0.1),
             aes(shape = Juxta_Not))+
  theme_classic()+
  scale_x_discrete(limits = c("E1",  "E2", "L"))+
  scale_shape_manual(values = c(21, 24))+
  scale_fill_manual(values = colorRampPalette(brewer.pal(4, "Spectral"))(4)[c(2, 4, 1, 3)])+
  theme(legend.position = "none")

```
Statistical testing
```{r}
anova_model <- aov(shannon_barcode~Group_PT, data = meta_plot)
summary(anova_model)

tukey_res <- TukeyHSD(anova_model)
tukey_res
```

Store Shannon index in Seurat object
```{r}
rna_seurat_umi_el$shannon_barcode_index <- meta_plot$shannon_barcode
```


Relationship with pseudotime
```{r}
plot(rna_seurat_umi_el$shannon_barcode_index~rna_seurat_umi_el$pseudotime)


data_plot <- data.frame(shannon = rna_seurat_umi_el$shannon_barcode_index,
                        pseudotime = rna_seurat_umi_el$pseudotime)

ggplot(data_plot, aes(x = pseudotime, y = shannon))+
  geom_point(shape = 21)+
  geom_smooth(method = "lm")+
  theme_classic()
```


##Distance to biological sample centroid (sample dispersion)
```{r}
pca_df$Group_PT <- rna_seurat_umi_el$Group_PT
variance <- rna_seurat_umi_el@reductions$balanced_pca@stdev[c(1, 3)] /100

# -------------------------------
# 1. PCA-based centroid computation per BioSample
# -------------------------------

centroids <- pca_df %>%
  dplyr::group_by(BioSample) %>%
  dplyr::summarise(across(paste0("BPC_", c(1, 3)), ~ mean(.x, na.rm = TRUE), .names = "c{.col}"))
head(centroids)

# -------------------------------
# 2. Euclidean distance to sample centroid
# -------------------------------

df_dist <- pca_df %>%
  left_join(centroids, by = c("BioSample")) %>%
  group_by(BioSample) %>%
  dplyr::rowwise() %>%
  dplyr::mutate( dist_to_centroid = sqrt(sum(
      (dplyr::c_across(paste0("BPC_", c(1, 3))) -
       dplyr::c_across(paste0("cBPC_", c(1, 3))))^2
    ))
  ) %>%
  ungroup()


#Visualization of dispersion 
base_pal <- c( "grey90","#C0D69F", "#61980E", "grey10")
extended_pal <- rev(colorRampPalette(base_pal)(29))

ggboxplot <- ggplot(subset(df_dist, is.na(BioSample)==F),
       aes(x = Group_PT, y = dist_to_centroid)) +
  geom_boxplot(outliers = F) +
  geom_jitter(aes(fill = Group_PT, shape = Juxta_Not), width = 0.2, alpha = 0.5, size = 1) +
  theme_classic(base_size = 14) +
  labs(y = "Distance to BioSample-centroid, PC1 & PC3", x = "") +
 scale_fill_manual(values = extended_pal) +  
  theme(legend.position = "none")+
  scale_shape_manual(values = c(21, 24) )+
  scale_fill_manual(values = colorRampPalette(brewer.pal(4, "Spectral"))(4)[c(2, 4, 1, 3)])


ggboxplot
```

```{r}
anova_model <- aov(dist_to_centroid~Group_PT, 
                   data = subset(df_dist, is.na(BioSample)==F))
summary(anova_model)

tukey_res <- TukeyHSD(anova_model)
tukey_res
```

Add to metadata
```{r}
rna_seurat_umi_el$Distance_to_Biosample_Centroid <- df_dist$dist_to_centroid
```


#Identification of pathways associated with pseudotime

1. Loading gene signatures and preparing pathway sets
```{r}
# -------------------------------
# Load curated mammary gland gene signatures
# -------------------------------

Signature_ID_CreN_10x <- read.delim("C:/Users/Marthe Laisne/Desktop/BI_references_signatures_annotations/Mammarygland_Signatures_Custom/SignatureCreN_scRNA10x_Epithelial_poles.txt")

LAlv_Chen2021  <- read.delim("C:/Users/Marthe Laisne/Desktop/BI_references_signatures_annotations/Mammarygland_Signatures_cellTypes/LAlv_Chen2021.txt")
LAlv_Chen2021$Type <- rep("Alv", nrow(LAlv_Chen2021))
colnames(LAlv_Chen2021) <- c("Gene", "Type")
LAlv_Chen2021 <- LAlv_Chen2021[, c(2,1)]
LAlv_Chen2021 <- rbind(LAlv_Chen2021, data.frame(Type = "Alv", Gene = "Spp1"))

Signature_ID_CreN_10x <- rbind(Signature_ID_CreN_10x,
                               LAlv_Chen2021)

# Group genes into signature lists
sig_list <- Signature_ID_CreN_10x %>%
  group_by(Type) %>%
  summarise(genes = list(unique(Gene))) %>%
  deframe()

# Restrict to genes present in the dataset
sig_list <- lapply(sig_list, function(g) {
  intersect(g, rownames(rna_seurat_umi_el))
})

```

2. Compute module scores per signature
```{r}
# -------------------------------
# Add Seurat module scores for each gene set
# -------------------------------

rna_seurat_umi_el <- AddModuleScore(
  object = rna_seurat_umi_el,
  features = sig_list,
  name = names(sig_list)
)

# Rename score columns for clarity
new_names <- paste0(names(sig_list), "_Score")

colnames(rna_seurat_umi_el@meta.data)[(ncol(rna_seurat_umi_el@meta.data)-3):ncol(rna_seurat_umi_el@meta.data)] <- new_names
```

3. Linking pseudotime to PCA loadings (pathway ranking strategy)
```{r}
# Extract PCA loadings from balanced PCA
loadings <- Loadings(rna_seurat_umi_el, reduction = "balanced_pca")

# Pseudotime
pt_scores <- rna_seurat_umi_el$pseudotime
names(pt_scores) <- colnames(rna_seurat_umi_el)

# Linear model coefficients defining pseudotime axis
pt_loadings <- coefficients(fit)[1] + 
  coefficients(fit)[2]*loadings[,1] +
  coefficients(fit)[3] *loadings[,3] 
pt_loadings <- sort(pt_loadings, decreasing = TRUE)

```

4. Gene set enrichment analysis (FGSEA)
```{r}
library(msigdbr)
library(fgsea)

# Hallmark gene sets (Mouse)
hallmark <- msigdbr(species = "Mus musculus", category = "H")

pathways_hallmark <- hallmark %>% 
  split(x = .$gene_symbol, f = .$gs_name) 

# Combine curated signatures + hallmark sets
pathways <- c(pathways_hallmark, sig_list)

# Run fgsea using ranked gene list
fgsea_res <- fgsea(
  pathways = pathways,
  stats = pt_loadings
)

# Filter significant pathways
fgsea_sig_pt <- fgsea_res |> 
  dplyr::filter(padj < 0.01) |> 
  arrange(desc(NES))

fgsea_sig_pt
fgsea_sig_pt[, c("pathway","padj", "NES")]

```

5. Export results
```{r}
write.table(fgsea_sig_pt[, c("pathway","padj", "NES")], paste0(Output_Dir, "fgsea_sig_pt.txt"), quote = F, sep = "\t")
```


```{r}
library(openxlsx)

gsea_export <- copy(fgsea_sig_pt)

gsea_export[, leadingEdge := sapply(leadingEdge, paste, collapse = ";")]

write.xlsx(gsea_export, "C:/Users/Marthe Laisne/Desktop/FlashSeq_MultiOneCell/Results/Fig5_MiniBulkRNA/gsea_sig_pt.xlsx")
```

6. Compute module scores for significant pathways
```{r}
sig_pathways <- fgsea_sig_pt$pathway 

for (pw in sig_pathways) {
  genes <- pathways[[pw]]
  
  rna_seurat_umi_el <- AddModuleScore(
    rna_seurat_umi_el,
    features = list(genes),
    name = pw
  )
}
```

## Pathway score heatmaps



```{r}
#Extract score
meta <- rna_seurat_umi_el@meta.data
colnames(meta) <- gsub("1$", "", colnames(meta))

meta <- meta %>% 
  mutate(Group_PT = case_when(pseudotime < 2 ~"E1", pseudotime > 5.5 ~"L", TRUE ~"E2"))

score_mat <- meta |>
  dplyr::select(dplyr::any_of(c(sig_pathways,
                                "LP_Score", "Alv_Score", "Basal_Score", "Luminal H-S_Score")))



score_mat <- score_mat[rownames(meta), ]


#Annot tumor bearing
meta$Tm_Bearing <- meta$Juxta_Not
```

```{r}
library(ComplexHeatmap)
library(circlize)

# palette Age
age_col_fun <- colorRamp2(
  seq(min(meta$Age), max(meta$Age), length.out = 4),
  c("white", "#C0D69F", "#61980E", "grey10"))

pseudoT_col_fun <- colorRamp2(
  seq(min(meta$pseudotime, na.rm = T)+1, max(meta$pseudotime, na.rm = T)+1, length.out = 5),
  c("white", "#9f97c8", "#514495", "#231e47", "grey10"))

Clusters_col <- setNames(  colorRampPalette(brewer.pal(4, "Spectral"))(4)[1:3],
  levels(factor(meta$Group_PT)))
  

# palette Shannon - distance
shannon_col_fun <- colorRamp2(
  seq(min(meta$shannon_barcode_index, na.rm = T), max(meta$shannon_barcode_index, na.rm = T), length.out = 4),
  c("white", "grey70", "grey30", "grey10"))



#  Tm_Bearing (facteur)
tm_levels <- unique(meta$Tm_Bearing)

tm_cols <- setNames(  c( "grey10", "grey50", "grey95")[seq_along(tm_levels)],   tm_levels)

# --- Mouse (facteur)
mouse_levels <- unique(meta$Mouse)

mouse_cols <- setNames(  colorRampPalette(brewer.pal(8, "Spectral"))(length(mouse_levels)),
  mouse_levels)

# --- Annotation heatmap
age_anno <- HeatmapAnnotation(
  pseudoT = meta$pseudotime,
  Shannon = meta$shannon_barcode_index,
  Cluster = meta$Group_PT,
  Age = meta$Age,
  Tm_Bearing = meta$Tm_Bearing,
 
  Mouse = as.character(meta$Mouse),
  
 
  col = list(
    Shannon = shannon_col_fun,
    Cluster = Clusters_col,
    Tm_Bearing = tm_cols,
    Mouse = mouse_cols,
    pseudoT = pseudoT_col_fun,
    Age = age_col_fun),
  
  annotation_name_side = "left")



############# Heatmap
heatmap_palette <- c( rev( c("grey99","#E9F2FC", "#C1DAF6",  "#99C1F0", "#71A9EA", "#4991E4", "#3283E0", "#1B63B6", "#154E8E", "#0F3866") )
  ,"grey99", "#FCE9F2", "#F6C1DA", "#F099C1", "#EA71A9",  "#E44991", "#E03283", "#B61B63", "#8E154E",  "#660F38")

heatmap_col_fun <- colorRamp2(
  seq(-2, 2, length.out = length(heatmap_palette)),
  heatmap_palette)


Heatmap(
  t(scale(score_mat[,  ], center = TRUE, scale = TRUE)),
  name = "Pathway score",
  top_annotation = age_anno,
  col = heatmap_col_fun,
  column_order = order(meta$pseudotime), # Order according to pseudot
  cluster_columns = F,
  cluster_rows = TRUE,
  clustering_distance_rows = "euclidean",
  clustering_method_rows = "ward.D2",
  show_column_names = FALSE,
  row_names_gp = gpar(fontsize = 8) )

```
```{r}
pdf(paste0(Output_Dir, "/Heatmap_Noscale_Mouse_BalancedPC.pdf")  , height =4, width = 10, onefile = T)
Heatmap(
  t(scale(score_mat[is.na(meta$pseudotime)==F,  ], center = TRUE, scale = TRUE)),
  name = "Pathway score",
  top_annotation = age_anno[is.na(meta$pseudotime)==F],
  col = heatmap_col_fun,
  cluster_columns = T,
    clustering_distance_columns  = "euclidean",
  clustering_method_columns  = "ward.D2",
  cluster_rows = TRUE,
  clustering_distance_rows = "euclidean",
  clustering_method_rows = "ward.D2",
  show_column_names = FALSE,
  row_names_gp = gpar(fontsize = 8) )

Heatmap(
  t(scale(score_mat[ is.na(meta$pseudotime)==F,], center = TRUE, scale = TRUE)),
  name = "Pathway score",
  top_annotation = age_anno[is.na(meta$pseudotime)==F],
  col = heatmap_col_fun,
  column_order = order(meta$pseudotime[is.na(meta$pseudotime)==F]), # Order according to pseudot
  cluster_columns = F,
  cluster_rows = TRUE,
  clustering_distance_rows = "euclidean",
  clustering_method_rows = "ward.D2",
  show_column_names = FALSE,
  row_names_gp = gpar(fontsize = 8) )


dev.off()

```

# Identification of differentially expressed genes


```{r}
#-----------------------------
# 1. Define comparisons
#-----------------------------
comparisons <- list(
  list(name = "Late_vs_(E1+E2)", ident.1 = "L",  ident.2 = c("E1", "E2"), group = "Group_PT"),
  list(name = "E2_vs_E1",      ident.1 = "E2", ident.2 = "E1",         group = "Group_PT"),
  list(name = "Late_vs_E2",    ident.1 = "L",  ident.2 = "E2",        group = "Group_PT")
)

#-----------------------------
# 2. Loop DEG analysis
#-----------------------------
deg_results <- list()

for (cmp in comparisons) {

  message("Running: ", cmp$name)

  deg <- FindMarkers(
    object = rna_seurat_umi_el,
    ident.1 = cmp$ident.1,
    ident.2 = cmp$ident.2,
    group.by = cmp$group,
    logfc.threshold = 0,
    min.pct = 0.1,
    test.use = "wilcox"
  )

  deg$gene <- rownames(deg)

  deg <- deg %>%
    mutate(
      comparison = cmp$name,
      signif = p_val_adj < 0.05,
      FC = case_when(
        signif & avg_log2FC > 0 ~ "UP",
        signif & avg_log2FC < 0 ~ "DW",
        TRUE ~ "NS"
      )
    )

  deg_results[[cmp$name]] <- deg

  #-----------------------------
  # Export tables
  #-----------------------------
  write.csv(
    deg,
    file.path(Output_Dir, paste0("DEG_", cmp$name, ".csv")),
    row.names = FALSE
  )

  write.csv(
    deg %>% filter(p_val_adj < 0.05),
    file.path(Output_Dir, paste0("DEG_", cmp$name, "_sig.csv")),
    row.names = FALSE
  )

  #-----------------------------
  # Volcano plot (minimal)
  #-----------------------------
  p <- ggplot(deg, aes(avg_log2FC, -log10(p_val_adj))) +
    geom_point(aes(color = signif), alpha = 0.6) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
    theme_classic() +
    labs(title = cmp$name)
  
  print(p)

  ggsave(
    file.path(Output_Dir, paste0("Volcano_", cmp$name, ".pdf")),
    p, width = 4, height = 4
  )
}

#-----------------------------
# 3. Optional: merged table
#-----------------------------
deg_merged <- Reduce(function(x, y) full_join(x, y, by = "gene"), deg_results)
```

#MiniBulk Scoring of significant pathways

```{r}
# --------------------------------------------------------------
# UCell scoring function (Yuna's implementation)
# --------------------------------------------------------------


UCell_scores_per_cell <- function(raw_counts, sig) {
     # Keep only genes present in the dataset
    sig_genes <- dplyr::intersect(unique(unlist(sig)), rownames(raw_counts)) # restrict to genes found in matrix
    sig <- lapply(sig, function(x) dplyr::intersect(x, sig_genes))
    # Compute Ucell scores
    scores_ucell <- as.data.frame(UCell::ScoreSignatures_UCell(raw_counts, features = sig))
    return(scores_ucell)
}


# Normalization
rna_seurat_umi_el <- NormalizeData(rna_seurat_umi_el, 
                            normalization.method = "LogNormalize",
                            scale.factor = 10000)

ucell_mat_MiniBulk <- matrix(nrow = dim(rna_seurat_umi_el)[2])

# Score pathway signatures (excluding selected cell-state programs)
for (pw in sig_pathways[!(sig_pathways %in% c("LP", "Basal", "Luminal H-S"))]) {
  genes <- pathways[[pw]]
  ucell_mat_MiniBulk <- cbind(ucell_mat_MiniBulk,
                        UCell_scores_per_cell(GetAssayData(rna_seurat_umi_el, slot = "data"), 
                                      sig = list(genes))
  )

}
# Format matrix
ucell_mat_MiniBulk <- ucell_mat_MiniBulk[,2:ncol(ucell_mat_MiniBulk)]
colnames(ucell_mat_MiniBulk) <- (sig_pathways[!(sig_pathways %in% c("LP", "Basal", "Luminal H-S"))]
)

ucell_mat_MiniBulk %>% head
```


```{r}
# UMAP projection and metadata integration
dat <- ucell_mat_MiniBulk

dat$BPC1 <-Embeddings(rna_seurat_umi_el , "balanced_pca")[, 1]
dat$BPC3 <- Embeddings(rna_seurat_umi_el , "balanced_pca")[, 3]
dat$pseudotime <- rna_seurat_umi_el$pseudotime

head(dat)
```




## Pathways assignment (global reference)
Binarization and state calling:
  1. Scores are centered and scaled across the cohort
  2. A threshold is defined per pathway (mean + SD)
  3. Samples  are assigned to active/inactive states
  4. Maximum scoring pathway defines dominant state per Sample
```{r}
meta <- rna_seurat_umi_el@meta.data
meta <- meta[, !(colnames(meta) %in% c(sig_pathways, "state_max", "state_max_relev"))]
metadata_cells <- meta

nb_sds <- 1

ucell_mat_MiniBulk_global <- ucell_mat_MiniBulk
df_MiniBulk <- meta


ucell_mat_MiniBulk_global <- ucell_mat_MiniBulk_global[ rownames(df_MiniBulk), ] %>% t

# --------------------------------------------------------------
#1. Scores are centered and scaled across the cohort
# --------------------------------------------------------------

ucell_mat_MiniBulk <- ucell_mat_MiniBulk_global 
ucell_mat_MiniBulk_global <- t(scale(t(ucell_mat_MiniBulk_global), scale = T))
ucell_mat_MiniBulk_global[is.nan(ucell_mat_MiniBulk_global)] <- 0

# --------------------------------------------------------------
#2. A threshold is defined per pathway (mean + SD)
# --------------------------------------------------------------

thr_vect <- apply(ucell_mat_MiniBulk[,  drop = FALSE], 1, function(x) {
    tmp <- x[x > 0]
    if (length(tmp) == 0) return(0)  # sécurité
    return(mean(tmp) + nb_sds * sd(tmp))
})

ucell_mat_global_MiniBulk_binary <- sweep(ucell_mat_MiniBulk, 1, thr_vect, "-")

{
    list_mat <- list()
    list_attrib <- list()
    list_attrib_max <- list()
    cell_id <- c()

    for (sample_id in unique(df_MiniBulk$BioSample)) {

        # subset state activity matrix for the right sample
        idx <- which(df_MiniBulk$BioSample == sample_id)
        ucell_mat_sub <- ucell_mat_global_MiniBulk_binary[, idx, drop = FALSE]

        # --- Correction : If only one sample, force matrix  ---
        if (is.null(dim(ucell_mat_sub))) {
            ucell_mat_sub <- matrix(ucell_mat_sub, ncol = 1)
            colnames(ucell_mat_sub) <- colnames(ucell_mat_global_MiniBulk_binary)[idx]
            rownames(ucell_mat_sub) <- rownames(ucell_mat_global_MiniBulk_binary)
        }

        # binary attribution
        thr <- 0
        tmp_attrib <- apply(ucell_mat_sub, 1, function(x) x >= thr)
        list_attrib[[sample_id]] <- tmp_attrib

        cell_id <- c(cell_id, colnames(ucell_mat_sub))

        # save processed state matrix by sample
        list_mat[[sample_id]] <- ucell_mat_global_MiniBulk_binary[, idx, drop = FALSE]

        # maximum attribution
        ucell_mat_sub2 <- ucell_mat_global_MiniBulk_binary[, idx, drop = FALSE]

        if (ncol(ucell_mat_sub2) == 1) {
            # --- 1 sample ---
            x <- ucell_mat_sub2[, 1]
            tmp_max <- max(x, na.rm = TRUE)
            if (is.finite(tmp_max) && tmp_max > thr) {
                tmp <- names(which.max(x, na.rm = TRUE))[1]
            } else {
                tmp <- "Unknown"
            }
            tmp_attrib2 <- setNames(tmp, colnames(ucell_mat_sub2))

        } else {
            # --- >1 sample ---
            tmp_attrib2 <- apply(ucell_mat_sub2, 2, function(x) {
                tmp_max <- max(x, na.rm = TRUE)
                if (is.finite(tmp_max) && tmp_max > thr) {
                    tmp <- rownames(ucell_mat_sub2)[which.max(x)]
                    tmp[1]
                } else {
                    "Unknown"
                }
            })
        }

        list_attrib_max[[sample_id]] <- tmp_attrib2
    }

    df_attrib <- do.call(rbind, list_attrib)
    rownames(df_attrib) <- cell_id
    df_attrib <- df_attrib[colnames(ucell_mat_MiniBulk_global), ]

    vect_attrib <- do.call(c, list_attrib_max)
    names(vect_attrib) <- cell_id
    vect_attrib <- vect_attrib[colnames(ucell_mat_MiniBulk_global)]
}

df_MiniBulk <- cbind(df_MiniBulk, df_attrib)
df_MiniBulk$state_max <- vect_attrib

head(df_MiniBulk)
```

```{r}
dim(df_MiniBulk) ; dim(rna_seurat_umi_el)
```


# ######### SAVE & RELOAD ######### 
```{r}
qs::qsave(rna_seurat_umi_el, file = paste0(Output_Dir, "rna_seurat_umi_el.qs") )
qs::qsave(rna_seurat_umi, file = paste0(Output_Dir, "rna_seurat_umi.qs") )


#PCA parameters
qs::qsave( VariableFeatures(rna_seurat_umi_el), file = paste0(Output_Dir, "features_PCA_MiniBulk.qs") )
qs::qsave(  Loadings(rna_seurat_umi_el[["balanced_pca"]])[ VariableFeatures(rna_seurat_umi_el), ], 
            file = paste0(Output_Dir, "loading_PCA_MiniBulk.qs") )
qs::qsave( center,  file = paste0(Output_Dir, "centers_PCA_MiniBulk.qs") )


#Pathways assignment
qs::qsave(df_MiniBulk, file = paste0(Output_Dir, "Pathways_MiniBulk.qs") )
qs::qsave( pathways,  file = paste0(Output_Dir, "pathways.qs") )
qs::qsave( sig_pathways,  file = paste0(Output_Dir, "sig_pathways.qs") )

```


















