---
title: "R Notebook"
output: html_notebook
---
```{r}
library(data.table)
library(tidyverse)
library(dplyr)
library(readxl)

library(Seurat)
library(patchwork)
library(DoubletFinder)

library(ggplot2)
library(gridExtra)
library(ggExtra)
library(viridis)
library(Matrix)
library("ggsci")
library(cowplot)
library(platetools)
library(scales)
library("gridExtra")
library(ggpubr)
library(RColorBrewer)
library(tidyr)
library(dplyr)
library(ggforce)

# Def Paths
Input_Dir <- "C:/Users/Marthe Laisne/OneDrive - INSTITUT CURIE/m_laisne/Publications/2026_Laisne/Scripts/Output/MiniBulkRNA/"
Output_Dir <- "C:/Users/Marthe Laisne/OneDrive - INSTITUT CURIE/m_laisne/Publications/2026_Laisne/Scripts/Output/MiniBulkRNA_Figures/"
```




# ######### Vizualisation: ######### 

```{r}
rna_seurat_umi_el <- qs::qread( paste0(Input_Dir, "rna_seurat_umi_el.qs") )

cells_sub <- qs::qread( paste0(Input_Dir, "cells_sub.qs") )

sig_pathways <- qs::qread( paste0(Input_Dir, "sig_pathways.qs") )
pathways <- qs::qread( paste0(Input_Dir, "pathways.qs") )

df_MiniBulk <- qs::qread( paste0(Input_Dir, "Pathways_MiniBulk.qs") )

```

## QC: Ext. Fig2 c-e

Extended Fig2d:
```{r}
dat_plot <- rna_seurat_umi_el@meta.data


 ggplot(dat_plot,
         aes(x = CellNumber , y = log10(nCount_RNA))) +
    ggpointdensity::geom_pointdensity(size = 4) +
    scale_color_viridis_c(option = "magma") +
    theme_classic(base_size = 12) +
    guides(color="none")
```

Make dataset for vizu:
```{r}
pca_df <- Embeddings(rna_seurat_umi_el, "balanced_pca")[, 1:10] %>% as.data.frame()

pca_df$BioSample <- rna_seurat_umi_el@meta.data$BioSample
pca_df$Type <- rna_seurat_umi_el@meta.data$Stage
pca_df$Age <- rna_seurat_umi_el@meta.data$Age
pca_df$Ncell <- rna_seurat_umi_el@meta.data$CellNumber
pca_df$plate <- rna_seurat_umi_el@meta.data$Plate_Source
pca_df$CellType <- rna_seurat_umi_el@meta.data$CellType
pca_df$Nreads <- rna_seurat_umi_el@meta.data$nCount_RNA
pca_df$SampleName <- rna_seurat_umi_el@meta.data$SampleName

pca_df$Mouse <- rna_seurat_umi_el@meta.data$Mouse %>% as.character()

pca_df$Nclone <- rna_seurat_umi_el@meta.data$n_barcode
pca_df$MaxCloneFraction <- rna_seurat_umi_el@meta.data$max_barcode_fraction
pca_df$BCM_major <- rna_seurat_umi_el@meta.data$BCM_major
pca_df$Juxta_Not <-rna_seurat_umi_el@meta.data$Juxta_Not %>% as.character()
pca_df$pseudotime <-rna_seurat_umi_el@meta.data$pseudotime 

col_mouse <- sample(colorRampPalette(brewer.pal(8, "Spectral"))(30))
```

##Extended 4b-c. PCA


```{r}

l_plots <- list(
ggplot(pca_df[cells_sub, ], aes(x = BPC_1, y = BPC_3)) +
  geom_point(size = 4, alpha = 0.5,
             aes(shape = Juxta_Not, fill = BioSample)) +
  theme_classic() + 
  scale_shape_manual(values = c(21, 24))+
  scale_fill_manual(values = col_mouse)+
  labs(title = "Balanced PCA - BioSample - sub",   x = "BPC1", y = "BPC3")+
ggplot(pca_df, aes(x = BPC_1, y = BPC_3)) +
  geom_point(size = 4, alpha = 0.5,
             aes(shape = Juxta_Not, fill = BioSample)) +
  theme_classic() + 
  scale_shape_manual(values = c(21, 24))+
  scale_fill_manual(values = col_mouse)+
  labs(title = "Balanced PCA - BioSample",  x = "BPC_1", y = "BPC3"),

ggplot(pca_df[cells_sub, ], aes(x = BPC_1, y = BPC_3)) +
  geom_point(size = 4, alpha = 0.5,
             aes(shape = Juxta_Not, fill = Mouse)) +
  theme_classic() + 
  scale_shape_manual(values = c(21, 24))+
  scale_fill_manual(values = col_mouse)+
  labs(title = "Balanced PCA - Mouse - sub",   x = "BPC1", y = "BPC3")+
ggplot(pca_df, aes(x = BPC_1, y = BPC_3)) +
  geom_point(size = 4, alpha = 0.5,
             aes(shape = Juxta_Not, fill = Mouse)) +
  theme_classic() + 
  scale_shape_manual(values = c(21, 24))+
  scale_fill_manual(values = col_mouse)+
  labs(title = "Balanced PCA - Mouse",  x = "BPC_1", y = "BPC3"),



ggplot(pca_df[cells_sub, ], aes(x = BPC_1, y = BPC_3)) +
  geom_point(size = 4, alpha = 0.5,
             aes(shape = Juxta_Not, fill = plate)) +
  theme_classic() + 
  scale_shape_manual(values = c(21, 24))+
  scale_fill_manual(values = (colorRampPalette(brewer.pal(8, "Spectral"))(5)))+
  labs(title = "Balanced PCA - plate - sub",   x = "BPC1", y = "BPC3")+
ggplot(pca_df, aes(x = BPC_1, y = BPC_3)) +
  geom_point(size = 4, alpha = 0.5,
             aes(shape = Juxta_Not, fill = plate)) +
  theme_classic() + 
  scale_shape_manual(values = c(21, 24))+
  scale_fill_manual(values = (colorRampPalette(brewer.pal(8, "Spectral"))(5)))+
  labs(title = "Balanced PCA - plate",   x = "BPC_1", y = "BPC3"),


ggplot(pca_df[cells_sub, ], aes(x = BPC_1, y = BPC_3)) +
  geom_point(size = 4, alpha = 0.5,color = "black",
             aes(shape = Juxta_Not, fill = Age)) +
  theme_classic() + 
  scale_shape_manual(values = c(21, 24))+
  scale_fill_gradientn(  colours = c("white", "#C0D69F", "#61980E", "grey10") , 
                         na.value = "white", limits = c(0, 8)  )+
  labs(title = "Balanced PCA - Age - sub",   x = "BPC1", y = "BPC3")+
ggplot(pca_df, aes(x = BPC_1, y = BPC_3)) +
  geom_point(size = 4, alpha = 0.5,color = "black",
             aes(shape = Juxta_Not, fill = Age)) +
  theme_classic() + 
  scale_shape_manual(values = c(21, 24))+
  scale_fill_gradientn(  colours = c("white", "#C0D69F", "#61980E", "grey10") , 
                         na.value = "white", limits = c(0, 8)  )+
  labs(title = "Balanced PCA - Age",  x = "BPC_1", y = "BPC3"),



ggplot(pca_df[cells_sub, ], aes(x =BPC_1, y = BPC_3, fill = Nclone, shape = Juxta_Not)) +
  geom_point(size = 4, color="black", alpha = 0.5,
             position = position_jitter(width = 0, heigh=0)) +
  theme_classic() +
  scale_shape_manual(values = c(21,24,  23))+
  scale_fill_gradientn(  colours = c("white", "#E03932", "grey10") , 
                         na.value = "white"  )+
  labs(title = "Balanced PCA - Nclone - sub",       x = "BPC1", y = "BPC3")+
ggplot(pca_df, aes(x = BPC_1, y = BPC_3, fill = Nclone, shape = Juxta_Not)) +
  geom_point(size = 4, color="black", alpha = 0.5,
             position = position_jitter(width = 0, heigh=0)) +
  theme_classic() +
  scale_shape_manual(values = c(21,24,  23))+
  scale_fill_gradientn(  colours = c("white", "#E03932", "grey10") , 
                         na.value = "white"  )+
  labs(title = "Balanced PCA - Nclone",       x = "BPC_1", y = "BPC3"),


ggplot(pca_df[cells_sub, ], aes(x = BPC_1, y = BPC_3, fill = Ncell, shape = Juxta_Not)) +
  geom_point(size = 4, color="black", alpha = 0.5,
             position = position_jitter(width = 0, heigh=0)) +
  theme_classic() +
  scale_shape_manual(values = c(21,24,  23))+
  scale_fill_gradientn(  colours = c("white", "#eba00a", "grey10") , 
                         na.value = "white" , limits = c(0, 100) )+
  labs(title = "Balanced PCA - Ncell - sub",
       x = "BPC1", y = "BPC3")+
ggplot(pca_df, aes(x = BPC_1, y = BPC_3, fill = Ncell, shape = Juxta_Not)) +
  geom_point(size = 4, color="black", alpha = 0.5,
             position = position_jitter(width = 0, heigh=0)) +
  theme_classic() +
  scale_shape_manual(values = c(21,24,  23))+
  scale_fill_gradientn(  colours = c("white", "#eba00a", "grey10") , 
                         na.value = "white"  , limits = c(0, 100) )+
  labs(title = "Balanced PCA - Ncell",
       x = "BPC_1", y = "BPC3"))

l_plots
```


##Fig2a. Visualization of pseudotime

```{r}
l_plots <- list( 
  
   # Age vs pseudotime relationship
ggplot(pca_df, aes( x = Age, y =pseudotime)) +
  geom_point(size = 4, alpha = 0.5,color = "black",
             aes(shape = Juxta_Not, fill = Mouse),
             position = position_jitter(width = 0.1, height = 0)) +
  geom_smooth( method = "loess", 
               color = "black", se = FALSE) +
  theme_classic() + 
  scale_shape_manual(values = c(21, 24))+
  scale_fill_manual(values = col_mouse) +
  geom_hline(yintercept =  c(2, 5.5)) +
  
  
  # PCA colored by pseudotime
ggplot(pca_df, aes(x = BPC_1, y = BPC_3, fill = pseudotime)) +
  geom_point(size = 4, alpha = 0.6, color = "black",
             aes(shape = Juxta_Not)) +
  theme_classic() +
  scale_shape_manual(values = c(21, 24)) +
    scale_fill_gradientn(
    colours = c("white", "#9f97c8", "#514495", "#231e47", "grey10"),
    values = rescale(seq(min(pca_df$pseudotime, na.rm = TRUE),
                         max(pca_df$pseudotime, na.rm = TRUE),
                         length.out = 5)) )+
  
      # Same plot restricted to subsampled cells
ggplot(pca_df[cells_sub, ], aes(x = BPC_1, y = BPC_3, fill = pseudotime)) +
  geom_point(size = 4, alpha = 0.6, color = "black",
             aes(shape = Juxta_Not)) +
  theme_classic() +
  scale_shape_manual(values = c(21, 24)) +
    scale_fill_gradientn(
    colours = c("white", "#9f97c8", "#514495", "#231e47", "grey10"),
    values = rescale(seq(min(pca_df$pseudotime, na.rm = TRUE),
                         max(pca_df$pseudotime, na.rm = TRUE),
                         length.out = 5)) )


)
```


```{r}
wrap_plots(l_plots, ncol = 1)
```

## Fig2b. Clonal composition: Donut by mouse

```{r}
# Extract PCA coordinates (balanced PCA: PC1 vs PC3)
pca <- Embeddings(rna_seurat_umi_el, "balanced_pca")[,c(1, 3)] |>
  as.data.frame() |>
  tibble::rownames_to_column("SampleName")

# Retrieve Seurat metadata
mm_meta <- rna_seurat_umi_el@meta.data
mm_meta$SampleName = rownames(mm_meta)

# Merge PCA coordinates with metadata
meta_plot <- mm_meta |>
  left_join(pca, by = "SampleName")

```
Global functions for visualization
```{r}
col_bcm <- sample(colorRampPalette(brewer.pal(8, "Spectral"))(550))

# -------------------------------
#1. PCA visualization per mouse
# -------------------------------
plot_mouse_points <- function(df, mouse_id){

  df_mouse <- df |> dplyr::filter(Mouse == mouse_id)
  Age <- unique(df_mouse$Age)

  ggplot() +

    # Background: all cells (grey reference)
    geom_point(  data = df,
      aes(BPC_1, BPC_3),
      color = "grey90",  size = 2  ) +

   # Highlight selected mouse
    geom_point( data = df_mouse,
      aes(  BPC_1, BPC_3,  fill = Gland), 
      alpha = 0.9, shape = 21 )  +  
    theme_classic() + 
    ggtitle(paste("Mouse", mouse_id, Age, "months"))
}

# -------------------------------
#2. Construction of clonal arc coordinates (donut geometry)
# -------------------------------
make_donut_df <- function(df_mouse){

  df_mouse %>%
    mutate(sample_id = row_number()) %>%
    unnest(c(BCM_LVmerged_list, fraction_list)) %>%
    
     # Compute cumulative angular positions for each clone fraction
    group_by(sample_id) %>%
    mutate(
      end = 2*pi*cumsum(fraction_list),
      start = lag(end, default = 0)  ) %>%
    ungroup()
}

# -------------------------------
#3. PCA embedding with clonal donut overlays
# -------------------------------
plot_mouse_donuts <- function(df, mouse_id){

  df_mouse <- df %>% filter(Mouse == mouse_id)

  donut_df <- make_donut_df(df_mouse)
  Age <- unique(df_mouse$Age)
  ggplot() +
    # Background PCA
    geom_point(  data = df,
      aes(BPC_1, BPC_3),
      color = "grey90",    size = 2) +
    # Clonal composition displayed as donut sectors
    geom_arc_bar(
      data = donut_df,
      aes(x0 = BPC_1,  y0 = BPC_3,  r0 = 0,  r = 2,   
        start = start,    end = end,  fill = BCM_LVmerged_list  ),  color = NA  ) +
    coord_equal() +
    scale_fill_manual(values =col_bcm)+
    theme_classic() +theme(legend.position = "none")+
    labs(title = paste("Mouse", mouse_id, Age, "months"))
}



# -------------------------------
#4. Grid representation (mouse × sample structure)
# -------------------------------
plot_pie_grid <- function(df, mouse_id){

  df_mouse <- df %>% dplyr::filter(Mouse == mouse_id)

  donut_df <- make_donut_df(df_mouse)

   # Convert categorical variables into grid coordinates
  donut_df <- donut_df %>%
    mutate(
      x = as.numeric(factor(Gland)),
      y = as.numeric(factor(CellID ))
    )

  ggplot(donut_df) +
    geom_arc_bar( aes(x0 = x,   y0 = y,  r0 = 0,r = 0.45,   start = start,  end = end,
        fill = BCM_LVmerged_list  ),color = NA) +
    scale_x_continuous(
      breaks = unique(donut_df$x),
      labels = unique(donut_df$Gland)    ) +
    scale_y_continuous(
      breaks = unique(donut_df$y),
      labels = unique(donut_df$CellID )    ) +
    coord_equal() +
    theme_classic() +
    labs(x = "Gland", y = "MiniBulk") +
    scale_fill_manual(values = col_bcm)+
    theme(legend.position = "none")
}

# -------------------------------
#5. Combined multi-panel visualization
# -------------------------------
plot_mouse_combined <- function(df, mouse_id){

  p_mid <- plot_mouse_points(meta_plot, mouse)
  p_left  <- plot_mouse_donuts(df, mouse_id)
  
  p_right <- plot_pie_grid(df, mouse_id)

  p_left | p_mid + p_right 
}
```


Iteration over all mice
```{r}
for (mouse in unique(meta_plot$Mouse)) {

  print(plot_mouse_combined(meta_plot, mouse))

}
```

```{r}

clones_per_sample <- meta_plot %>%
  unnest(BCM_LVmerged_list) %>%
  group_by(BioSample) %>%
  summarise(n_clones = n_distinct(BCM_LVmerged_list))

clones_per_sample %>% left_join(rna_seurat_umi_el@meta.data[, c("BioSample", "Age")], by = "BioSample") %>% distinct
```

##Fig2c. Barcode and transcriptomic diversities through pseudotime
Figure 2c

```{r}
library(rstatix)
library(ggpubr)

data <- data.frame(pseudotime = rna_seurat_umi_el$pseudotime,
                   Cluster = rna_seurat_umi_el$Group_PT,
                   BarcodeDiversity = rna_seurat_umi_el$shannon_barcode_index,
                   TranscriptoimcDiversity = rna_seurat_umi_el$Distance_to_Biosample_Centroid,
                   Juxta_Not = rna_seurat_umi_el$Juxta_Not,
                   Ncells = rna_seurat_umi_el$CellNumber,
                   BioSample = rna_seurat_umi_el$BioSample,
                   SampleName = rna_seurat_umi_el$SampleName,
                   mouse = rna_seurat_umi_el$Mouse %>% as.factor)

df <- data %>%
  subset(Ncells == 100 & is.na(BioSample)==F)

# Tukey test
tuk <- df %>%
  tukey_hsd(BarcodeDiversity ~ Cluster) %>%
  add_xy_position(x = "Cluster")


g1 <- ggplot(df, aes(x = Cluster, y = BarcodeDiversity)) +
  geom_boxplot(aes(fill = Cluster), outliers = FALSE, color = "black") +
  geom_point(aes(shape = Juxta_Not, fill = Cluster),
             size = 1, alpha = 0.5, color = "black",
             position = position_jitter(width = 0.1, height = 0)) +
  stat_compare_means(method = "anova",
                     label.y = max(df$BarcodeDiversity, na.rm = TRUE) * 1.5) +
  stat_pvalue_manual(tuk, label = "p.adj.signif", step.increase = 0.1,  xmin = "group1",
  xmax = "group2") +
  scale_shape_manual(values = c(21, 24)) +
  scale_fill_manual(values = (colorRampPalette(brewer.pal(4, "Spectral"))(4))[c(1,2,3)]) +
  theme_classic() +
  theme(legend.position = "none")


tuk2 <- df %>%
  tukey_hsd(TranscriptoimcDiversity ~ Cluster) %>%
  add_xy_position(x = "Cluster")

g2 <- ggplot(df, aes(x = Cluster, y = TranscriptoimcDiversity)) +
  geom_boxplot(aes(fill = Cluster), outliers = FALSE, color = "black") +
  geom_point(aes(shape = Juxta_Not, fill = Cluster),
             size = 1, alpha = 0.5, color = "black",
             position = position_jitter(width = 0.1, height = 0)) +
  stat_compare_means(method = "anova",
                     label.y = max(df$TranscriptoimcDiversity, na.rm = TRUE) * 1.5) +
  stat_pvalue_manual(tuk2, label = "p.adj.signif", step.increase = 0.1,  xmin = "group1",
  xmax = "group2") +
   scale_shape_manual(values = c(21, 24))+
  scale_fill_manual(values = (colorRampPalette(brewer.pal(4, "Spectral"))(4))[c(1,2,3)])+
    theme_classic() +  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "none"  )

g1 + g2
```


```{r}
library(ggpubr)

g1 <- data %>%
  subset(Ncells ==100 & is.na(BioSample)==F) %>%
  subset(SampleName %in% cells_sub) %>%
  ggplot( aes(x = pseudotime, y = BarcodeDiversity)) +
  geom_point(aes(shape = Juxta_Not, fill = mouse), 
             size = 2,  alpha = 0.5, color = "black",
             position = position_jitter(width = 0.1, heigh=0)) +
   geom_smooth(method = "lm", color = "black", se = FALSE) +
  stat_cor(method = "spearman",
           label.x = Inf, label.y = Inf, hjust = 1.5, vjust = 1.5,   size = 4) +
  scale_shape_manual(values = c(21, 24)) +
  scale_fill_manual(values = col_mouse) +
  theme_classic() +
  theme(strip.text = element_text(face = "bold"),
        legend.position = "none")


g2 <- data %>%
  subset(Ncells ==100 & is.na(BioSample)==F) %>%
  subset(SampleName %in% cells_sub) %>% 
  ggplot(aes(x = pseudotime, y = TranscriptoimcDiversity)) +
  geom_point(aes(shape = Juxta_Not, fill = mouse), 
             size = 2,  alpha = 0.5, color = "black",
             position = position_jitter(width = 0.1, heigh=0)) +
   geom_smooth(method = "lm", color = "black", se = FALSE) +
  stat_cor(method = "spearman",
           label.x = Inf, label.y = Inf, hjust = 1.6, vjust = 1.5, size = 4) +
  scale_shape_manual(values = c(21, 24)) +
  scale_fill_manual(values = col_mouse) +
  theme_classic() +
  theme(strip.text = element_text(face = "bold"),
        legend.position = "none")

g1 + g2
```

##Fig2d. Pathway score heatmaps

Figure 2d

```{r}
#Extract score
meta <- rna_seurat_umi_el@meta.data
colnames(meta) <- gsub("1$", "", colnames(meta))
colnames(meta)[colnames(meta) %in% c("Basal_Score"  , "LP_Score"  ,"Luminal H-S_Score"  )] <- c("Basal", "LP", "Luminal H-S")

score_mat <- meta |>
  dplyr::select(dplyr::any_of(sig_pathways))

score_mat <- score_mat[rownames(meta), ]

#Annot tumor bearing
meta$Tm_Bearing <- meta$Juxta_Not
```

```{r}
library(ComplexHeatmap)
library(circlize)

# palette Age
age_col_fun <- colorRamp2(
  seq(min(meta$Age), max(meta$Age), length.out = 4),
  c("white", "#C0D69F", "#61980E", "grey10"))

pseudoT_col_fun <- colorRamp2(
  seq(min(meta$pseudotime, na.rm = T)+1, max(meta$pseudotime, na.rm = T)+1, length.out = 5),
  c("white", "#9f97c8", "#514495", "#231e47", "grey10"))

Clusters_col <- setNames(  colorRampPalette(brewer.pal(4, "Spectral"))(4)[1:3],
  levels(factor(meta$Group_PT)))
  

# palette Shannon 
shannon_col_fun <- colorRamp2(
  seq(min(meta$shannon_barcode_index, na.rm = T), max(meta$shannon_barcode_index, na.rm = T), length.out = 4),
  c("white", "grey70", "grey30", "grey10"))

# palette distance
Distance_col_fun <- colorRamp2(
  seq(min(meta$Distance_to_Biosample_Centroid, na.rm = T), max(meta$Distance_to_Biosample_Centroid, na.rm = T), length.out = 4),
  c("white", "grey70", "grey30", "grey10"))



#  Tm_Bearing (facteur)
tm_levels <- unique(meta$Tm_Bearing)

tm_cols <- setNames(  c( "grey10", "grey50", "grey95")[seq_along(tm_levels)],   tm_levels)

# --- Mouse (facteur)
mouse_levels <- unique(meta$Mouse)

mouse_cols <- setNames(  colorRampPalette(brewer.pal(8, "Spectral"))(length(mouse_levels)),
  mouse_levels)

# --- Annotation heatmap
age_anno <- HeatmapAnnotation(
  pseudoT = meta$pseudotime,
  Shannon = meta$shannon_barcode_index,
  Cluster = meta$Group_PT,
  Age = meta$Age,
  Tm_Bearing = meta$Tm_Bearing,
 
  Mouse = as.character(meta$Mouse),
  
 
  col = list(
    Shannon = shannon_col_fun,
    Cluster = Clusters_col,
    Tm_Bearing = tm_cols,
    Mouse = mouse_cols,
    pseudoT = pseudoT_col_fun,
    Age = age_col_fun),
  
  annotation_name_side = "left")



############# Heatmap
heatmap_palette <- c( rev( c("grey99","#E9F2FC", "#C1DAF6",  "#99C1F0", "#71A9EA", "#4991E4", "#3283E0", "#1B63B6", "#154E8E", "#0F3866") )
  ,"grey99", "#FCE9F2", "#F6C1DA", "#F099C1", "#EA71A9",  "#E44991", "#E03283", "#B61B63", "#8E154E",  "#660F38")

heatmap_col_fun <- colorRamp2(
  seq(-2, 2, length.out = length(heatmap_palette)),
  heatmap_palette)


```


```{r}
age_anno_sub <- HeatmapAnnotation(
  pseudoT = meta$pseudotime[rownames(meta) %in% cells_sub],
  Shannon = meta$shannon_barcode_index[rownames(meta) %in% cells_sub],
  Distance = meta$Distance_to_Biosample_Centroid[rownames(meta) %in% cells_sub],

  Age = meta$Age[rownames(meta) %in% cells_sub],
  Tm_Bearing = meta$Tm_Bearing[rownames(meta) %in% cells_sub],
 
  Mouse = as.character(meta$Mouse)[rownames(meta) %in% cells_sub],
  
 
  col = list(
    Shannon = shannon_col_fun,
    Distance = Distance_col_fun, 
    pseudoT =pseudoT_col_fun,
    Tm_Bearing = tm_cols,
    Mouse = mouse_cols,
    Age = age_col_fun),
  
  annotation_name_side = "left")

Heatmap(
  t(scale(score_mat[rownames(meta) %in% cells_sub,  ], center = TRUE, scale = TRUE)),
  name = "Pathway score",
  top_annotation = age_anno_sub,
  col = heatmap_col_fun,
  column_order = order(meta$pseudotime[rownames(meta) %in% cells_sub]), # Order according to pseudot
  cluster_columns = F,
  cluster_rows = TRUE,
  clustering_distance_rows = "euclidean",
  clustering_method_rows = "ward.D2",
  show_column_names = FALSE,
  row_names_gp = gpar(fontsize = 8) )

```


##Fig2e. Boxplots DEGs of interest

For Figure 2e & extended Fig. 4 i

```{r}
genes_of_interest <- c("Krt14", "Acta2", "Trp63","Zeb1","Snai1", "Twist1",
                       "Tspan8",  "Kit","Csn2","Wap", "Elf5","Bcl11b",  "Procr","Lgr5",
                       "Sox14", "Foxi1","Hey1", 
                       "Col1a2", "Vim","Serpine1","Mmp3",
                       
                       "Apobec3", "Rnase1", "Cox7b", "Cox7c",
                       "Deptor", "Gas5", "Igfbp5", "Irs1",
                       "Irf7", "Oasl1", "Parp14", "Isg15", "Fancd2",
                       "Tmem173", "Cgas", "Tbk1",
                       "Tlr3", "Hes1", "Bcl11a",
                       
                       "Il10", "Ccl9","Il23a", "Cd73", "Il6", "Irf3","Tnfrsf12a",
                       "Ccl7", "Il18r1","Csf1", "Ccl2","Cxcl2","Slpi","Cd74",
                       "Tgtp1", "Tgtp2", "Itgp",
                       
                       "Pdcd1", "Cxcl10", "Ctla4","Cxcl12","Cd73",
                       "H2-Aa","H2-Ab1", "Cd74","Hba-a1",
                       "Irs1", "B2m",
                       "Kdm6b", "Setdb1", "Crabp1",
                       "Fabp4",  "Chac1", "Cd36")

genes_of_interest <- intersect(
  genes_of_interest,
  rownames(rna_seurat_umi_el)
)

expr_mat <- GetAssayData(rna_seurat_umi_el, slot = "data")

df_plot <- as.data.frame(t(expr_mat[genes_of_interest, ]))
df_plot$Mouse <- rna_seurat_umi_el@meta.data$Mouse %>% as.character()
df_plot$Nclone <- rna_seurat_umi_el@meta.data$n_barcode

df_plot$Cluster <- rna_seurat_umi_el@meta.data$Group_PT
df_plot$BioSample <- rna_seurat_umi_el@meta.data$BioSample
df_plot$plate <- rna_seurat_umi_el@meta.data$Plate_Source
df_plot$pseudotime <- rna_seurat_umi_el@meta.data$pseudotime


df_plot <- df_plot %>%
  mutate(Juxta_Not = case_when(plate %in% c("LatePlq1", "LatePlq2") & 
                                 BioSample != "9341_G4MixedGFPpos" ~ "Jx",
                               TRUE ~ "Gl"))




df_long <- df_plot |>
  pivot_longer(
    cols = all_of(genes_of_interest),
    names_to = "Gene",
    values_to = "Expression"
  )

df_long$Gene <- factor(df_long$Gene, levels = genes_of_interest)



ggplot(df_long, aes(x = Cluster, y = Expression, fill = Cluster)) +
  geom_boxplot(notch = F, outliers = F, color = "black")+
  geom_point(aes(shape = Juxta_Not), 
             size = 1,  alpha = 0.5, color = "black",
             position = position_jitter(width = 0.1, heigh=0)) +
   scale_shape_manual(values = c(21, 24))+
  scale_color_manual(values =  c("white", "#C0D69F", "#61980E", "grey10"))+
  scale_fill_manual(values =  c("white", "#C0D69F", "#61980E", "grey10"))+
  scale_x_discrete(limits = c("E1","E2", "L")) +
  
  facet_wrap(~ Gene, scales = "free_y", ncol = 4) +
  theme_classic() +  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "none"  )

```



## Ext4g. PCA - Pathways assignments
```{r}
df_MiniBulk$BPC1 <-Embeddings(rna_seurat_umi_el , "balanced_pca")[, 1]
df_MiniBulk$BPC3 <- Embeddings(rna_seurat_umi_el , "balanced_pca")[, 3]
df_MiniBulk$pseudotime <- rna_seurat_umi_el$pseudotime


#Pathways annotation grouping
df_MiniBulk <- df_MiniBulk %>% 
  mutate(state_max_relev = case_when(state_max %in% c( "HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION", "HALLMARK_MYOGENESIS"  ) ~ "EMT",
                                     state_max %in% c("HALLMARK_ANGIOGENESIS", "HALLMARK_HYPOXIA"  ) ~ "Angiogenesis", 
                                     state_max %in% c("HALLMARK_COAGULATION"  ) ~ "Coagulation",
                                     state_max %in% c( "HALLMARK_APOPTOSIS" ,  "HALLMARK_P53_PATHWAY"  ,"HALLMARK_UNFOLDED_PROTEIN_RESPONSE" ,"HALLMARK_XENOBIOTIC_METABOLISM"  ) ~ "Stress",
                                     state_max %in% c( "HALLMARK_INTERFERON_GAMMA_RESPONSE" ,  "HALLMARK_INTERFERON_ALPHA_RESPONSE" ) ~ "Interferons",
                                      state_max %in% c( "HALLMARK_E2F_TARGETS" , "HALLMARK_G2M_CHECKPOINT"   ,   "HALLMARK_MITOTIC_SPINDLE") ~ "CellCycle",
                                     state_max == "HALLMARK_KRAS_SIGNALING_UP" ~"Kras",
                                     state_max == "HALLMARK_ADIPOGENESIS" ~"Adipogenesis",
                                     state_max == "HALLMARK_MTORC1_SIGNALING" ~"mTorC1",
                                     state_max == "HALLMARK_TNFA_SIGNALING_VIA_NFKB" ~"TNFa_NFkB",
                                      state_max == "HALLMARK_MYC_TARGETS_V1" ~"Myc",
                                     TRUE ~ state_max))

state_colors <- c(
  "Adipogenesis" = "#FFBF50FF", 
  "Angiogenesis" = "#1F83B4FF",  
  "CellCycle"    = "#8A60B0FF",  
  "Coagulation"  = "#12A2A8FF", 
  "EMT"          = "#2CA030FF",  
  "Interferons"  = "#D63A3AFF",  
  "Kras"         = "#C7519CFF",  
  "mTorC1"       = "#BA43B4FF",  #
  "Myc"          = "#6F63BBFF",  
  "Stress"       = "#BCBD22FF",  
  "TNFa_NFkB"    = "#FF7F0EFF",  
  "Unknown"      = "#D3D3D3"   # gris clair
)


ggplot(df_MiniBulk, aes(x = BPC1, y = BPC3)) +
  geom_point(aes(fill = state_max_relev),
             shape = 21,   size = 3,     color = "black",      stroke = 0.1) +
  theme_classic() +
  scale_fill_manual(values = state_colors)+
  theme(
    strip.text = element_text(size = 8),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )
```


##Ext4i. Barplot - Cumulative pathway activity by sample
```{r}
df_state_attrib <- df_MiniBulk |>
  dplyr::select(dplyr::any_of(sig_pathways)) 

head(df_state_attrib)

df_state_attrib$CumState <- rowSums(df_state_attrib)
df_state_attrib$Group <- rna_seurat_umi_el@meta.data$Group_PT
df_state_attrib$pseudotime <- rna_seurat_umi_el@meta.data$pseudotime




df_state_attrib %>%
  ggplot( aes(x = Group, y = CumState))+
  geom_point(alpha = 0.5, shape = 21,position = position_jitter(width = 0.1, height = 0))+
  geom_boxplot()+
  theme_classic()

ggplot(df_state_attrib, aes(x = pseudotime, y = CumState))+
  geom_point(aes(fill = Group), 
             size = 2,  alpha = 0.5, color = "black",
             position = position_jitter(width = 0.1, heigh=0)) +
  geom_smooth(method = "loess", color = "black")+
  scale_shape_manual(values = c(21, 24))+
  scale_color_manual(values = (colorRampPalette(brewer.pal(4, "Spectral"))(4))[c(2, 4, 1, 3)])+
  scale_fill_manual(values = (colorRampPalette(brewer.pal(4, "Spectral"))(4))[c(2, 4, 1, 3)])+
  theme_classic()

```

```{r}
anova_model <- aov(CumState~Group, data = df_state_attrib)
summary(anova_model)

tukey_res <- TukeyHSD(anova_model)
tukey_res
```



## Ext4j. PCA - Pathways assignments
```{r}
#Barplot cumulatif cell state per Group
df_MiniBulk$Group_PT <- rna_seurat_umi_el$Group_PT

# Calcul des proportions
prop_df <- df_MiniBulk %>%
  filter(state_max_relev != "Unknown") %>%
  group_by(Group_PT, state_max_relev) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Group_PT) %>%
  mutate(prop = n / sum(n))


# Barplot

ggplot(prop_df, aes(x = Group_PT, y = prop, fill = state_max_relev)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
   scale_fill_manual(values = state_colors)+
  labs(
    title = "Proportions de state_max par Group",
    x = "Group",
    y = "Proportion",
    fill = "state_max"
  ) +
  theme_minimal()


```





##Ext4k. Heatmap genes


```{r}
expr_mat <- GetAssayData(rna_seurat_umi_el, slot = "data")

#color scale
col_fun <- colorRamp2(
  c(-2, -1, 0, 1, 2),
  c("#3B6FB6", "#A6C0E8", "#F2F2F2", "#E6A0A0", "#B22222")
)

```
```{r}
genes_of_interest <- c( "H2-Aa","H2-Ab1", "Cd74",    "Foxi1","Hey1", "Deptor","Irf7", "Oasl1", "Parp14","Tbk1",  "Isg15",  "Gas5", "Igfbp5",  "B2m","Irs1","Cxcl10","Cd74", "Irf3" ,#Immune DW
                        "Bcl11a", "Apobec3", #DW 
                       "Tspan8", #putative MaSC merker
                       "Kit","Elf5",  #LP markers
                      
                       
                      "Setdb1","Rnase1", "Cox7b", "Cox7c", "Hes1",  "Tlr3",   "Tmem173","Sox14",  "Hba-a1", #Latency UUP
                   
                       
                       
                      
                         "Kdm6b"  , #latency dw
                       "Csn2", "Wap", #Alv 
                         "Krt14", "Acta2", #basal marker
                         "Col1a2", "Vim","Serpine1","Tnfrsf12a","Procr", "Slpi", "Fabp3",   "Chac1",#UP
                        "Ccl7", "Csf1", "Ccl2","Il6","Ccl9","Il23a","Cxcl12"    #Immune Up
                       
                      
 )

genes_of_interest <- intersect(
  genes_of_interest,
  rownames(rna_seurat_umi_el)
)

mat <- expr_mat[genes_of_interest, rownames(meta) %in% cells_sub] %>% as.matrix %>%t


Heatmap(
  t(scale(mat, center = TRUE, scale = TRUE)),
        cluster_columns = FALSE,
        cluster_rows = F, 
        column_order = order(meta$pseudotime[rownames(meta) %in% cells_sub]), # Order according to pseudot
        show_column_names = FALSE,
   top_annotation = age_anno_sub,
  col = col_fun,
  row_names_gp = gpar(fontsize = 6) )
```




## Fig2f - pies Pathways
```{r}
library(scatterpie)
# Binary hallmarks
hallmark_cols <- grep("^HALLMARK_", colnames(df_MiniBulk)[55:ncol(df_MiniBulk)], value = TRUE)

# Long format
df_long <- df_MiniBulk %>%
  dplyr::select(BioSample, BPC1, BPC3, all_of(hallmark_cols)) %>%
  pivot_longer(
    cols = all_of(hallmark_cols),
    names_to = "hallmark",
    values_to = "value"
  ) %>%
  dplyr::filter(value == TRUE) %>%
  mutate(
    state_group = case_when(
      hallmark %in% c("HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION", "HALLMARK_MYOGENESIS") ~ "EMT",
      hallmark %in% c("HALLMARK_ANGIOGENESIS", "HALLMARK_HYPOXIA") ~ "Angiogenesis",
      hallmark %in% c("HALLMARK_COAGULATION") ~ "Coagulation",
      hallmark %in% c("HALLMARK_APOPTOSIS", "HALLMARK_P53_PATHWAY", "HALLMARK_UNFOLDED_PROTEIN_RESPONSE", "HALLMARK_XENOBIOTIC_METABOLISM") ~ "Stress",
      hallmark %in% c("HALLMARK_INTERFERON_GAMMA_RESPONSE", "HALLMARK_INTERFERON_ALPHA_RESPONSE") ~ "Interferons",
      hallmark %in% c("HALLMARK_E2F_TARGETS", "HALLMARK_G2M_CHECKPOINT", "HALLMARK_MITOTIC_SPINDLE") ~ "CellCycle",
      hallmark == "HALLMARK_KRAS_SIGNALING_UP" ~ "Kras",
      hallmark == "HALLMARK_ADIPOGENESIS" ~ "Adipogenesis",
      hallmark == "HALLMARK_MTORC1_SIGNALING" ~ "mTorC1",
      hallmark == "HALLMARK_TNFA_SIGNALING_VIA_NFKB" ~ "TNFa_NFkB",
      hallmark == "HALLMARK_MYC_TARGETS_V1" ~ "Myc",
      TRUE ~ "Other"
    )
  )

#hallmark per sample - size pie
df_pie <- df_long %>% dplyr::select(BioSample, BPC1 , BPC3, state_group )  %>% distinct() %>%
  group_by( BPC1, BPC3, state_group ) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(BPC1, BPC3) %>%
  mutate(freq = n / sum(n))


df_wide <- df_pie %>%
  pivot_wider(
    names_from = state_group,
    values_from = freq,
    values_fill = 0
  ) %>%
  dplyr::filter(
    is.finite(BPC1),
    is.finite(BPC3)
  )
cols_use <- setdiff(colnames(df_wide), c("BioSample", "BPC1", "BPC3", "n"))


plot_tot <- ggplot() +
  geom_scatterpie(
    data = df_wide,
    aes(x = BPC1, y = BPC3),
   cols = cols_use,
    pie_scale = 0.5
  ) +
  scale_fill_manual(values = state_colors)+
  coord_equal() +
  theme_classic()

plot_tot
```
```{r}
#PCA generale
df_bg <- df_MiniBulk %>%
  select(BioSample, BPC1, BPC3)

#add biosample
df_wide <- df_wide %>%
  left_join(
    df_MiniBulk %>% select(BioSample, BPC1, BPC3),
    by = c("BPC1", "BPC3")
  )


cols_use <- setdiff(colnames(df_wide), c("BioSample", "BPC1", "BPC3", "n"))

samples <- unique(df_wide$BioSample)
df_wide$r <- 1

```


```{r}
plots <- lapply(samples, function(s) {
  
  df_sub <- df_wide %>% filter(BioSample == s)
  
  # -------------------------
  # LEFT PANEL : PCA
  # -------------------------
  p_left <- ggplot() +
    geom_point(
      data = df_bg,
      aes(x = BPC1, y = BPC3),
      color = "grey85",
      size = 1
    ) +
    geom_point(
      data = df_sub,
      aes(x = BPC1, y = BPC3),
      color = "black",
      size = 2
    ) +
    coord_equal() +
    theme_classic() +
    labs(title = paste0("BioSample: ", s))
  
  
  # -------------------------
  # RIGHT PANEL : GRID PIE
  # -------------------------
df_sub <- df_wide %>% ungroup() %>% 
  filter(BioSample == s) %>%
  group_by(BPC1, BPC3) %>%
  summarise(    across(  all_of(cols_use),
      ~ sum(.x, na.rm = TRUE)
    ),
    .groups = "drop"
  ) %>% 
    rowwise() %>%
  mutate(
    total = sum(c_across(all_of(cols_use)))
  ) %>%
  mutate(
    across(all_of(cols_use), ~ ifelse(total > 0, .x / total, 0))
  ) %>%
  ungroup() %>%
  mutate(id = row_number())

ncol_grid <- 2

df_sub <- df_sub %>%
  mutate(
    col = (id - 1) %% ncol_grid,
    row = floor((id - 1) / ncol_grid)
  )


df_sub <- df_sub %>%
  mutate(
    x_plot = col ,
    y_plot = -row,
    r_plot = 0.2   # rayon constant adapté
  )
  
  p_right <- ggplot() +
    geom_scatterpie(
      data = df_sub,
      aes(x = x_plot, y = y_plot, r = r_plot),
      cols = cols_use
    ) +
    scale_fill_manual(values = state_colors) +
    coord_equal() +
    theme_void() +
    labs(title = "Hallmark composition")
  
  
  # -------------------------
  # COMBINE
  # -------------------------
  p_left + p_right + plot_layout(widths = c(1.2, 1))
})

plots[[12]]
```

```{r}
for (s in 1:length(samples)) {

  print(plots[[s]])

}

```

## Fig2g. State dynamics along pseudotime (alluvial / GAM smoothing)
```{r}
# Binary columns
hallmark_cols <- grep("^HALLMARK_", colnames(df_MiniBulk)[55:ncol(df_MiniBulk)], value = TRUE)


df_long <- df_MiniBulk %>%
  select(BioSample, BPC1, BPC3, pseudotime, all_of(hallmark_cols)) %>%
  pivot_longer(
    cols = all_of(hallmark_cols),
    names_to = "hallmark",
    values_to = "value"
  ) %>%
  filter(value == TRUE) %>%
  mutate(
    state_group = case_when(
      hallmark %in% c("HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION", "HALLMARK_MYOGENESIS") ~ "EMT",
      hallmark %in% c("HALLMARK_ANGIOGENESIS", "HALLMARK_HYPOXIA") ~ "Angiogenesis",
      hallmark %in% c("HALLMARK_COAGULATION") ~ "Coagulation",
      hallmark %in% c("HALLMARK_APOPTOSIS", "HALLMARK_P53_PATHWAY", "HALLMARK_UNFOLDED_PROTEIN_RESPONSE", "HALLMARK_XENOBIOTIC_METABOLISM") ~ "Stress",
      hallmark %in% c("HALLMARK_INTERFERON_GAMMA_RESPONSE", "HALLMARK_INTERFERON_ALPHA_RESPONSE") ~ "Interferons",
      hallmark %in% c("HALLMARK_E2F_TARGETS", "HALLMARK_G2M_CHECKPOINT", "HALLMARK_MITOTIC_SPINDLE") ~ "CellCycle",
      hallmark == "HALLMARK_KRAS_SIGNALING_UP" ~ "Kras",
      hallmark == "HALLMARK_ADIPOGENESIS" ~ "Adipogenesis",
      hallmark == "HALLMARK_MTORC1_SIGNALING" ~ "mTorC1",
      hallmark == "HALLMARK_TNFA_SIGNALING_VIA_NFKB" ~ "TNFa_NFkB",
      hallmark == "HALLMARK_MYC_TARGETS_V1" ~ "Myc",
      TRUE ~ "Other"
    )
  )%>% select( pseudotime, state_group )  %>% distinct()

head(df_long)
```

```{r}
#pseudotime Bining
df_long <- df_long %>%
  mutate(pt_bin = cut(pseudotime, breaks = 20))  
df_prop <- df_long %>%
  group_by(pt_bin, state_group) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(pt_bin) %>%
  mutate(prop = n / sum(n))
```
```{r}
library("ggforce")

levels_x <- names(state_colors)
alpha <- 0.7

#pseudotime Bining
df_prop <- df_long %>%
  filter(is.na(pseudotime)==F) %>%

    group_by(pseudotime, state_group) %>%
  summarise(n = n(), .groups = "drop") %>% ungroup()%>%
  complete(pseudotime, state_group = levels_x, fill = list(n = 0)) %>% # Put 0 instead of null
  group_by(pseudotime) %>%
  mutate(freq = n / sum(n)) %>%
  ungroup()

g1 <- ggplot(df_prop,
       aes(x = pseudotime,
           y = freq,
           fill = state_group)) +
  geom_area(position = "fill", alpha = 0.9) +
  scale_fill_manual(values = state_colors) +
  theme_classic() +
  theme(legend.position = "right")

```


```{r}

#oyu GAM
library(mgcv)

df_smooth <- df_prop %>%
  group_by(state_group) %>%
  do({
    df <- .
    
    # Avoid 0 & 1 
    df$freq_adj <- pmin(pmax(df$freq, 1e-5), 1 - 1e-5)
    
    # fit GAM
    m <- gam(freq_adj ~ s(pseudotime, k = 15),
             data = df,
             family = quasibinomial())
    
    # predictions
    df$freq_smooth <- predict(m, type = "response")
    
    df
  }) %>%
  ungroup()

df_smooth <- df_smooth %>%
  group_by(pseudotime) %>%
  mutate(freq_smooth = freq_smooth / sum(freq_smooth)) %>%
  ungroup()

g2 <- ggplot() +
  geom_area(data = df_smooth,
       aes(x = pseudotime, y = freq_smooth, fill = state_group),
       alpha = 0.9) +
  scale_fill_manual(values = state_colors) +
  theme_classic()

grid.arrange(g1, g2, ncol =1)
```


##Fig2h. Shannon diversity of pathways
```{r}
# cBinary columns hall.
hallmark_cols <- grep("^HALLMARK_", colnames(df_MiniBulk)[55:ncol(df_MiniBulk)], value = TRUE)

#long
df_long <- df_MiniBulk %>%
  dplyr::select(BioSample, BPC1, BPC3, all_of(hallmark_cols)) %>%
  pivot_longer(
    cols = all_of(hallmark_cols),
    names_to = "hallmark",
    values_to = "value"
  ) %>%
  dplyr::filter(value == TRUE) %>%
  mutate(
    state_group = case_when(
      hallmark %in% c("HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION", "HALLMARK_MYOGENESIS") ~ "EMT",
      hallmark %in% c("HALLMARK_ANGIOGENESIS", "HALLMARK_HYPOXIA") ~ "Angiogenesis",
      hallmark %in% c("HALLMARK_COAGULATION") ~ "Coagulation",
      hallmark %in% c("HALLMARK_APOPTOSIS", "HALLMARK_P53_PATHWAY", "HALLMARK_UNFOLDED_PROTEIN_RESPONSE", "HALLMARK_XENOBIOTIC_METABOLISM") ~ "Stress",
      hallmark %in% c("HALLMARK_INTERFERON_GAMMA_RESPONSE", "HALLMARK_INTERFERON_ALPHA_RESPONSE") ~ "Interferons",
      hallmark %in% c("HALLMARK_E2F_TARGETS", "HALLMARK_G2M_CHECKPOINT", "HALLMARK_MITOTIC_SPINDLE") ~ "CellCycle",
      hallmark == "HALLMARK_KRAS_SIGNALING_UP" ~ "Kras",
      hallmark == "HALLMARK_ADIPOGENESIS" ~ "Adipogenesis",
      hallmark == "HALLMARK_MTORC1_SIGNALING" ~ "mTorC1",
      hallmark == "HALLMARK_TNFA_SIGNALING_VIA_NFKB" ~ "TNFa_NFkB",
      hallmark == "HALLMARK_MYC_TARGETS_V1" ~ "Myc",
      TRUE ~ "Other"
    )
  )

#hallmark per sample - size pie
df_pie <- df_long %>% dplyr::select(BioSample, BPC1 , BPC3, state_group )  %>% distinct() %>%
  group_by( BPC1, BPC3, state_group ) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(BPC1, BPC3) %>%
  mutate(freq = n / sum(n))


df_wide <- df_pie %>%
  pivot_wider(
    names_from = state_group,
    values_from = freq,
    values_fill = 0
  ) %>%
  dplyr::filter(
    is.finite(BPC1),
    is.finite(BPC3)
  )
```

```{r}
cols_use <- setdiff(colnames(df_wide), c("BioSample", "BPC1", "BPC3", "r", "n"))

df_shannon <- df_wide %>%
  group_by(BPC1, BPC3) %>%
  summarise(    across(  all_of(cols_use),
      ~ sum(.x, na.rm = TRUE)
    ),
    .groups = "drop"
  ) %>%
  ungroup() %>%
  rowwise() %>% 
  mutate(
    shannon = {
      p <- c_across(all_of(cols_use))
      p <- p[p > 0]
      -sum(p * log(p))
    }
  ) %>%
    left_join(
    df_MiniBulk %>% dplyr::select( BPC1, BPC3, Group_PT,BioSample, Juxta_Not, pseudotime),
    by = c( "BPC1", "BPC3")
  ) %>% distinct()


p1 <- ggplot(df_shannon, aes(x = Group_PT, y = shannon, fill = Group_PT)) +
  geom_boxplot(outliers = F, color = "black")+
  geom_point(aes(shape = Juxta_Not), 
             size = 1,  alpha = 0.5, color = "black",
             position = position_jitter(width = 0.1, heigh=0)) +
  stat_compare_means(method = "anova",
                     label.y = max(df_shannon$shannon, na.rm = TRUE) * 1.5) +
  stat_pvalue_manual(tuk, label = "p.adj.signif", step.increase = 0.1,  xmin = "group1",
  xmax = "group2") +
  scale_shape_manual(values = c(21, 24))+
  scale_color_manual(values = (colorRampPalette(brewer.pal(4, "Spectral"))(4))[c(2, 4, 1, 3)])+
  scale_fill_manual(values = (colorRampPalette(brewer.pal(4, "Spectral"))(4))[c(2, 4, 1, 3)])+
  theme_classic()

p1
```

```{r}
anova_model <- aov(shannon~Group_PT, data = df_shannon)
summary(anova_model)

tukey_res <- TukeyHSD(anova_model)
tukey_res
```

```{r}
p2 <- ggplot(df_shannon, aes(x = pseudotime, y = shannon))+
  geom_point(aes(shape = Juxta_Not, fill = Group_PT), 
             size = 2,  alpha = 0.5, color = "black",
             position = position_jitter(width = 0.1, heigh=0)) +
  geom_smooth(method = "lm", color = "black")+
  scale_shape_manual(values = c(21, 24))+
  scale_color_manual(values = (colorRampPalette(brewer.pal(4, "Spectral"))(4))[c(2, 4, 1, 3)])+
  scale_fill_manual(values = (colorRampPalette(brewer.pal(4, "Spectral"))(4))[c(2, 4, 1, 3)])+
  theme_classic()

p2
```

```{r}
cor(df_shannon$shannon, df_shannon$pseudotime, method = "spearman", use = "complete.obs")
```
```{r}
p1 + p2

```






