---
title: "R Notebook"
output: html_notebook
---


```{r}
library(data.table)
library(tidyverse)
library(dplyr)
library(readxl)

library(Seurat)
library(patchwork)
library(DoubletFinder)

library(ggplot2)
library(gridExtra)
library(ggExtra)
library(viridis)
library(Matrix)
library("ggsci")
library(cowplot)
library(platetools)
library(scales)
library("gridExtra")
library(ggpubr)
library(RColorBrewer)
library(tidyr)
library(dplyr)
library(ggforce)

#To add filename
read_plus <- function(flnm) {
    read.csv(flnm, header = TRUE, na.strings=c("", " ")) %>% 
        mutate(filename = flnm)
}

# Def Paths
Output_Dir <- "C:/Users/Marthe Laisne/OneDrive - INSTITUT CURIE/m_laisne/Publications/2026_Laisne/Scripts/Output/sc_Pseudotime/"
```


# Score single-cell
The pre-tumoral pseudotime score was derived at single-cell resolution by projecting the sc-RNAseq dataset onto the previously defined “MiniBulk” principal component space. 
Cells were mapped using shared gene loadings and centered expression values, enabling transfer of the PCA structure across datasets. 
The same linear model relating principal components to biological age, previously defined, was then fitted, and pseudotime was defined as a weighted combination of projected PC scores using the learned coefficients, providing an estimation of pre-tumoral progression at the single-cell level


```{r}
seurat_rna <- qs::qread(file.path("C:/Users/Marthe Laisne/OneDrive - INSTITUT CURIE/m_laisne/Publications/2026_Laisne/seurat_rna_step7_new.qs"))

seurat <- qs::qread(file.path("C:/Users/Marthe Laisne/OneDrive - INSTITUT CURIE/m_laisne/Publications/2026_Laisne/seurat_step5_new.qs"))

#Bulk
rna_seurat_umi_el <- qs::qread(file.path("C:/Users/Marthe Laisne/OneDrive - INSTITUT CURIE/m_laisne/Publications/2026_Laisne/Scripts/Output/MiniBulkRNA/rna_seurat_umi_el.qs"))

#PCA parameters
features <- qs::qread(file.path("C:/Users/Marthe Laisne/OneDrive - INSTITUT CURIE/m_laisne/Publications/2026_Laisne/Scripts/Output/MiniBulkRNA/features_PCA_MiniBulk.qs"))

loadding <-  qs::qread(file.path("C:/Users/Marthe Laisne/OneDrive - INSTITUT CURIE/m_laisne/Publications/2026_Laisne/Scripts/Output/MiniBulkRNA/loading_PCA_MiniBulk.qs"))

center <-  qs::qread(file.path("C:/Users/Marthe Laisne/OneDrive - INSTITUT CURIE/m_laisne/Publications/2026_Laisne/Scripts/Output/MiniBulkRNA/centers_PCA_MiniBulk.qs"))

```



```{r}
DefaultAssay(seurat_rna) <- ("RNA_umi")

common_genes <- intersect(features, rownames(seurat_rna ) )

seurat_rna <- NormalizeData(seurat_rna)
seurat_rna  <- ScaleData(
 seurat_rna ,
  features = common_genes,
  verbose = FALSE
)

common_load <- intersect(rownames(loadings) , rownames(seurat_rna))

######## proj

scaled <- GetAssayData( seurat_rna ,  slot = "data")[common_load, ] %>% t() 
scaled <-   scale(scaled, center = center[common_load], scale = F)



# projection = X * rotation
proj <- t(loadings[common_load,]) %*% t(scaled)

#######

seurat_rna[["pca_proj"]] <- CreateDimReducObject(
  embeddings = t(proj),
  key = "projPC_"
)

DimPlot(seurat_rna, reduction = "pca_proj", group.by = "CellFate_RNA", dims = c(1, 3))

DimPlot(seurat_rna, reduction = "pca_proj", group.by = "Group", dims = c(1, 3))
```


Pseudotemps
```{r}
coefs <- coefficients(fit)

seurat_rna@meta.data$pseudotime <- 
  coefs["(Intercept)"] +
  coefs["BPC_1"] * seurat_rna@reductions$pca_proj@cell.embeddings[,1] +
  coefs["BPC_3"] * seurat_rna@reductions$pca_proj@cell.embeddings[,3]

```



```{r}
qs::qsave(seurat_rna, file = paste0(Output_Dir, "seurat_rna_withPseudotime.qs") )

```


