---
title: "R Notebook"
output: html_notebook
---

# Intro

```{r}
cols_Group <- c("CreP_EARLY" = "#C0D69F", 
                "CreN" = "grey",
                "CreP_LATE"  = "#61980E",
                "Tm"    = "#841811")
```


```{r}
library(CellBarcode)
library("ggnewscale") ; library(ggpubr)


library(data.table)
library(tidyverse)
library(dplyr)
library(readxl)

library(Seurat)
library(patchwork)
library(DoubletFinder)

library(ggplot2)
library(gridExtra)
library(ggExtra)
library(viridis)
library(Matrix)
library("ggsci")
library(cowplot)
library(platetools)
library(scales)
library("gridExtra")
library(qs)
library(RColorBrewer)

library(stringdist)
library(Biostrings)
library(msa)
library(seqinr)
library(ggmsa)
```


Before the following workflow was done, to construct and annotated the DRAG-related object
- import sc metadata and DRAG FASTQ files;
- extract lineage barcodes from R1, R2 and hybrid (R2 + UMI) reads > *Bold sequences* indicated the motif grepped (no filter on UMI / depth)

5’-acctcctcgaggtca*tcgaagtatcaagtcca*gtag   V sequence
5'-tctactatcgttacgac                        D sequence
5'-gtagctact*accgtagcaagctcgagagtagacCTACTG*      J sequence + linker GFP
length: 86nt

- summarize all detected barcodes for each cell;
- merge barcode information with sc metadata;
- Add Pgen
- save the annotated metadata for downstream analyses.



```{r}
# Define the output directory and load the metadata table containing raw DRAG barcode information
outputDir <- "C:/Users/Marthe Laisne/OneDrive - INSTITUT CURIE/m_laisne/Publications/2026_Laisne/Scripts/Output/sc_DRAG/"

DRAG_All <- qs::qread(paste0(outputDir, "sc_Metadata_withAllDRAG.qs"))

Pgen_table <- read.delim("C:/Users/Marthe Laisne/OneDrive - INSTITUT CURIE/m_laisne/Publications/2026_Laisne/Tables/SupplementaryTable_Pgen.txt")

```

```{r}
#Add complementary fixed part to the barcodes sequences
DRAG_All$R1_seq_raw_long <- lapply(DRAG_All$R1_seq_raw,
                                   function(X){
                                     paste0(toupper("acctcctcgaggtcatcgaagtatcaagtcca"),
                                            X,
                                            toupper("accgtagcaagctcgagagtagac"))
                                   }
)

```

# QC

## Before barcode quality control
### Evaluate barcode generation probability (Pgen) prior to filtering

- Extract all raw R1 lineage barcodes from the single-cell metadata and annotate each barcode with its estimated probability of generation (Pgen).

- Quantify the number of distinct mice in which each barcode is detected to estimate barcode sharing across animals.

- Visualize the distribution of barcode generation probabilities according to the degree of barcode sharing between mice. Frequently shared barcodes are expected to exhibit higher Pgen values, consistent with independent recurrent generation rather than true clonal relationships.

- Assess the impact of increasingly stringent Pgen thresholds on barcode sharing by calculating, for each threshold (10^-x), the fraction of barcodes detected in more than one mouse. This analysis guides the selection of a Pgen cutoff that minimizes independently generated recurrent barcodes while retaining the largest possible number of informative lineage barcodes.

```{r}
# Extract all raw R1 lineage barcodes and associated mouse identifiers.
AllBCM1_R1 <- DRAG_All %>%
  select(MouseNumber, R1_seq_raw_long) %>%
  unnest(R1_seq_raw_long)  # Expand cells containing multiple barcodes into one row per barcode.


# Annotate each unique barcode with its estimated probability of generation (Pgen).
AllBCM1_R1 <- AllBCM1_R1 %>%
  left_join(
    Pgen_table %>% 
      dplyr::select(sequence, Pgen) %>%
      distinct(),
    by = c("R1_seq_raw_long" = "sequence")
  )

# Count the number of distinct mice in which each barcode is detected.
AllBCM1_R1 <- AllBCM1_R1 %>%
  ungroup() %>% group_by( R1_seq_raw_long) %>%
  mutate(Barcode_sharing_mice_N = n_distinct(MouseNumber))

# SClassify barcodes ccording to the number of mice in which they are detcted

AllBCM1_R1 %>%
  dplyr::filter(!is.nan(Pgen), !is.na(Pgen),is.finite(Pgen)) %>%   
  ungroup() %>% group_by(R1_seq_raw_long) %>%
  summarise(
    Pgen = dplyr::first(Pgen),
    Barcode_sharing_mice_N = max(Barcode_sharing_mice_N),
    .groups = "drop"
  ) %>%
  mutate(Barcode_sharing_mice_Factor = case_when(Barcode_sharing_mice_N == 1  ~ "unique",
                                                 Barcode_sharing_mice_N > 1 & Barcode_sharing_mice_N < 4 ~ "2-3",
                                                 Barcode_sharing_mice_N>=4 ~  "4-7")) %>%
  mutate(Barcode_sharing_mice_Factor = factor(Barcode_sharing_mice_Factor,
                                              levels = c("unique", 
                                                         "2-3",
                                                         "4-7"))) %>%
  
    # Visualize the Pgen distribution according to barcode sharing across mice. 
  ggplot(aes(x = -log10(Pgen), fill = Barcode_sharing_mice_Factor)) +
  geom_histogram(alpha = 0.5, aes(y = after_stat(count))) +
   geom_vline(xintercept = 3.8)+
  labs(
    x = "Pgen",
    fill = "Barcode sharing (# mice)",
    title = "Pgen distribution by barcode sharing") +
  theme_classic()
```

```{r}
# Evaluate the effect of progressively more stringent Pgen thresholds on barcode sharing between mice.

df <-AllBCM1_R1 %>%
  dplyr::filter(!is.nan(Pgen), is.finite(Pgen)) %>%   # évite NaN / Inf
   ungroup() %>%group_by(R1_seq_raw_long) %>%
  summarise(
    Pgen = dplyr::first(Pgen),
    Barcode_sharing_mice_N = max(Barcode_sharing_mice_N),
    .groups = "drop"  ) 

# Define a grid of Pgen thresholds (10^-x).
thr_grid <- seq(0, 11, by = 0.5)


# For each threshold, calculate the fraction of barcodes shared by more than one mouse.
res <- lapply(thr_grid, function(x){

  thr <- 10^-x

  sub <- df %>% dplyr::filter(Pgen <= thr)

  if(nrow(sub) == 0){
    return(data.frame(x = x, frac = NA))
  }

  frac_shared <- mean(sub$Barcode_sharing_mice_N > 1)

  data.frame(
    x = x,
    frac = frac_shared
  )

}) %>% bind_rows()


# Record the number of retained barcodes at each threshold.
res$n <- sapply(thr_grid, function(x){
  sum(df$Pgen <= 10^-x)
})

# Plot the fraction of shared barcodes as a function of the Pgen threshold.
ggplot(res, aes(x = x, y = frac)) +
  geom_point(size = 3) +
  geom_text(aes(label = n), vjust = -1, size = 3)+
  labs(
    x = expression("Pgen threshold 10"^-x),
    y = "Fraction barcode shared between mice"  ) +
  theme_classic()
```

### Barcode capture rate assessment

The proportion of cells with at least one detected DRAG barcode sequence (from R1 and/or R2 reads) is evaluated before applying any filtering step.

```{r}
# Compute the proportion of cells with at least one detected barcode sequence (R1 or R2) for each mouse and experimental group


DRAG_All %>%
  mutate(has_barcode = !is.na(R1_seq_raw) | !is.na(R2_seq_raw)) %>%
  group_by(Group, MouseNumber) %>%
  summarise(
    has_barcode = sum(has_barcode),
    n_cells = n(),
    prop_barcoded = has_barcode / n_cells,
       .groups = "drop"  ) %>%


# Visualize barcode capture rate across experimental groups.
# Bars represent the mean proportion of barcoded cells across mice, with error bars showing the standard deviation.

  ggplot( aes(x = Group, y = prop_barcoded, fill = Group)) +
    stat_summary(fun = mean, geom = "bar", width = 0.7, color = "black") +
    stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1),
               geom = "errorbar", width = 0.2) +
    geom_jitter(width = 0.12, size = 2, alpha = 0.8) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 1)) +
    labs( x = NULL,
        y = "Cells with R1 or R2 barcode (%)"  ) +
    theme_classic() +
    theme(    legend.position = "none",
      axis.text.x = element_text(angle = 45, hjust = 1)  )
```

```{r}
# Quantify cells lacking any barcode detection in R1 and R2 according to Cre status
table((is.na(DRAG_All$R1_seq_raw)& is.na(DRAG_All$R2_seq_raw)), DRAG_All$Cre)
```

```{r}
1291/(1291+555)
```


Barcode detection efficiency was further evaluated separately for R1 and R2 reads. The proportion of cells with a barcode detected in each read was compared to the overall capture rate (R1 OR R2).

```{r}
# Define barcode detection status independently for R1 and R2
df_prop <- DRAG_All %>%
  mutate(
    R1 = !is.na(R1_seq_raw),
    R2 = !is.na(R2_seq_raw),
    Either = R1 | R2
  ) %>%
  
    # Calculate proportions per mouse and experimental group
  group_by(Group, MouseNumber) %>%
  summarise(
    R1 = mean(R1),
    R2 = mean(R2),
    Either = mean(Either),
    .groups = "drop"
  ) %>%
  
    # Convert to long format for visualization
  tidyr::pivot_longer(R1:Either,
                      names_to = "Method",
                      values_to = "Proportion")

df_prop

# Plot barcode capture rate according to the sequencing read
ggplot(df_prop, aes(x = Group, y = Proportion)) +
  stat_summary(fun = mean, geom = "bar", width = 0.7, color = "black", aes(fill = Method)) +
  stat_summary(data = subset(df_prop, Method == "Either"),
    fun.data = mean_sdl, fun.args = list(mult = 1),
               geom = "errorbar", width = 0.2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 1)) +
  labs(
    x = NULL,
    y = "Cells with R1 or R2 barcode (%)"
  ) +
  theme_classic() +
  theme(
    #legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

Distribution of sequencing depth supporting the primary detected barcode.
This analysis is used to evaluate the detection threshold applied during QC.

```{r}
# Replace empty barcode count lists by NA values
DRAG_All$R1_count_raw[sapply(DRAG_All$R1_count_raw, is.null)] <- NA


# Extract the read count of the primary barcode detected in each cell
st <- data.frame(count = lapply(DRAG_All$R1_count_raw, function(X){X[1]}) %>% unlist,
                 class = DRAG_All$Group)
head(st)

# Visualize barcode read count distribution.
# The horizontal line indicates the selected minimum detection threshold
# (3 reads).
s <- ggplot(st, aes(x=class, y = log10(1+count), fill = class)) +
  geom_boxplot()+
  geom_jitter(shape = 21, alpha = 0.5)+
  theme_classic()+
  theme(legend.position = "top")+
  geom_hline(yintercept = log10(3+1))
s %>% print
```

## QC1 - Detection threshold

We assess the consistency between R1 and R2 barcode detection by comparing the number of reads supporting each barcode in both sequencing reads.

```{r}
## A minimum detection threshold is evaluated based on the correlation between R1 and R2 barcode counts.

st <- data.frame(count_R1 = lapply(DRAG_All$R1_count_raw, function(X){sum(X)}) %>% unlist,
                 count_R2 = lapply(DRAG_All$R2_count_raw, function(X){sum(X)}) %>% unlist,
                 class =DRAG_All$Group)


# -------------------------------------------------------------- 
# Visualize the correlation between R1 and R2 barcode counts.
# --------------------------------------------------------------
# Hexbin plot highlights the density of cells according to their barcode counts.
# Individual cells are overlaid and colored by experimental group.

g1 <- ggplot(st, aes(x = count_R1+1, y = count_R2+1))+
  
  geom_hex(bins = 20, color = "black")  +
  scale_fill_gradientn(colours=brewer.pal(7,"Greys")) +
  
   new_scale_fill() +
  
  geom_jitter(aes(fill = class),
              width = 0.03, height = 0.03, shape = 21)+
  
    # Linear regression to evaluate the correlation between R1 and R2 counts
  geom_smooth(method=lm, se=TRUE, color = "black")+
    # Display regression equation and adjusted R²
  stat_regline_equation(aes(label =  paste(..eq.label.., ..adj.rr.label.., sep = "~~~~")),
                        label.x=0, label.y=2.5, size=2, colour="black") + 
  
  scale_fill_manual(values = cols_Group )+
  
   # Highlight the detection threshold of 3 reads
  geom_vline(xintercept = 3)+
  geom_hline(yintercept = 3)+
  theme_classic2()+ 
  theme(legend.position = "right") +
  scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
              labels = trans_format("log10", math_format(10^.x)), limits = c(-10, 10^3))+
  scale_x_log10(breaks = trans_breaks("log10", function(x) 10^x),
              labels = trans_format("log10", math_format(10^.x)), limits = c(-10, 3*10^3))+
  annotation_logticks(side = "bl", outside = F)
```




```{r}
## Cells with multiple detected barcodes were further inspected.
## We evaluated whether secondary barcodes corresponded to low-count events.

# Identify cells containing more than one detected barcode
idx <- which(lapply(DRAG_All$R1_seq_raw, length) > 1)

# Number of cells containing multiple barcodes before filtering
idx %>% length # 455 cells with multiple bcm if no filter on counts


# Extract read counts of the first detected barcode and additional barcodes
count_first <- lapply(DRAG_All$R1_count_raw[idx], function(X){X[1]}) %>% unlist
count_other <- lapply(DRAG_All$R1_count_raw[idx], function(X){X[-1]}) %>% unlist


# Compare read counts between primary and secondary barcodes
dt <- data.frame(count = c(count_first, count_other),
                 class = rep(c("first", "other"), 
                             times = c(length(count_first), length(count_other))))


# Distribution of barcode counts for primary and secondary barcodes
ggplot(dt, aes(x=class, y = log10(1+count), fill = class)) +
  geom_boxplot(notch = T, outliers = F)+
  geom_jitter(shape = 21, alpha = 0.5, width = 0.03, height = 0.03)+
  theme_classic()+
  theme(legend.position = "top")+
  geom_hline(yintercept = log10(3))

```




```{r}
# -------------------------------------------------------------- 
# Filter out barcodes supported by fewer than 3 reads.
# --------------------------------------------------------------
# This filtering is applied independently to R1 and R2.
# Barcodes below this threshold are considered unreliable detections.


# Keep only barcodes with at least 3 supporting reads in R1
DRAG_All$R1_count_filter <- DRAG_All$R1_count_raw
DRAG_All$R1_seq_filter <- DRAG_All$R1_seq_raw

filtered_counts <- Map(function(count, seq) {
  idx_rm <- which(count < 3)  
  
  if(length(idx_rm) == 0){
    list(count = count, seq = seq)
    }else{
    list(count = count[-idx_rm], seq = seq[-idx_rm])
      }  
  }, 
  DRAG_All$R1_count_raw, 
  DRAG_All$R1_seq_raw)

# Store filtered R1 barcode counts and sequences
DRAG_All$R1_count_filter <- lapply(filtered_counts, `[[`, "count")
DRAG_All$R1_seq_filter <- lapply(filtered_counts, `[[`, "seq")


#######################
# Same filtering applied to R2
DRAG_All$R2_count_filter <- DRAG_All$R2_count_raw
DRAG_All$R2_seq_filter <- DRAG_All$R2_seq_raw

filtered_counts <- Map(function(count, seq) {
  idx_rm <- which(count < 3)  
  if(length(idx_rm) == 0){
    list(count = count, seq = seq)
    }else{
    list(count = count[-idx_rm], seq = seq[-idx_rm]) 
      } 
  },
  DRAG_All$R2_count_raw, 
  DRAG_All$R2_seq_raw)

DRAG_All$R2_count_filter <- lapply(filtered_counts, `[[`, "count")
DRAG_All$R2_seq_filter <- lapply(filtered_counts, `[[`, "seq")


# Replace empty barcode lists by NA values to facilitate downstream analyses
DRAG_All$R1_seq_filter[which(DRAG_All$R1_seq_filter %>% lapply(., length) == 0)] <- NA
DRAG_All$R2_seq_filter[which(DRAG_All$R2_seq_filter %>% lapply(., length) == 0)] <- NA

```

```{r}
# Quantify the number of cells containing multiple barcodes before and after filtering

#R1 
idx <- which(lapply(DRAG_All$R1_seq_raw, length) > 1)
length(idx) # 455 before

res <- which(lapply(DRAG_All$R1_seq_filter, length) > 1)
res %>% length # 97 cells with multiple bcm after filter

#R2 
idx <- which(lapply(DRAG_All$R2_seq_raw, length) > 1)
length(idx) # 383 before

res <- which(lapply(DRAG_All$R2_seq_filter, length) > 1)
res %>% length # 61 cells with multiple bcm after filter

```

```{r}
#Validation of R1/R2 barcode count correlation after filtering


# Recalculate total barcode counts after removing low-confidence barcodes
st <- data.frame(count_R1 = lapply(DRAG_All$R1_count_filter, function(X){sum(X)}) %>% unlist,
                 count_R2 = lapply(DRAG_All$R2_count_filter, function(X){sum(X)}) %>% unlist,
                 class = DRAG_All$Group)


# Same visualization as before, after applying the detection threshold
g3 <- ggplot(st, aes(x = count_R1+1, y = count_R2+1))+
  
  geom_hex(bins = 20, color = "black")  +
  scale_fill_gradientn(colours=brewer.pal(7,"Greys")) +
  
   new_scale_fill() +
  
  geom_jitter(aes(fill = class),
              width = 0.03, height = 0.03, shape = 21)+
  geom_smooth(method=lm, se=TRUE, color = "black")+
  stat_regline_equation(aes(label =  paste(..eq.label.., ..adj.rr.label.., sep = "~~~~")),
                        label.x=0, label.y=2.5, size=2, colour="black") + #from ggpubr package
  
  geom_vline(xintercept = 3)+
  geom_hline(yintercept = 3)+
  scale_fill_manual(values = cols_Group)+
  theme_classic2()+ 
  theme(legend.position = "right") +
  
  scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
              labels = trans_format("log10", math_format(10^.x)), limits = c(-10, 10^3))+
  scale_x_log10(breaks = trans_breaks("log10", function(x) 10^x),
              labels = trans_format("log10", math_format(10^.x)), limits = c(-10, 3*10^3))+
  annotation_logticks(side = "bl", outside = F)

```

```{r}
# Compare barcode detection before and after filtering
grid.arrange(g1,g3, ncol = 2)

```





## QC2 - Remove PRC errors

Multiple barcode sequences detected within the same cell may result from PCR amplification errors generating closely related barcode variants.

We identify and correct potential PCR-derived barcode errors based on:

1. Levenshtein distance between barcode sequences within each cell
2. Relative abundance of barcode variants (fold-change compared to minor variants)

Barcodes with highly similar sequences are merged, while ambiguous barcode assignments are discarded.


```{r}
## Evaluation of intra-cell barcode sequence similarity
##
## Levenshtein distances are computed between barcode sequences detected within the same cell to estimate the frequency of PCR-derived variants.

#R1 
idx <- which(lapply(DRAG_All$R1_seq_filter, length) > 1)
length(idx) # 455 before

res_iC <- lapply(DRAG_All$R1_seq_filter[idx] , 
       function(X){
             # Convert barcode sequences into DNAStringSet format
         dnaSet <- DNAStringSet(X)
         
             # Multiple sequence alignment
         align.msa = msa(dnaSet,method="ClustalOmega")
         
             # Convert alignment object for string distance computation
         alignnment <- msaConvert(align.msa, "seqinr::alignment")
         
             # Compute pairwise Levenshtein distances between barcodes
         stringdistmatrix(  
           a = alignnment$seq[1],  
           b = alignnment$seq[-1], 
           useNames = T,    
           method='lv',  
           weight = c(d = 1, 
                      i = 1, 
                      s = 0.1, 
                      t = 1))} )

# Distribution of intra-cell barcode distances
# Thresholds are visualized to define barcode merging criteria
res_iC %>%   melt() %>% .$value %>% hist(., 100)  ; abline(v=2) ; abline(v=0.2)
```

```{r}
## Barcode correction parameters

# Maximum allowed Levenshtein distance between barcode variants
thr_Lv = 0.2

# Minimum abundance ratio between major and minor barcode variants
thr_FC = 2
```

Correction of R1 barcode assignments

```{r}
# Initialize corrected barcode sequence and count columns
DRAG_All$R1seq_filter_percell <- rep(NA, nrow(DRAG_All))
DRAG_All$R1counts_filter_percell <- rep(NA, nrow(DRAG_All))

# Process each cell independently
for (k in 1:nrow(DRAG_All)) {
  
# Case 1: zero or one barcode detected in the cell
   # No correction is required
  if(length(DRAG_All$R1_seq_filter[[k]]) <= 1){
    
    DRAG_All$R1seq_filter_percell[[k]] <- DRAG_All$R1_seq_filter[[k]]
    
    if ( DRAG_All$R1_count_filter[[k]] %>% length > 0) {
      DRAG_All$R1counts_filter_percell[[k]] =  DRAG_All$R1_count_filter[[k]]
      }else{
          DRAG_All$R1counts_filter_percell[[k]]  <- NA
          }
    
    }else{
   
# Case 2: multiple barcodes detected
    # Compute Levenshtein distance between the major barcode
    # and all secondary barcodes
     
      Leven_intra <- stringdistmatrix(
      a = DRAG_All$R1_seq_filter[[k]][1],
      b = DRAG_All$R1_seq_filter[[k]][-1],
      useNames = TRUE,
      method = "lv",
      weight = c(
        d = 1,
        i = 1,
        s = 0.1,
        t = 1
      )
    ) %>%
      reshape2::melt()
    
# Case 2.1:
    # All barcode variants are within the Levenshtein distance threshold.
    # They are considered PCR-derived variants of the same original barcode.
      
      if( sum(Leven_intra$value > thr_Lv & 
              Leven_intra$value != 0)==0){
        
        # If one barcode is clearly dominant, keep it as reference
        
            if (DRAG_All$R1_count_filter[[k]][1] != DRAG_All$R1_count_filter[[k]][2]) {
            
              selected_bcm <- which(DRAG_All$R1_count_filter[[k]] %>% unlist
                                    == max(DRAG_All$R1_count_filter[[k]] %>% unlist ) )
        
              # Keep the dominant barcode sequence
              DRAG_All$R1seq_filter_percell[[k]] =  
                    DRAG_All$R1_seq_filter[[k]][selected_bcm]
              # Merge counts from all related barcode variantss
               DRAG_All$R1counts_filter_percell[[k]] = 
                    DRAG_All$R1_count_filter[[k]] %>% unlist %>% sum

        # If no dominant barcode exists, randomly select one reference sequence since all variants are highly similar
               
            }else{
              selected_bcm <- sample(1:length(DRAG_All$R1_count_filter[[k]]), size = 1)
               DRAG_All$R1seq_filter_percell[[k]] =  
                    DRAG_All$R1_seq_filter[[k]][selected_bcm]
               DRAG_All$R1counts_filter_percell[[k]] = 
                    DRAG_All$R1_count_filter[[k]] %>% unlist %>% sum
            }
      
# Case 2.2:
    # Some barcode variants are distant, but the major barcode is strongly
    # enriched compared to minor variants.
    # Only close variants are merged.

      }else if(
        sum(Leven_intra$value > thr_Lv
            & Leven_intra$value != 0) !=0 &
               (DRAG_All$R1_count_filter[[k]][1] / sum(DRAG_All$R1_count_filter[[k]][-1])) >= thr_FC
        ){
#Keep the dominant barcode sequence, Merge counts from all related barcode variants
        selected_id <- which(Leven_intra$value <= thr_Lv)
        
        DRAG_All$R1seq_filter_percell[[k]] = DRAG_All$R1_seq_filter[[k]][1]
        DRAG_All$R1counts_filter_percell[[k]] = DRAG_All$R1_count_filter[[k]][selected_id] %>% unlist %>% sum
        
# Case 2.3:
    # Ambiguous barcode assignment.
    # Barcode is discarded.
        
      }else{ 
        DRAG_All$R1seq_filter_percell[[k]] = "uncertain"
        DRAG_All$R1counts_filter_percell[[k]] = "uncertain"
      }
    }
}


```



The same PCR-error correction strategy was applied independently to R2 barcodes.
The same Levenshtein distance and abundance ratio thresholds were used.
```{r}
DRAG_All$R2seq_filter_percell <- rep(NA, nrow(DRAG_All))
DRAG_All$R2counts_filter_percell <- rep(NA, nrow(DRAG_All))

for (k in 1:nrow(DRAG_All)) {
  
  if(length(DRAG_All$R2_seq_filter[[k]]) <= 1){
    DRAG_All$R2seq_filter_percell[[k]] = DRAG_All$R2_seq_filter[[k]]

    
    if ( DRAG_All$R2_count_filter[[k]] %>% length > 0) {
      DRAG_All$R2counts_filter_percell[[k]] =
        DRAG_All$R2_count_filter[[k]]}else{
          DRAG_All$R2counts_filter_percell[[k]]  = NA}
    
    }else{
   
      Leven_intra <- stringdistmatrix( a = DRAG_All$R2_seq_filter[[k]][1], 
                                       b = DRAG_All$R2_seq_filter[[k]][-1], useNames = T,
                      method='lv', weight = c(d = 1, i = 1, s = 1, t = 1))  %>% reshape2::melt()
    
      if( sum(Leven_intra$value > thr_Lv & Leven_intra$value != 0)==0){
            if (DRAG_All$R2_count_filter[[k]][1] != DRAG_All$R2_count_filter[[k]][2]) {
            
              selected_bcm <- which(DRAG_All$R2_count_filter[[k]] %>% unlist == DRAG_All$R2_count_filter[[k]] %>% unlist %>% max() )
        
               DRAG_All$R2seq_filter_percell[[k]] =  
                    DRAG_All$R2_seq_filter[[k]][selected_bcm]
               DRAG_All$R2counts_filter_percell[[k]] = 
                    DRAG_All$R2_count_filter[[k]] %>% unlist %>% sum
            }else{
              selected_bcm <- sample(1:length(DRAG_All$R2_count_filter[[k]]), size = 1)
               DRAG_All$R2seq_filter_percell[[k]] =  
                    DRAG_All$R2_seq_filter[[k]][selected_bcm]
               DRAG_All$R2counts_filter_percell[[k]] = 
                    DRAG_All$R2_count_filter[[k]] %>% unlist %>% sum
            }
      
      }else if(sum(Leven_intra$value > thr_Lv & Leven_intra$value != 0) !=0 &
               (DRAG_All$R2_count_filter[[k]][1] / sum(DRAG_All$R2_count_filter[[k]][-1]) ) >= thr_FC){
        
        selected_id <- which(Leven_intra$value <= thr_Lv)
        
        DRAG_All$R2seq_filter_percell[[k]] = DRAG_All$R2_seq_filter[[k]][1]
        DRAG_All$R2counts_filter_percell[[k]] = DRAG_All$R2_count_filter[[k]][selected_id] %>% unlist %>% sum
        
      }else{ 
        DRAG_All$R2seq_filter_percell[[k]] = "uncertain"
        DRAG_All$R2counts_filter_percell[[k]] = "uncertain"
      }
    }
}
```



```{r}
# Convert corrected barcode counts to numeric format
DRAG_All$R1counts_filter_percell <- as.numeric(DRAG_All$R1counts_filter_percell)
DRAG_All$R2counts_filter_percell <- as.numeric(DRAG_All$R2counts_filter_percell)

# Quantify the number of unresolved barcode assignments after correction
idx2 <- which( DRAG_All$R1seq_filter_percell ==  "uncertain") 
idx2%>% length ## Number of cells for which barcode correction failed: 4

```

## By cell, combine read1 and read2 barcode information

```{r}
### Validate barcode concordance between R1 and R2

# Load Biostrings to compute reverse complements of R2 barcode sequences.
library(Biostrings)

DRAG_All$R1seq_filter_percell[2] %>% DNAString(); 
DRAG_All$R2seq_filter_percell[2] %>% DNAString() %>% reverseComplement()

```



```{r}
# Summarize the proportion of cells with concordant R1 and R2 barcode sequences.
res = rep(NA, nrow(DRAG_All))
for (k in 1:nrow(DRAG_All)) {
  if(is.na(DRAG_All$R1seq_filter_percell[k]) == F &
     DRAG_All$R1seq_filter_percell[k] != "uncertain" & 
     is.na(DRAG_All$R2seq_filter_percell[k]) == F & 
     DRAG_All$R2seq_filter_percell[k] != "uncertain" 
     ){
    if (length(DNAString(DRAG_All$R1seq_filter_percell[k])) > 3) {
      
    
    res[k] =  reverseComplement(DNAString( DRAG_All$R2seq_filter_percell[k]) )== DNAString( DRAG_All$R1seq_filter_percell[k] )
  }}else{res[k] = NA}
}
sum(res, na.rm = T) / length(na.omit(res)) * 100 # 99.5519% R1 == R2



#plot
data.frame(NBCM = c(sum(res, na.rm = T),length(na.omit(res)) ),
                 type = factor(c("R2 == R1", "All"),
                               levels = c( "R2 == R1","All" )) ,
                 x = c(1, 1)) %>%
  mutate(freq = NBCM / length(na.omit(res)) * 100) %>%

ggplot(aes(x = type,  y=freq, fill=type) ) +
  geom_bar(position = "stack", stat="identity", color = "black")+
  geom_text(aes(y=freq+10, label=NBCM), vjust=1,  color="black", size=3.5)+
  scale_y_continuous(limits = c(0,120))+
  theme_classic()
```


```{r}
### Rescue barcode assignments using R2

# For cells lacking a confident R1 barcode but containing a confident R2 barcode, recover the barcode sequence from the reverse-complemented R2 read. Otherwise, retain the R1 barcode assignment.

FUN_revComp_R1 <- function(X){
  
  R1 = X$R1seq_filter_percell
  R2 = X$R2seq_filter_percell
  res = c()
  
  if (is.na(R1) == T & is.na(R2) ==F & R2 != "uncertain") {
    res = R2 %>% DNAString() %>% reverseComplement() %>% as.character 
    
  }else if (is.na(R1) == F) {
    res = R1
    
  }else{
    res = c(NA)
  }
  
  res
}

DRAG_All$BCM_seq_noLV_beforeFilterMultipleMouse <- rep(NA, nrow(DRAG_All))

DRAG_All$BCM_seq_noLV_beforeFilterMultipleMouse <- apply(DRAG_All, 1, FUN_revComp_R1) %>% unlist

# Number of additional cells rescued by using R2 information.
sum(is.na(DRAG_All$R1seq_filter_percell)) - sum(is.na(DRAG_All$BCM_seq_noLV_beforeFilterMultipleMouse))
#88 saved!
```


```{r}
### Characterize rescued barcode detections

# Identify cells rescued exclusively from R2 and compare their barcode read counts with all other detected barcodes.

id_R2 <- which(is.na(DRAG_All$BCM_seq_noLV_beforeFilterMultipleMouse) ==F & is.na(unlist(DRAG_All$R1seq_filter_percell)) == T) %>% as.vector()

dt <- data.frame(counts = c(as.vector(DRAG_All$R2counts_filter_percell)[id_R2],
                            as.vector(DRAG_All$R2counts_filter_percell)[-id_R2]),
                 count_tot = c(as.vector(DRAG_All$nCount_RNA)[id_R2],
                            as.vector(DRAG_All$nCount_RNA)[-id_R2]),
                 type = c(rep("saveR2", length(id_R2)),
                          rep("other", c(length(DRAG_All$R2counts_filter_percell) - length(id_R2)))),
                 mouse = as.factor(c( DRAG_All$Group[id_R2],
                            DRAG_All$Group[-id_R2])) )

ggplot(data = dt[-which(is.na(dt$counts)==T),], 
       aes(x = mouse, y = counts+1, fill = type))+
  geom_boxplot(outlier.size = 0)+
  geom_point(shape = 21, position = position_jitterdodge(), 
              alpha = 0.6)+
  geom_hline(yintercept = c(1,4))+
  scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
              labels = trans_format("log10", math_format(10^.x)), limits = c(1, 2*10^3))+
  theme_classic()+ 
  theme(legend.position = "right") +
  scale_fill_manual(values = (colorRampPalette(brewer.pal(8, "Spectral"))(2)[c(2,1)]))

# Compare total coverage between rescued cells and all other barcode-positive cells.
ggplot(data = dt[-which(is.na(dt$counts)==T),], 
       aes(x = mouse, y = count_tot+1, fill = type))+
  geom_boxplot(outlier.size = 0)+
  geom_point(shape = 21, position = position_jitterdodge(), 
              alpha = 0.6)+
  scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
              labels = trans_format("log10", math_format(10^.x)), limits = c(1, 2*10^4))+
  theme_classic()+ 
  theme(legend.position = "right") +
  scale_fill_manual(values = (colorRampPalette(brewer.pal(8, "Spectral"))(2)[c(2,1)]))

# Compare barcode read depth between rescued and non-rescued cells.
ggplot(data = dt[-which(is.na(dt$counts)==T),], 
       aes(x = type, y = counts+1, fill = type))+
  geom_boxplot(outlier.size = 0, notch = T)+
  geom_point(shape = 21, position = position_jitterdodge(jitter.width = 0.1), 
              alpha = 0.6)+
  scale_x_discrete(limits = c( "other", "saveR2"))+
  scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
              labels = trans_format("log10", math_format(10^.x)), limits = c(1, 2*10^3))+
  theme_classic()+ 
  theme(legend.position = "none") +
  scale_fill_manual(values = (colorRampPalette(brewer.pal(8, "Spectral"))(2)[c(2,1)]))
```


```{r}
### Merge barcode read counts

# Combine R1 and R2 barcode read counts into a single per-cell
# barcode abundance measure.

DRAG_All$BCM_counts_noLV_beforeFilterMultipleMouse <- apply(
  DRAG_All, 
  1, 
  function(X){
  sum(c(X$R1counts_filter_percell  , 
        X$R2counts_filter_percell), 
      na.rm = T )
    }
  )

head(DRAG_All$BCM_counts_noLV_beforeFilterMultipleMouse )
```




##QC3 - Remove shared bcm1 bw mive

```{r}
### Identify recurrent lineage barcodes across mice

# Compile the list of unique lineage barcodes detected in each mouse,
# remove missing and ambiguous barcode assignments, 
# and identify barcodes observed in more than one mouse (blacklist).

List_bcm = by(DRAG_All$BCM_seq_noLV_beforeFilterMultipleMouse, DRAG_All$MouseNumber, unique)

List_bcm <- lapply(List_bcm, function(x){x[!is.na(x)]}) #remove NA
List_bcm <- lapply(List_bcm, function(x){x[which(x %in% c("tooClose", "uncertain")==F)]}) #remove NA
List_bcm <- unlist(List_bcm)
Blacklist <- List_bcm[duplicated(List_bcm)]

length(Blacklist) #136 - 38%
length(Blacklist) / DRAG_All$BCM_seq_noLV_beforeFilterMultipleMouse %>% unique %>% length() * 100

```


```{r}
### Remove recurrent barcode assignments

# Remove recurrent (blacklisted) and uncertain barcode assignments,
# together with their associated barcode read counts.

DRAG_All$BCM_STRINGEANT <- DRAG_All$BCM_seq_noLV_beforeFilterMultipleMouse

id <- which(DRAG_All$BCM_STRINGEANT %in% c("uncertain", Blacklist))
DRAG_All$BCM_STRINGEANT[id] <- NA

DRAG_All$BCM_counts <- DRAG_All$BCM_counts_noLV_beforeFilterMultipleMouse
DRAG_All$BCM_counts[id] <- NA
```


```{r}
### Generate the final filtered barcode annotation : Store the final filtered lineage barcode annotation for downstream analyses.
DRAG_All$Barcode <- DRAG_All$BCM_STRINGEANT
```

```{r}
### Compute clone size

# Calculate clone size as the number of cells carrying each lineage
# barcode before and after recurrent barcode filtering.

# Clone size before filtering
taille <- table(DRAG_All$BCM_seq_noLV_beforeFilterMultipleMouse)
DRAG_All$BCM_CloneSize_noLV <- taille[DRAG_All$BCM_seq_noLV_beforeFilterMultipleMouse]

# Clone size after filtering
taille <- table(DRAG_All$BCM_STRINGEANT)
DRAG_All$BCM_CloneSize_STRINGEANT <- taille[DRAG_All$BCM_STRINGEANT]
```


```{r}
### Characterize recurrent barcodes

# Compare the clone sizes of recurrent (blacklisted) barcodes
# with those of retained lineage barcodes.

id <- which(DRAG_All$BCM_seq_noLV_beforeFilterMultipleMouse %in% c( "uncertain", Blacklist))

dt <- data.frame(size = c(as.vector(DRAG_All$BCM_CloneSize_noLV)[id],
                            as.vector(DRAG_All$BCM_CloneSize_noLV)[-id]),
                 type = c(rep("Blacklist", 
                              length(id)),
                          rep("other", 
                              c( length( DRAG_All$BCM_seq_noLV_beforeFilterMultipleMouse ) - length( id )
                                 )
                              )
                          ),
                 mouse = as.factor(c( DRAG_All$Group[id],
                            DRAG_All$Group[-id])) 
                 )

ggplot(data = dt, 
       aes(x = mouse, y = (size), fill = type))+
  geom_boxplot(outlier.size = 0)+
  geom_point(shape = 21, position = position_jitterdodge(), 
              alpha = 0.6)+
  scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
              labels = trans_format("log10", math_format(10^.x)))+
  theme_classic()+ 
  scale_fill_manual(values = (colorRampPalette(brewer.pal(8, "Spectral"))(2)))
```






## QC4 - Add Pgen


```{r}
# Add the full barcode sequence used for Pgen estimation by concatenating:
# - the constant 5' and 3' flanking regions of the DRAG barcode
# - the variable barcode sequence
# Then retrieve the corresponding Pgen value from the reference Pgen table.

DRAG_All <- DRAG_All %>%
  mutate(Barcode_long = paste0(
    toupper("acctcctcgaggtcatcgaagtatcaagtcca"), # 5' constant region
    BCM_seq_noLV_beforeFilterMultipleMouse,                                     # variable barcode sequence
    toupper("accgtagcaagctcgagagtagac")          # 3' constant region
                             )
    ) %>%
  
  # Add Pgen values by matching the reconstructed barcode sequence
  # with the reference table containing barcode sequences and their generation probabilities.
  left_join(
    Pgen_table %>% 
      dplyr::select(sequence, Pgen) %>% 
      distinct(),
    by = c("Barcode_long" = "sequence")
  ) 
head(DRAG_All, 15)

```




# FIGURES


## Pgen vs Nmice
```{r}

# Compute the number of distinct mice in which each barcode is detected.
# This allows assessing whether recurrent/shared barcodes are enriched
# among barcodes with high probability of generation (high Pgen).

# Count the number of distinct mice in which each barcode is detected.

DRAG_All <- DRAG_All %>%
  ungroup() %>% 
  group_by(BCM_seq_noLV_beforeFilterMultipleMouse) %>%
  mutate(Barcode_sharing_mice_N = n_distinct(MouseNumber))


# Summarize one entry per barcode and classify barcodes according
# to the number of mice in which they are detected.

df <-DRAG_All %>%
  dplyr::filter(!is.nan(Pgen),!is.na(Pgen),  is.finite(Pgen)) %>%  
  ungroup() %>% 
  group_by(BCM_seq_noLV_beforeFilterMultipleMouse) %>%
  summarise(
    Pgen = dplyr::first(Pgen),
    Barcode_sharing_mice_N = max(Barcode_sharing_mice_N),
    .groups = "drop"
  ) 

# Define a grid of Pgen thresholds (10^-x).
thr_grid <- seq(0, 11, by = 0.5)

# For each threshold, calculate the fraction of barcodes
# shared by more than one mouse.

res <- lapply(thr_grid, function(x){

  thr <- 10^-x

  sub <- df %>% dplyr::filter(Pgen <= thr)

  if(nrow(sub) == 0){
    return(data.frame(x = x, frac = NA))
  }

  frac_shared <- mean(sub$Barcode_sharing_mice_N > 1)

  data.frame(
    x = x,
    frac = frac_shared
  )

}) %>% bind_rows()

# Record the number of retained barcodes at each threshold.
res$n <- sapply(thr_grid, function(x){
  sum(df$Pgen <= 10^-x)
})



# Plot the fraction of shared barcodes as a function of the Pgen threshold.
ggplot(res, aes(x = x, y = frac)) +
  geom_point(size = 3, shape = 21) +
  geom_text(aes(label = n), vjust = -1, size = 3)+
  labs(
    x = expression("Pgen threshold 10"^-x),
    y = "Fraction barcode shared between mice"
  ) +
  theme_classic()
```



```{r}
# Classify samples according to whether they correspond to pooled glands
# or individual gland-level samples.
DRAG_All <- DRAG_All %>%
  mutate(type_gl = case_when(WhichGland %in% c("AllGl.Pooled", "All") ~"AllGl.Pooled",
                             TRUE ~"Mono.Gl"))
```

```{r}
DRAG_All %>%
  filter(!is.na(BCM_STRINGEANT)) %>% filter(Group != "CreN") %>%
  group_by(SampleBio, type_gl) %>%
  mutate(
    n_barcodes_sample = n_distinct(BCM_STRINGEANT),
    .groups = "drop"
  ) %>% ungroup() %>%
  group_by(type_gl) %>%
  summarize(
         mean = mean(n_barcodes_sample))
```
Average number of detected clones per sample and gland type
```{r}
# Calculate the average number of distinct barcodes (clones)
# detected per sample for each gland processing strategy.
# CreN control samples are excluded.

DRAG_All %>%
  filter(!is.na(BCM_STRINGEANT)) %>% filter(Group != "CreN") %>%
  group_by(Group) %>%
  summarize(
    n_barcodes_sample = n_distinct(BCM_STRINGEANT),
    .groups = "drop"
  ) 

```

Clone number comparison between samples and groups


```{r}
# Calculate:
# - number of distinct clones detected in each biological sample
# - average clone number within each experimental group and gland type


df_nclone <- DRAG_All %>%
  filter( !is.na(BCM_STRINGEANT)) %>%
  
  group_by(SampleBio, Group, type_gl) %>%
  mutate(
    n_barcodes_sample = n_distinct(BCM_STRINGEANT),
    .groups = "drop"
  )%>% ungroup() %>%
  
    # Compute the mean number of clones for each group and gland type
  group_by( Group, type_gl) %>%
  mutate(
    n_barcodes_group = mean(n_barcodes_sample),
    .groups = "drop"
  )%>%
  
  select(Group, 
         SampleBio, 
         n_barcodes_group, 
         n_barcodes_sample , 
         type_gl) %>% 
  distinct()


df_nclone %>%
  filter(type_gl=="Mono.Gl") %>%
  filter(SampleBio %in% c("8415_G2-3", "8718_D2-3", "8964_G4")) %>%
ggplot( aes(x = Group, y = n_barcodes_group, fill = Group)) +
  geom_bar(stat = "identity", position = position_dodge()) +
    geom_point(aes(y =n_barcodes_sample), 
             shape = 21,  position = position_jitter()) +
  theme_classic() +
  labs(y = "Number of distinct barcodes",
       x = "Sample")
```











##  Summary barplots: barcode recovery and clone complexity metrics
```{r}
# Create a unique biological sample identifier by combining:
# - mouse identifier
# - gland identifier
DRAG_All$SampleBio <- paste0(DRAG_All$MouseNumber, "_", DRAG_All$WhichGland)

# Manually replace missing mouse identifier for one sample
DRAG_All$SampleBio[which(DRAG_All$SampleBio  == "NA_G2-3")] <- "8415_G2-3"



## Barcode detection rate before QC3

# Calculate the percentage of cells containing a detected BCM barcode
# before removal of barcodes shared between multiple mice (QC3).
#
# For each biological sample:
# - numerator: number of cells assigned to a barcode
# - denominator: total number of cells

res = table(
  DRAG_All$BCM_seq_noLV_beforeFilterMultipleMouse,
  DRAG_All$SampleBio) %>% 
  colSums() / 
  table(DRAG_All$SampleBio) *
  100 %>% 
  round(., 0)


# Build summary dataframe containing:
# - total number of cells
# - number of cells with barcode detection
# - barcode recovery frequency
# - gland processing type
# - experimental group

dt <- data.frame(
  Ncell = colSums(table(
  DRAG_All$BCM_seq_noLV_beforeFilterMultipleMouse, 
  DRAG_All$SampleBio
  ))  ,
  Ncell_wBCM = table(DRAG_All$SampleBio) ,
  frequency = res,
  glandType = c("Mix", "Mix", "Mix",
                              "Mono", "Mono", "Mono","Mono", "Mono", 
                              "Mix", "Mix",
                              "Mono", "Mono", "Mono", "Mix", "Mix", NA, NA),
  group = c("CreP_Late", "CreP_Early", "CreP_Early",
                          "CreP_Late", "CreP_Late", "CreP_Late", "CreP_Late","CreP_Late",
                          "CreN","CreN",
                          "CreP_Early","CreP_Early", "CreP_Early",
                          "CreN",  "CreN","CreN","NA"))
by(dt$Ncell, dt$group, sum)


# Compute mean barcode recovery rate per experimental group
dt <- dt %>% 
  group_by(group ) %>%
  mutate(mean = mean(frequency.Freq))

# Plot barcode detection rate before QC3
# Dots represent individual samples
# Dot size represents the number of analyzed cells

g1 <- ggplot(dt, aes(x = group, y = frequency.Freq, fill = group))+
  geom_bar(aes(y = mean), position="identity", stat="identity", color = "black")+
  scale_y_continuous(limits = c(0,100))+
  geom_point(aes(size = Ncell), 
             shape =21, position = position_jitter(width = 0.2))+
  geom_text(aes(y=frequency.Freq, label=Ncell), vjust=1.6,  color="black", size=3.5,
            position = position_jitter(width = 0.2,height = 0))+
  theme_classic()
```


```{r}
# Calculate barcode recovery after QC filtering.
# Only barcodes retained after QC are considered (BCM_STRINGEANT).


res = table(
  DRAG_All$BCM_STRINGEANT, 
  DRAG_All$SampleBio) %>% 
  colSums() / 
  table(DRAG_All$SampleBio) *
  100 %>% 
  round(., 0)

dt <- data.frame(Ncell = colSums(table(
  DRAG_All$BCM_STRINGEANT, 
  DRAG_All$SampleBio))  ,
                 Ncell_wBCM = table(DRAG_All$SampleBio) ,
                 frequency = res,
                glandType = c("Mix", "Mix", "Mix",
                              "Mono", "Mono", "Mono","Mono", "Mono", 
                              "Mix", "Mix",
                              "Mono", "Mono", "Mono", "Mix", "Mix", NA, NA),
                group = c("CreP_Late", "CreP_Early", "CreP_Early",
                          "CreP_Late", "CreP_Late", "CreP_Late", "CreP_Late","CreP_Late",
                          "CreN","CreN",
                          "CreP_Early","CreP_Early", "CreP_Early",
                          "CreN",  "CreN","CreN","NA"))
by(dt$Ncell, dt$group, sum)

dt <- dt %>% 
  group_by(group ) %>%
  mutate(mean = mean(frequency.Freq))


g2 <- ggplot(dt, aes(x = group, y = frequency.Freq, fill = group))+
  geom_bar(aes(y = mean), position="identity", stat="identity", color = "black")+
  scale_y_continuous(limits = c(0,100))+
  geom_point(aes(size = Ncell), 
             shape =21, position = position_jitter(width = 0.2))+
  geom_text(aes(y=frequency.Freq, label=Ncell), vjust=1.6,  color="black", size=3.5,
            position = position_jitter(width = 0.2,height = 0))+
  theme_classic()

# Compare barcode recovery before and after QC
grid.arrange(g1, g2, ncol = 2)
```



#SAVE
```{r}
qs::qsave(DRAG_All,
          paste0(outputDir, "sc_Metadata_withAllDRAG_filt.qs"))

```





