---
title: "R Notebook"
output: html_notebook
---
#Intro

To evaluate the relationship between, respectively, clone poly-functionnality, intra-clonal transcriptional heterogeneity, or cell fate diversification and the probability of barcode generation (Pgen), clone-level number of FPs, distances to centroid and cell fate transition score were integrated with Pgen values derived from the IGOR pipeline. Analyses were restricted to clones with available Pgen estimates. 
We then assessed how these metrics varied as a function of Pgen thresholds. A series of decreasing Pgen cutoffs (10⁰ to 10⁻⁷ in logarithmic steps) was applied, and for each threshold, clones with Pgen below the cutoff were retained. Within each subset, we summarized the median score for each of these metrics, stratified by either clone size or mouse age. 


##Librairies
```{r}
library(rstatix)
library(ggpubr)
library(purrr)

library(Signac) ; library(Seurat) ; library(ggplot2)

library(mgcv)
library(entropy)

library(data.table)
library(tidyverse)
library(ggraph)
library(dplyr)
library(readxl)
library(ggrepel)
```


##Data 

```{r}
Res_dir <- "C:/Users/Marthe Laisne/OneDrive - INSTITUT CURIE/m_laisne/Publications/2026_Laisne/Scripts/Output/sc_DRAG_Analyses"

```



```{r}
seurat_rna <- qs::qread(file.path("C:/Users/Marthe Laisne/OneDrive - INSTITUT CURIE/m_laisne/Publications/2026_Laisne/Data/seurat_rna_step7_new.qs"))


seurat <- qs::qread(file.path("C:/Users/Marthe Laisne/OneDrive - INSTITUT CURIE/m_laisne/Publications/2026_Laisne/Data/seurat_step5_new.qs"))
```

##Pre-process data Pre-tumoral set
Subset: only CreP GFP+ pre-tumoral cells
```{r}
seu <- subset(seurat_rna, Group %in% c( "CreN", "Rosa"), invert = T)


seuEpi <- subset(seurat, Group %in% c( "CreN", "Rosa"), invert = T)
```

```{r}
#Normalize & RD
seu <- NormalizeData(seu)
seu <- FindVariableFeatures(seu)
seu <- ScaleData(seu)
seu <- RunPCA(seu, 
              features = VariableFeatures(seu), 
              npcs = 50)
```


Epigenome:
```{r}
DefaultAssay(seuEpi) <- "bin_50k_h3k4me1"




DefaultAssay(seuEpi) <- "bin_50k_h3k27me3"

#Normalize & RD
seuEpi <- RunTFIDF(seuEpi)
seuEpi <- FindTopFeatures(seuEpi)
seuEpi <- RunSVD(seuEpi)
seuEpi <- RunUMAP(seuEpi, reduction = 'lsi',  dims = 2:30)
```


## Define clone size categories 

```{r}
library(dplyr)

# Create a unique sample identifier combining mouse and gland IDs
seu@meta.data$SampleBio <- paste0(seu@meta.data$MouseNumber, "_", seu@meta.data$WhichGland )


# Compute clone size and relative abundance within each sample
meta <- seu@meta.data %>%
  as.data.frame() %>%
  dplyr::filter(!is.na(Barcode)) %>%                              #  retain barcoded cells only
  dplyr::group_by(SampleBio) %>%
  dplyr::mutate(TotalCells = n()) %>%                             # total number of barcoded cells per sample
  dplyr::group_by(SampleBio, Barcode) %>%
  dplyr::mutate(CloneSize = n()) %>%                              # number of cells per clone
  dplyr::ungroup() %>%
  dplyr::mutate(PercentClone = 100 * CloneSize / TotalCells)      # normalized clone abundance

# Inspect clone size distribution
meta %>% 
  dplyr::select(CloneSize, SampleBio) %>% 
  distinct %>% 
  arrange(CloneSize)

# Merge clone-level information back into Seurat metadata
meta2 <- data.frame(seu@meta.data )
head(meta2)
meta2 <- meta2 %>% 
  dplyr::left_join(meta %>% dplyr::select(SampleBio, Barcode, CloneSize, TotalCells, PercentClone),
                           by = c("Barcode"),multiple = "first")
head(meta2)
rownames(meta2) <- rownames(seu@meta.data )

seu@meta.data <- meta2

# Retain normalized clone abundance only for monoclonal glands
seu@meta.data <- seu@meta.data %>% 
  mutate(BCM_PercentClone_MonoGl = case_when(SampleBio.x %in% 
                                               c("8964_G4", "8718_D2-3", "8415_G2-3") ~ PercentClone,
                                             TRUE ~ NA))

# Classify clones according to their relative abundance
seu@meta.data <- seu@meta.data %>%
  mutate(ExpandedClone = case_when(SampleBio.x %in% c("8964_G4", "8718_D2-3", "8415_G2-3") & 
                                     PercentClone >= 10 ~"Expanded",
                                   SampleBio.x %in% c("8964_G4", "8718_D2-3", "8415_G2-3") & 
                                     PercentClone < 2 ~"Small",
                                   SampleBio.x %in% c("8964_G4", "8718_D2-3", "8415_G2-3") & 
                                     PercentClone < 10 &
                                     PercentClone >= 2 ~"Int",
                                   TRUE ~ NA))

table(seu@meta.data$ExpandedClone)
```




# Clone expansion and pseudotime


```{r}
plot_df <- seu@meta.data %>%
  filter(!is.na(Barcode),
         !is.na(pseudotime),
         !is.na(CloneSize)) %>%
  group_by(Barcode) %>%
  summarise(
    pseudotime = median(pseudotime),
    CloneSize  = dplyr::first(CloneSize),      
    ExpandedClone = dplyr::first(ExpandedClone),
    n_cells = n(),
    .groups = "drop"
  )

ggplot(plot_df,
       aes(x = pseudotime,
           y = (CloneSize))) +
  geom_point(aes( colour = ExpandedClone), size = 2.5, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, colour = "black") +
  stat_regline_equation(
    aes(label = paste(..rr.label.., sep = "~~~~")),
    label.x.npc = "left",
    label.y.npc = "top",
    colour = "black"
  ) +
  theme_classic() +
  labs(
    x = "Mean clone pseudotime",
    y = "Clone size",
    colour = "Expanded clone"
  )
```


#Clone-level transcriptional heterogeneity:

Compute distance to clone centroid in PCA space

```{r}
N_PC <- 10 # Number of principal components used

# Extract PCA embeddings
pcs <- Embeddings(seu,
                  reduction = "pca")[,1:N_PC] %>% as.data.frame()

# Build cell-level data frame
dt <- data.frame(Clone =seu$Barcode,
                       Expand =seu$ExpandedClone,
                       Sample = seu$SampleBio.x,
                       Group = seu$Group) %>%
  cbind(pcs)

head(dt)


# --------------------------------------------------------------
# Compute clone centroids in PCA space
# --------------------------------------------------------------
centroids <- dt %>%
  dplyr::group_by(Clone) %>%
  dplyr::summarise(across(paste0("PC_", 1:N_PC), ~ mean(.x, na.rm = TRUE), .names = "c{.col}"))
head(centroids)

# Add centroid coordinates to each cell
df_dist <- dt %>%
  left_join(centroids, by = c("Clone"))



# --------------------------------------------------------------
# Compute Euclidean distance between each cell and the centroid of its corresponding clone
# --------------------------------------------------------------
df_dist <- df_dist %>%
  dplyr::add_count(Clone, name = "n_cells") %>%  
  dplyr::filter(n_cells >= 2) %>%          # retain multicellular clones only
  group_by(Clone) %>%
  dplyr::rowwise() %>%
  dplyr::mutate( dist_to_centroid = sqrt(sum(
      (dplyr::c_across(starts_with("PC_")) -
       dplyr::c_across(starts_with("cPC_")))^2
    ))
  ) %>%
  ungroup()


# --------------------------------------------------------------
# Export table
# --------------------------------------------------------------
df_dist %>% write.table(., paste0(Res_dir, "/RNADistCentroid.txt"),
                              quote = F, sep = "\t", row.names = F)
```



```{r}
# --------------------------------------------------------------
# Compute clone-level summary statistics
# --------------------------------------------------------------

df_clone_dist <- df_dist %>%
  group_by(Clone) %>%
  summarise(
    mean_dist = mean(dist_to_centroid, na.rm = TRUE),
    median_dist = median(dist_to_centroid, na.rm = TRUE),
    n_cells = dplyr::first(n_cells),
    ExpandedClone = dplyr::first(Expand),
    .groups = "drop"
  )


# Add Pgen estimates
df_clone_dist <- df_clone_dist %>%
  left_join(seurat_rna@meta.data[, c("Barcode", "Pgen")] , by = c("Clone" = "Barcode")) %>%
  distinct()

df_dist <- df_dist %>%
  left_join(seurat_rna@meta.data[, c("Barcode", "Pgen")] , by = c("Clone" = "Barcode")) %>%
  distinct()


head(df_clone_dist)
```

Evaluate the effect of increasingly stringent Pgen filtering
```{r}
# --------------------------------------------------------------
# Define a series of decreasing Pgen thresholds
# --------------------------------------------------------------
thresholds <- 10^seq(0, -7, by = -0.5)
thresholds
```

```{r}
# --------------------------------------------------------------
# For each threshold, retain clones with lower Pgen values
# and compute transcriptional heterogeneity statistics
# --------------------------------------------------------------

res_pgen_dist <- map_dfr(thresholds, function(thr){

  tmp <- df_clone_dist %>%
    filter(Pgen <= thr)

  tmp %>%
    group_by(ExpandedClone) %>%
    summarise(
      threshold = thr,
      n_clones = n(),
      mean_dist = mean(mean_dist, na.rm = TRUE),
      median_dist = median(mean_dist, na.rm = TRUE),
      sem = sd(mean_dist, na.rm = TRUE)/sqrt(n()),
      .groups = "drop"
    )
})

ggplot(subset(res_pgen_dist, is.na(ExpandedClone )==F), 
       aes(x = -log10(threshold),  y = median_dist,   colour = ExpandedClone)) +
  geom_line(size = 1) +
  theme_classic() +
   geom_text(aes(label = n_clones),
    vjust = -1,    size = 3,    show.legend = FALSE  )+
  scale_color_manual(values = c("#d0e0ea","#4384ac", "#0a334c") %>%rev)+
  labs(x = "Maximum Pgen retained (-log10)",
       y = "Mean distance to centroid")
# --------------------------------------------------------------
# For each threshold, AOV
# --------------------------------------------------------------
stats <- purrr::map_dfr(thresholds[1:12], function(thr){
  dat <- df_dist %>%
    filter(  Pgen <= thr , 
              is.na(Expand)==F)

  fit <- aov(dist_to_centroid  ~ Expand, data = dat)

  broom::tidy(fit) %>%
    filter(term == "Expand") %>%
    transmute(
      Thr = thr,
      logThr = -log10(thr),
      p = p.value
    )
})

stats <- stats %>%
  mutate(
    y.position = max(res_pgen_dist$median_dist, na.rm = TRUE) * 1.05,
    label = paste0(signif(p, 2))
  )
stats

# --------------------------------------------------------------
# Export table
# --------------------------------------------------------------
res_pgen_dist %>% write.table(., paste0(Res_dir, "/RNADistCentroid_Pgen.txt"),
                              quote = F, sep = "\t", row.names = F)

```





#Clone-level epigenetic heterogeneity:

> distance to clone centroid in chromatin space
 
Clone-level epigenetic heterogeneity was quantified as the Euclidean distance between individual cells and the centroid of their corresponding clone in latent semantic indexing (LSI) space. Analyses were performed independently for H3K4me1 and H3K27me3 CUT&Tag datasets.

Only multicellular clones (≥ 2 cells) were considered. Clone expansion categories were transferred from matched transcriptomic profiles using CellID_RNA identifiers.

## H3K4me1 dataset
```{r}
DefaultAssay(seuEpi) <- "bin_50k_h3k4me1"
#Normalize & RD
seuEpi <- RunTFIDF(seuEpi)
seuEpi <- FindTopFeatures(seuEpi)
seuEpi <- RunSVD(seuEpi)
seuEpi <- RunUMAP(seuEpi, reduction = 'lsi',  dims = 2:30)

# Number of LSI dimensions used
N_PC <- 11

# Extract LSI embeddings (excluding the first component)
pcs <- Embeddings(seuEpi,
                  reduction = "lsi")[,2:N_PC] %>% as.data.frame()

# Create cell-level metadata table
dt <- data.frame(Clone =seuEpi@meta.data$LineageBarcode[which(seuEpi@meta.data$mark == "h3k4me1" )],
                 BC_RNA = seuEpi@meta.data$BC_RNA[which(seuEpi@meta.data$mark == "h3k4me1" )],
                       Sample = seuEpi@meta.data$SampleBio[which(seuEpi@meta.data$mark == "h3k4me1" )],
                       Group = seuEpi@meta.data$Group[which(seuEpi@meta.data$mark == "h3k4me1" )]) %>%
  cbind(pcs)

# --------------------------------------------------------------
# Compute clone centroids in LSI space
# --------------------------------------------------------------
centroids <- dt %>%
  dplyr::group_by(Clone) %>%
  dplyr::summarise(across(paste0("LSI_", 2:N_PC), ~ mean(.x, na.rm = TRUE), .names = "c{.col}"))

# Add centroid coordinates to each cell
df_dist <- dt %>%
  left_join(centroids, by = c("Clone"))

# --------------------------------------------------------------
# Compute Euclidean distance to clone centroid
# --------------------------------------------------------------
df_dist <- df_dist %>%
  dplyr::add_count(Clone, name = "n_cells") %>%  # number of cells per clone
  dplyr::filter(n_cells >= 2) %>%          # retain multicellular clones only
  group_by(Clone) %>%
  dplyr::rowwise() %>%
  dplyr::mutate( dist_to_centroid = sqrt(sum(
      (dplyr::c_across(starts_with("LSI_")) -
       dplyr::c_across(starts_with("cLSI_")))^2
    ))
  ) %>%
  ungroup()

# --------------------------------------------------------------
# Transfer clone expansion status from matched RNA profiles
# --------------------------------------------------------------
df_dist <- df_dist %>%
 left_join(seu@meta.data[, c("BC_RNA", "ExpandedClone")] , by = c("BC_RNA" = "BC_RNA")) %>% distinct()


# --------------------------------------------------------------
# Export table
# --------------------------------------------------------------
df_dist %>% write.table(., paste0(Res_dir, "/H3K4me1DistCentroid.txt"),
                              quote = F, sep = "\t", row.names = F)
```

## H3K27me3 dataset
```{r}
DefaultAssay(seuEpi) <- "bin_50k_h3k27me3"
#Normalize & RD
seuEpi <- RunTFIDF(seuEpi)
seuEpi <- FindTopFeatures(seuEpi)
seuEpi <- RunSVD(seuEpi)
seuEpi <- RunUMAP(seuEpi, reduction = 'lsi',  dims = 2:30)

# Number of LSI dimensions used
N_PC <- 11

# Extract LSI embeddings (excluding the first component)
pcs <- Embeddings(seuEpi,
                  reduction = "lsi")[,2:N_PC] %>% as.data.frame()

# Create cell-level metadata table
dt <- data.frame(Clone =seuEpi@meta.data$LineageBarcode[which(seuEpi@meta.data$mark == "h3k27me3" )],
                 BC_RNA = seuEpi@meta.data$BC_RNA[which(seuEpi@meta.data$mark == "h3k27me3" )],
                       Sample = seuEpi@meta.data$SampleBio[which(seuEpi@meta.data$mark == "h3k27me3" )],
                       Group = seuEpi@meta.data$Group[which(seuEpi@meta.data$mark == "h3k27me3" )]) %>%
  cbind(pcs)

head(dt)


# --------------------------------------------------------------
# Compute clone centroids in LSI space
# --------------------------------------------------------------
centroids <- dt %>%
  dplyr::group_by(Clone) %>%
  dplyr::summarise(across(paste0("LSI_", 2:N_PC), ~ mean(.x, na.rm = TRUE), .names = "c{.col}"))
head(centroids)

# Add centroid coordinates to each cell
df_dist <- dt %>%
  left_join(centroids, by = c("Clone"))

# --------------------------------------------------------------
# Compute Euclidean distance to clone centroid
# --------------------------------------------------------------
df_dist <- df_dist %>%
  dplyr::add_count(Clone, name = "n_cells") %>%  # number of cells per clone
  dplyr::filter(n_cells >= 2) %>%          # retain multicellular clones only
  group_by(Clone) %>%
  dplyr::rowwise() %>%
  dplyr::mutate( dist_to_centroid = sqrt(sum(
      (dplyr::c_across(starts_with("LSI_")) -
       dplyr::c_across(starts_with("cLSI_")))^2
    ))
  ) %>%
  ungroup()

# --------------------------------------------------------------
# Transfer clone expansion status from matched RNA profiles
# --------------------------------------------------------------
df_dist <- df_dist %>%
 left_join(seu@meta.data[, c("BC_RNA", "ExpandedClone")] , by = c("BC_RNA" = "BC_RNA")) %>% distinct()


# --------------------------------------------------------------
# Export table
# --------------------------------------------------------------
df_dist %>% write.table(., paste0(Res_dir, "/H3K27me3DistCentroid.txt"),
                              quote = F, sep = "\t", row.names = F)
```


#Clone polyfunctionality analysis

```{r}
# Load cell-level metadata containing functional program (FP) scores
meta_all_rna <- seurat_rna@meta.data

```

## Cell state assignment

To focus on non-lineage-associated transcriptional programs,  lineage-related programs (MP1: Basal fate and MP2: Alveolar fate) were excluded from downstream analyses.
```{r}
meta <- meta_all_rna
metadata_cells <- meta

# Retain only molecular programs MP3-MP16
ucell_mat_global <- meta[, (colnames(meta) %in% paste0("MP", 3:16))] 

# Harmonize sample identifiers
df <- meta_all_rna
df[is.na(df$SampleBio ), "SampleBio"] <-  df[is.na(df$SampleBio ),"Exp"]
df[df$SampleBio == "8724_NA", "SampleBio"] <-  "8724_AllGl.Pooled"
df[df$SampleBio == "9475_NA", "SampleBio"] <-  "9475_AllGl.Pooled"


# Reformat score matrix (programs in rows, cells in columns)
ucell_mat_global <- ucell_mat_global[ rownames(df), ] %>% t

# Center and scale program activities across the entire dataset
ucell_mat <- ucell_mat_global 
ucell_mat_global <- t(scale(t(ucell_mat_global), scale = T))
ucell_mat_global[is.nan(ucell_mat_global)] <- 0

# -------------------------------------------------------------- 
# Define active molecular programs 
# --------------------------------------------------------------
# For each program, an activation threshold was defined as: 
# mean positive score + 1 standard deviation

nb_sds <- 1

thr_vect <- apply(ucell_mat, 1, function(x) {
    tmp <- x[x > 0]
    tmp <- na.omit(tmp)
    if (length(tmp) == 0) return(0)  # sécurité
    return(mean(tmp) + nb_sds * sd(tmp))
})

# Binarize program activity
ucell_mat_global_binary <- sweep(ucell_mat, 1, thr_vect, "-")

# Cell states were assigned independently within each sample. 
# A program was considered active when its score exceeded the threshold.
{
    list_mat <- list()
    list_attrib <- list()
    list_attrib_max <- list()
    cell_id <- c()

    for (sample_id in unique(df$SampleBio)) {

        # Extract cells from the current sample
        idx <- which(df$SampleBio == sample_id)
        ucell_mat_sub <- ucell_mat_global_binary[, idx, drop = FALSE]

        # --- Handle samples containing a single cell ---
        if (is.null(dim(ucell_mat_sub))) {
            ucell_mat_sub <- matrix(ucell_mat_sub, ncol = 1)
            colnames(ucell_mat_sub) <- colnames(ucell_mat_global_binary)[idx]
            rownames(ucell_mat_sub) <- rownames(ucell_mat_global_binary)
        }

        # Binary attribution of active programs
        thr <- 0
        tmp_attrib <- apply(ucell_mat_sub, 1, function(x) x >= thr)
        list_attrib[[sample_id]] <- tmp_attrib

        cell_id <- c(cell_id, colnames(ucell_mat_sub))

        # save processed state matrix by sample
        list_mat[[sample_id]] <- ucell_mat_global[, idx, drop = FALSE]

       # Assign the dominant program to each cell
        ucell_mat_sub2 <- ucell_mat_global[, idx, drop = FALSE]
          
      tmp_attrib2 <- apply( ucell_mat_sub2,   2,    
                            function(x){
                              if(all(is.na(x))){
                                return(NA_character_)        }

        tmp_max <- max(x, na.rm = TRUE)

        if(tmp_max > thr){
            rownames(ucell_mat_sub2)[which.max(x)][1]
        } else {
            "Unknown"
        }
    }
)

        list_attrib_max[[sample_id]] <- tmp_attrib2
    }

    df_attrib <- do.call(rbind, list_attrib)
    rownames(df_attrib) <- cell_id
    df_attrib <- df_attrib[colnames(ucell_mat_global), ]

    vect_attrib <- do.call(c, list_attrib_max)
    names(vect_attrib) <- cell_id
    vect_attrib <- vect_attrib[colnames(ucell_mat_global)]
}

df <- cbind(df, df_attrib)
df$state_max <- vect_attrib
```
```{r}

colnames(df)[123:ncol(df)] <- paste0("Bin_", colnames(df)[123:ncol(df)])
hallmark_cols <- grep("^Bin_MP", colnames(df)[123:ncol(df)], value = TRUE)

df$umap1 <- Embeddings(seurat_rna , "UMAP_spectral_rna_umi_harmony")[, 1]
df$umap2 <- Embeddings(seurat_rna , "UMAP_spectral_rna_umi_harmony")[, 2]
```




```{r}
# -------------------------------------------------------------- 
# Functional annotation of programs 
# --------------------------------------------------------------

# Molecular programs were grouped into biologically interpretable 
# functional categories.
df <- df %>% 
  mutate(state_max_relev = case_when(state_max == "MP1" ~ "Basal fate",
                                     state_max == "MP2" ~ "Alveolar fate",
                                     state_max == "MP3" ~ "AP1 activation",
                                     state_max == "MP4" ~ "Angiogenic remodeling",
                                     state_max == "MP7" ~ "Inflammation, cytokines",
                                     state_max == "MP9" ~ "Focal adhesion",
                                     state_max == "MP10" ~ "Prostaglandin synthesis",
                                     state_max == "MP12" ~ "OxPhos",
                                     state_max == "MP14" ~ "Genome maintenance",
                                     state_max == "MP15" ~ "Lipid metabolism",
                                     state_max == "MP16" ~ "EMT",
                                     TRUE ~ state_max))



# Compute clone size and relative abundance within each sample
df_Clone <- df %>%
  as.data.frame() %>%
  dplyr::filter(!is.na(Barcode)) %>%                              #  retain barcoded cells only
  dplyr::group_by(SampleBio) %>%
  dplyr::mutate(TotalCells = n()) %>%                             # total number of barcoded cells per sample
  dplyr::group_by(SampleBio, Barcode) %>%
  dplyr::mutate(CloneSize = n()) %>%                              # number of cells per clone
  dplyr::ungroup() %>%
  dplyr::mutate(PercentClone = 100 * CloneSize / TotalCells) %>%       # normalized clone abundance
  mutate(BCM_PercentClone_MonoGl = case_when(SampleBio %in% 
                                               c("8964_G4", "8718_D2-3", "8415_G2-3") ~ PercentClone,
                                             TRUE ~ NA)) %>%
  mutate(ExpandedClone = case_when(SampleBio %in% c("8964_G4", "8718_D2-3", "8415_G2-3") & 
                                     PercentClone >= 10 ~"Expanded",
                                   SampleBio %in% c("8964_G4", "8718_D2-3", "8415_G2-3") & 
                                     PercentClone < 2 ~"Small",
                                   SampleBio %in% c("8964_G4", "8718_D2-3", "8415_G2-3") & 
                                     PercentClone < 10 &
                                     PercentClone >= 2 ~"Int",
                                   TRUE ~ NA))

# --------------------------------------------------------------
# Export table
# --------------------------------------------------------------
df %>% 
  write.table(., paste0(Res_dir, "/FunctionalProgramAssignment_all.txt"),
                              quote = F, sep = "\t", row.names = T)

df_Clone %>% 
  write.table(., paste0(Res_dir, "/FunctionalProgramAssignment.txt"),
                              quote = F, sep = "\t", row.names = F)
```

## NMF Annotations

```{r}
r <- FindAllMarkers(seurat_rna, group.by = "state_max_relev")


```
```{r}
r %>% group_by(cluster) %>% 
  filter(cluster == "Stress survival" & avg_log2FC > 0) %>% #Stress survival  Genome maintenance Mitochondrial-stress response
  arrange(p_val_adj) %>% 
  select(gene) %>% head(100) %>% View
```


```{r}
anno_FP <-  read.delim(paste0(Input_Dir, "/FunctionalProgramAssignment_all.txt"))
head(anno_FP)
```


FP3 - PID_AP1_PATHWAY ; 
      BIOCARTA_FOSB_PATHWAY 
        *AP1*
FP4 - WP_VEGFAVEGFR2_SIGNALING_PATHWAY ; 
      ABE_VEGFA_TARGETS 
        *Angiogenesis*
FP7 - BIOCARTA_MSP_PATHWAY (Macrophage stimulatiing prot) ;
      WP_CYTOKINES_AND_INFLAMMATORY_RESPONSE
      KEGG_CYTOKINE_CYTOKINE_RECEPTOR_INTERACTION
        *Inflammation, cytokines*
FP9 - WP_FOCAL_ADHESION ;
      KEGG_FOCAL_ADHESION 
        *focal Adhesion / lncRNA*
FP10 - WP_PROSTAGLANDIN_SYNTHESIS_AND_REGULATION
        *psotaglandin synthesis*

FP12 - KEGG_OXIDATIVE_PHOSPHORYLATION
       WP_OXIDATIVE_PHOSPHORYLATION
       WP_ELECTRON_TRANSPORT_CHAIN_OXPHOS_SYSTEM_IN_MITOCHONDRIA
       HALLMARK_OXIDATIVE_PHOSPHORYLATION
         *OxPhos*

FP14 - WP_DNA_MISMATCH_REPAIR
       KAUFFMANN_DNA_REPLICATION_GENES
       KEGG_DNA_REPLICATION
       WP_DNA_REPLICATION
       REACTOME_DNA_REPAIR
       KEGG_RNA_DEGRADATION
        *Genome maintenance*
        
FP15 - KEGG_GLYCEROLIPID_METABOLISM
       REACTOME_METABOLISM_OF_LIPIDS
       WP_MAMMARY_GLAND_DEVELOPMENT_PATHWAY_PREGNANCY_AND_LACTATION_STAGE_3_OF_4
       *lipid metabolism*
       
FP16 - GOTZMANN_EPITHELIAL_TO_MESENCHYMAL_TRANSITION_UP
       HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION
        *EMT*
        
```{r}
library(Seurat)
library(UCell)
library(msigdbr)
library(dplyr)
library(ggplot2)

msig <- msigdbr(species = "Mus musculus")



FP_pathways <- list(

FP3 = c(
"PID_AP1_PATHWAY" #,"BIOCARTA_FOSB_PATHWAY"
),

FP4 = c(
"WP_VEGFAVEGFR2_SIGNALING_PATHWAY" #,"ABE_VEGFA_TARGETS"
),

FP7 = c(
#"BIOCARTA_MSP_PATHWAY",
"WP_CYTOKINES_AND_INFLAMMATORY_RESPONSE" #,"KEGG_CYTOKINE_CYTOKINE_RECEPTOR_INTERACTION"
),

FP9 = c(
"WP_FOCAL_ADHESION"#,"KEGG_FOCAL_ADHESION"
),

FP10 = c(
"WP_PROSTAGLANDIN_SYNTHESIS_AND_REGULATION"
),

FP12 = c(
"KEGG_OXIDATIVE_PHOSPHORYLATION"#,"WP_OXIDATIVE_PHOSPHORYLATION","WP_ELECTRON_TRANSPORT_CHAIN_OXPHOS_SYSTEM_IN_MITOCHONDRIA","HALLMARK_OXIDATIVE_PHOSPHORYLATION"
),

FP14 = c(
"WP_DNA_MISMATCH_REPAIR",
#"KAUFFMANN_DNA_REPLICATION_GENES",
#"KEGG_DNA_REPLICATION",
"WP_DNA_REPLICATION" #,"REACTOME_DNA_REPAIR","KEGG_RNA_DEGRADATION"
),

FP15 = c(
#"KEGG_GLYCEROLIPID_METABOLISM"#,
  "REACTOME_METABOLISM_OF_LIPIDS"#,"WP_MAMMARY_GLAND_DEVELOPMENT_PATHWAY_PREGNANCY_AND_LACTATION_STAGE_3_OF_4"
),

FP16 = c(
#"GOTZMANN_EPITHELIAL_TO_MESENCHYMAL_TRANSITION_UP",
"HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION"
)

)

#consruct gene sets 
gene_sets <- list()

for(fp in names(FP_pathways)){

    for(gs in FP_pathways[[fp]]){

        genes <- msig %>%
            filter(gs_name == gs) %>%
            pull(gene_symbol) %>%
            unique()

        genes <- intersect(genes, rownames(seurat_rna))

        if(length(genes) >= 10){

            gene_sets[[gs]] <- genes

        }

    }

}

#compute UCell score
sic <- AddModuleScore_UCell(
    seurat_rna,
    features = gene_sets
)
#Make table
scores <- FetchData(
    sic,
    vars = paste0(names(gene_sets),"_UCell")
)
#add state
scores$cell <- rownames(scores)
scores$state_max_relev <- anno_FP$state_max_relev

#mean by state
heat <- scores %>%
  filter( !is.na(state_max_relev)) %>%
group_by(state_max_relev) %>%
summarise(across(ends_with("_UCell"), mean))

mat <- heat %>%
column_to_rownames("state_max_relev") %>%
as.matrix()


mat <- t(scale((mat)))

my.cols <- colorRampPalette(c(
    "#3B7EA1",   # bleu doux
    "#F8F4E8",   # ivoire
    "#A6365D"    # bordeaux
))(101)

pheatmap(
  mat[, c("AP1 activation",          
        "Angiogenic remodeling",
        "Inflammation, cytokines" ,
         "Focal adhesion"  ,
         "Prostaglandin synthesis",
         "OxPhos",
         "Genome maintenance",
        "Lipid metabolism",
        "EMT")],
  cluster_rows=F,
  cluster_cols=F,
    color = my.cols,
    breaks = seq(-2, 2, length.out = 102),
border_color=NA
)
```
```{r}

grep("ox", unique(msig_wp$gs_name),
     ignore.case = TRUE, value = TRUE)


VEGF_sig <- msig_wp %>%
    filter(gs_name == "HALLMARK_OXIDATIVE_PHOSPHORYLATION") %>%
    pull(gene_symbol) %>%
    str_to_title() %>%
    unique() %>%
  intersect(.,  rownames(seurat_rna) )


#compute module
seurat_rna <- AddModuleScore_UCell(
    seurat_rna,
    features = list(WP_VEGFA = VEGF_sig)
)

#Add score
anno_FP$WP_VEGFA <-seurat_rna$WP_VEGFA_UCell[
    match(rownames(anno_FP), colnames(seurat_rna))
]
#make graph
ggplot( anno_FP,  
        aes(x = state_max_relev, y = scale(WP_VEGFA), fill = state_max_relev)) +
  geom_violin(alpha=0.5)+
  geom_boxplot(alpha=0.5, width = 0.2, fill = "white") +
  geom_hline(yintercept = 0)+
  theme_classic() +
  scale_fill_manual(values= state_colors )+
  scale_x_discrete(limits = state_colors %>% names) +
  labs(
    x="Maximum functional programme relevance")
```

```{r}
msig_h <- msigdbr(
  species = "Mus musculus",
  category = "H"
)
```
```{r}
grep("Angio", unique(msig_h$gs_name),
     ignore.case = TRUE, value = TRUE)

test_sig <- msig_h %>%
    filter(gs_name == "HALLMARK_ANGIOGENESIS") %>%
    pull(gene_symbol) %>%
    unique()

#check which genes are presents
test_sig <- intersect(
    test_sig,
    rownames(seurat_rna)
)


#compute module
seurat_rna <- AddModuleScore_UCell(
    seurat_rna,
    features = list(test = test_sig)
)

#Add score
anno_FP$test <-seurat_rna$test_UCell[
    match(rownames(anno_FP), colnames(seurat_rna))
]
#make graph
ggplot( anno_FP,  aes(x = state_max_relev, y = scale(test), fill = state_max_relev)) +
  geom_violin(alpha=0.5)+
  geom_boxplot(alpha=0.5, width = 0.2, fill = "white") +
  geom_hline(yintercept = 0)+
  theme_classic() +
  scale_fill_manual(values= state_colors )+
  scale_x_discrete(limits = state_colors %>% names) +
  labs(
    x="Maximum functional programme relevance",
    y="test (UCell)"
)
```




## Quantification of clone poly-functionality 

Cells from monoclonal samples were retained and clones were stratified into Small, Intermediate, or Expanded categories according to sample-specific clone size distributions.
```{r}

df_sub <- df  %>%
  dplyr::filter(
    SampleBio %in% c("8964_G4", "8718_D2-3", "8415_G2-3"),
    !is.na(Barcode)  )%>%
  left_join(seu@meta.data[, c("Barcode", "ExpandedClone")], by = "Barcode") %>%
  distinct

```

Clone-level functional diversity was quantified using: 
(i) MP entropy: 
Shannon entropy calculated from the distribution of active molecular programs across all cells within a clone. 

(ii) Number of active MPs: 
Total number of distinct molecular programs detected within a clone, used as a measure of clone poly-functionality.

```{r}
library(entropy)
mp_cols <- grep("^Bin_MP", colnames(df), value = TRUE)


clone_stats <- df_sub %>%
  group_by(Barcode) %>%
  summarise(
    clone_size = n(),
    mean_pseudotime = mean(pseudotime, na.rm = TRUE),

    MP_entropy = {
      mp_counts <- colSums( across(all_of(mp_cols), ~ . == 1), na.rm = TRUE)
      mp_counts <- mp_counts[mp_counts > 0]
      entropy(mp_counts / sum(mp_counts)) #Shannon's entropy
    },

    n_active_MP = {
      mp_counts <- colSums(across(all_of(mp_cols), ~ . == 1), na.rm = TRUE )
      sum(mp_counts > 0)
    },    .groups = "drop"  ) %>%
  
  left_join(df_sub[, c("Barcode","SampleBio","ExpandedClone")] %>% 
              distinct(), by = "Barcode")


# --------------------------------------------------------------
# Export table
# --------------------------------------------------------------
clone_stats %>% write.table(., paste0(Res_dir, "/FP_byClone.txt"),
                              quote = F, sep = "\t", row.names = F)
```


## Integration with barcode generation probability (Pgen)
Clone-level poly-functionality metrics were integrated with Pgen values estimated using the IGOR pipeline.

```{r}
clone_stats <- clone_stats %>%
  left_join(seurat_rna@meta.data[, c("Barcode", "Pgen")] , by = c("Barcode" = "Barcode")) %>%
  distinct()
head(clone_stats)
```
To evaluate the influence of barcode generation probability on clone poly-functionality, progressively more stringent Pgen thresholds were applied (10^0 to 10^-7).

```{r}
library(purrr)

res_pgen_MP <- map_dfr(thresholds, function(thr){

  tmp <- clone_stats %>%
    filter(Pgen <= thr)

  tmp %>%
    group_by(ExpandedClone) %>%
    summarise(
      threshold = thr,
      n_clones = n(),
      mean_MP = mean(n_active_MP, na.rm = TRUE),
      median_MP = median(n_active_MP, na.rm = TRUE),
      sem = sd(n_active_MP, na.rm = TRUE)/sqrt(n()),
      .groups = "drop"
    )
})
# --------------------------------------------------------------
# For each threshold, AOV
# --------------------------------------------------------------
stats <- purrr::map_dfr(thresholds[1:12], function(thr){
  dat <- clone_stats %>%
    filter(  Pgen <= thr)

  fit <- aov(n_active_MP  ~ ExpandedClone, data = dat)

  broom::tidy(fit) %>%
    filter(term == "ExpandedClone") %>%
    transmute(
      Thr = thr,
      logThr = -log10(thr),
      p = p.value
    )
})
stats
# --------------------------------------------------------------
# Export tableS
# --------------------------------------------------------------
res_pgen_MP %>% write.table(., paste0(Res_dir, "/FP_byClone_Pgen.txt"),
                              quote = F, sep = "\t", row.names = F)

```



# Lineage switch score : 
```{r}
# build the single-cell metadata table
df <-  seurat_rna@meta.data %>%
  dplyr::filter(!Group %in% c("Tumor", "CreN")) %>%  # remove tumor and control samples
  dplyr::select(WhichGland, Barcode, annot_rna_joint_alv, Group, Pgen) %>%
  dplyr::filter(!WhichGland %in% c("AllGl.Pooled") ) %>%
  dplyr::filter(!is.na(Barcode), !is.na(annot_rna_joint_alv))  

# Metadata at clone ( lineage barcode) level
metadata_barcode <- seurat_rna@meta.data %>%
  dplyr::filter(!Group %in% c("Tumor", "CreN")) %>%
  dplyr::select(WhichGland, Barcode, annot_rna_joint_alv, Group, Pgen) %>%
  dplyr::filter(!WhichGland %in% c("AllGl.Pooled") ) %>%
  dplyr::filter(!is.na(Barcode), !is.na(annot_rna_joint_alv))  %>%
  dplyr::select( c("Group",  "Barcode", "Pgen")) %>%
  dplyr::filter(!duplicated(Barcode))

# ------------------------------------------------------------
# Compute clone-level lineage switching metrics
# ------------------------------------------------------------

df_clone_stats <-df %>% 
  group_by(Barcode) %>%
  summarise(
    n_total = n(),# total number of cells per clone
    
     # number of cells in each lineage state
    n_LP = sum(annot_rna_joint_alv == "LP", na.rm = TRUE),
    n_Basal = sum(annot_rna_joint_alv == "Basal", na.rm = TRUE),
    n_Alv = sum(annot_rna_joint_alv == "Alv", na.rm = TRUE),
    
      # number of cells not in LP state (i.e., switched states)
    n_other = sum(annot_rna_joint_alv != "LP" & !is.na(annot_rna_joint_alv)),
    
     # lineage switching score: fraction of non-LP cells per clone
    switch_score = n_other / n_total  ) %>%
  ungroup() %>%
  dplyr::select(n_total  ,n_LP, n_Basal, n_Alv,n_other ,switch_score, Barcode   ) %>%
  left_join(metadata_barcode, by = "Barcode") %>% # add clone-level metadata (Group, Pgen)
  dplyr::filter(!duplicated(Barcode)) # ensure unique clones


# --------------------------------------------------------------
# Export table
# --------------------------------------------------------------
df_clone_stats %>% write.table(., paste0(Res_dir, "/CellFateSwitch.txt"),
                              quote = F, sep = "\t", row.names = F)
```

## Integration with barcode generation probability (Pgen)
```{r}
# ------------------------------------------------------------
# Define Pgen thresholds for filtering clones
# ------------------------------------------------------------
thresholds <- 10^seq(0, -7, by = -0.25)
thresholds
```


```{r}
# ------------------------------------------------------------
# Evaluate switch score as a function of Pgen threshold
# ------------------------------------------------------------


res_pgen <- map_dfr(thresholds, function(thr){

   # keep only clones with Pgen below threshold (more "likely independent" barcodes)
  df_filt <- df_clone_stats %>%
    filter(Pgen <= thr)

  #  compute switch statistics at group level
  summary_group <- df_filt %>%
    group_by(Group) %>%
    summarise(
      n_clones = n(),
      
      # central tendency of switch score
      mean_switch = mean(switch_score, na.rm = TRUE),
      median_switch = median(switch_score, na.rm = TRUE),
      prop_switch = mean(switch_score > 0),
      .groups = "drop"
    )

   # extract early and late groups separately
  early <- summary_group %>%
    filter(Group == "CreP_EARLY")

  late <- summary_group %>%
    filter(Group == "CreP_LATE")

   # assemble results for each threshold
  tibble(
    threshold = thr,
    n_Early = early$n_clones,
    n_Late = late$n_clones,

    mean_switch_Early = early$mean_switch,
    mean_switch_Late = late$mean_switch,

    median_switch_Early = early$median_switch,
    median_switch_Late = late$median_switch,

    prop_switch_Early = early$prop_switch,
    prop_switch_Late = late$prop_switch,

      # ratios late vs early
    ratio_mean = late$mean_switch / early$mean_switch,
    ratio_median = late$median_switch / early$median_switch,
    ratio_prop = late$prop_switch / early$prop_switch
  )
})


# --------------------------------------------------------------
# For each threshold, AOV
# --------------------------------------------------------------
stats <- purrr::map_dfr(thresholds[1:27], function(thr){
  dat <- df_clone_stats %>%
    filter(  Pgen <= thr)

  fit <- aov(switch_score  ~ Group, data = dat)

  broom::tidy(fit) %>%
    filter(term == "Group") %>%
    transmute(
      Thr = thr,
      logThr = -log10(thr),
      p = p.value
    )
})
stats

# --------------------------------------------------------------
# Export table
# --------------------------------------------------------------
res_pgen %>% write.table(., paste0(Res_dir, "/CellFateSwitch_Pgen.txt"),
                              quote = F, sep = "\t", row.names = F)
```

# Expanded clone markers – DEG analysis (Fig 3 / related)

```{r}
# ------------------------------------------------------------
# Build metadata table for clones
# ------------------------------------------------------------
dt <- data.frame(fractionclone = seu$PercentClone,
                 value = seu$CloneSize,
                 sample = seu$SampleBio.x,
                 fill = seu$Group,
                 bcm = seu$Barcode)

# Subset: selected Gland entierly screened & clones >= 5 cells
mini_dt <- dt %>% 
  subset(sample %in% c("8964_G4", "8718_D2-3", "8415_G2-3")) %>%
  subset(value >= 5) 

table(mini_dt$bcm, mini_dt$sample)
```

```{r}
# ------------------------------------------------------------
# Define expanded clone identities in Seurat object
# ------------------------------------------------------------
# Rule:
# - True expanded clones (size >= 5) keep their barcode
# - Small clones (<2% frequency) are grouped as "Small_<sample>"
seu@meta.data <- seu@meta.data %>%
  mutate(ExpandedClone_nameSup5 = case_when(SampleBio.x %in% c("8964_G4", "8718_D2-3", "8415_G2-3") & 
                                     CloneSize >= 5 ~ Barcode,
                                   SampleBio.x %in% c("8964_G4", "8718_D2-3", "8415_G2-3") & 
                                     PercentClone < 2 ~paste0("Small", "_", SampleBio.x),
                                   TRUE ~ NA))

# ------------------------------------------------------------
# Build table of clone × sample with clone sizes
# ------------------------------------------------------------
SampleCloneSup5 <- ClonesSup5 <- table(mini_dt$bcm, mini_dt$sample) %>% 
  melt %>% 
  filter(!value <5)

colnames(SampleCloneSup5) <- c("Barcode", "Sample", "CloneSize")

# ------------------------------------------------------------
# Initialize DEG results metrix
# ------------------------------------------------------------
DEGs_sup5 <- data.frame(p_val= NULL, avg_log2FC=NULL, 
                        pct.1 = NULL, pct.2 = NULL, p_val_adj = NULL,
                        clone = NULL, sampleOrigine = NULL, clonesize = NULL, gene = NULL
                        )
# ============================================================
# Loop over expanded clones to compute DEGs
# ============================================================
for (i in 1:nrow(SampleCloneSup5)) {
  
  clone <- SampleCloneSup5$Barcode[i]
  sampleOrigine <- SampleCloneSup5$Sample[i]
  clonesize <- SampleCloneSup5$CloneSize[i]
  
  # Differential expression: clone vs small clones
  
  res <- FindMarkers(seu ,
                         group.by = "ExpandedClone_nameSup5", 
                         ident.1 = clone ,
                         ident.2 = c("Small_8415_G2-3",  "Small_8718_D2-3", "Small_8964_G4"),
                         test.use = 'wilcox',only.pos = F,
  recorrect_umi = FALSE)
  
  
  # Keep significant DEGs only
  
  signature_Clone <- subset(res, p_val_adj < 0.05)
  annot <- data.frame(clone = rep(clone, nrow(signature_Clone)),
                      sampleOrigine = rep(sampleOrigine, nrow(signature_Clone)),
                      clonesize = rep(clonesize, nrow(signature_Clone)))
  dt_int <- cbind(signature_Clone, annot)
  dt_int$gene = rownames(signature_Clone)
  rownames(dt_int) <- paste0(dt_int$gene, dt_int$sampleOrigine)
  
  DEGs_sup5 <- rbind(DEGs_sup5, dt_int)
  
  # Volcano plot
  DEG <- res

  DEG$signif <- DEG$p_val_adj < 0.05
  DEG$FC <- rep(NA, nrow(DEG))
  DEG$FC[which(DEG$avg_log2FC > 0 & DEG$signif ==TRUE)] = "UP"
  DEG$FC[which(DEG$avg_log2FC < 0 & DEG$signif ==TRUE)] = "DW"

  DEG$gene <- rownames(res)

  sub <- sample(1:nrow(DEG), size = 100, replace = F) 



      gg1 <- ggplot(DEG[sub,],
          aes(x = (avg_log2FC), y = -log10(p_val_adj), 
              fill = FC , size = signif, alpha = signif) )+
        geom_point(shape = 21, colour = "black")+ 
        geom_point(data = subset(DEG, signif == TRUE), 
             shape = 21, colour = "black") +
        geom_label_repel(data = subset(DEG, signif == TRUE), 
                   aes(label = gene),  force = 2,  max.overlaps =100) +
  
        geom_hline(yintercept = -log10(0.05), linetype = "dashed") + 
        geom_vline(xintercept = c(log2(0.5), log2(2)),linetype = "dashed") +
        
        ggtitle(paste0(clone,"\n", sampleOrigine, "_n=",clonesize))+
  
        scale_fill_manual(values = c("#d81159")) + 
        scale_size_manual(values = c(2,2,1)) + 
        scale_alpha_manual(values = c(0.5, 1,1)) +
        scale_x_continuous(breaks = c(seq(-10, 10, 2)),       
                     limits = c(-10, 10))  +
        scale_y_continuous(limits = c(0,10))+
        theme_bw() +
        theme(panel.border = element_rect(colour = "black", fill = NA, size= 0.5),    
              panel.grid.minor = element_blank(),
              panel.grid.major = element_blank())

  gg1 %>% print()

}
```

```{r}
# ------------------------------------------------------------
# Summary checks
# ------------------------------------------------------------
DEGs_sup5 %>% head()

# Most frequent DE genes
DEGs_sup5$gene %>% table %>% sort() %>% tail(., 40)
```





```{r}
# ============================================================
# Build binary DEG matrix (presence/absence per clone)
# ============================================================
mat_deg <- DEGs_sup5 %>%
  dplyr::filter(avg_log2FC > 0) %>%  # keep only UP-regulated genes
  mutate(DEG_present = 1) %>%
  dplyr::select(gene, clone, DEG_present) %>%
  distinct() %>%
  pivot_wider(names_from = clone, values_from = DEG_present, values_fill = 0)


# Save full DEG table and matrix
write.csv(mat_deg, paste0(Res_dir,  "/mat_degSup5.csv"))
write.csv(DEGs_sup5, paste0(Res_dir,  "/DEGs_sup5.csv"))

# ------------------------------------------------------------
# Identify genes shared across multiple expanded clones
# ------------------------------------------------------------
genes_multi <- mat_deg$gene[rowSums(mat_deg[2:20]) > 1]
mat_multi <- mat_deg %>% 
  filter(gene %in% genes_multi)

# Save
write.csv(mat_multi, paste0(Res_dir,  "/mat_degSup5_UP_multi.csv"))
```







