---
title: "R Notebook"
output: html_notebook
---


# Introduction

```{r}
# -------------------------------------------------------------- 
# Libraries
# --------------------------------------------------------------
library(Seurat) ; library(Signac)
library(Matrix) ; library(uwot)

library(msigdbr) ; library(clusterProfiler)

library(dplyr)

library(ggplot2)
library(ggforce)
library(ggrepel)
library(gridExtra)

# -------------------------------------------------------------- 
# Color definition
# --------------------------------------------------------------
Bin_colors <- c(
  "Bin_MP3"  = "#fdea44",  
  "Bin_MP4"  = "#a7d066", 
  "Bin_MP7"  = "#e9407b",  
  "Bin_MP9"  = "#d3e44c",  
  "Bin_MP10" = "#b16680",  
  "Bin_MP12" = "#973356",  
  "Bin_MP14" = "#3f0418",  
  "Bin_MP15" = "#f9e8cc",  
  "Bin_MP16" = "#6c9d1e",  
  "No_MP"    = "grey95" 
)

state_colors <- c(
  "Basal fate"                     = "#236ec4", 
  "Alveolar fate"                  = "#F7DCAD",  
  "AP1 activation"                 = "#fdea44",  
  "Angiogenic remodeling"          = "#a7d066", 
  "Inflammatory cytokines"         = "#e9407b",  
  "Mechanostransduction"           = "#d3e44c",  
  "Stress survival"                = "#b16680",  
  "Mitochondrial-stress response"  = "#973356",  
  "Genome maintenance"             = "#3f0418",  
  "Primed Alveolar"                = "#f9e8cc",  
  "EMT / plasticity"               = "#6c9d1e",  
  "Unknown"                        = "grey95"   
)



cols_fate <- c("basal"  = "#236ec4", 
               "LP"     = "#a01858",      
               "HS"     = "#f9c40c",  
               "Alv"    = "#F7DCAD")


cols_Group <- c("CreP_EARLY" = "#C0D69F",  
                "CreP_LATE"  = "#61980E",
                "CreN_LP"    = "#a01858")

cols_MonoGland <- c("8415_G2-3" = "#375708",  
                    "8718_D2-3"     = "#61980E",
                    "8964_G4"       = "#C0D69F")
```


```{r}
Output_Dir <- "C:/Users/Marthe Laisne/OneDrive - INSTITUT CURIE/m_laisne/Publications/2026_Laisne/Scripts/Output/sc_MutantvsWT"

# Load sc RNA & Multiome

seurat_rna <- qs::qread(file.path("C:/Users/Marthe Laisne/OneDrive - INSTITUT CURIE/m_laisne/Publications/2026_Laisne/seurat_rna_step7_new.qs"))

seurat <- qs::qread(file.path("C:/Users/Marthe Laisne/OneDrive - INSTITUT CURIE/m_laisne/Publications/2026_Laisne/seurat_step5_new.qs"))

```

# scRNA

## Fig. 3b - Violin - Median cell cell transcriptomic correlation 

```{r}
# ============================================================
# Median cell–cell transcriptomic correlation
# ============================================================

# Extract raw RNA count matrix
expr <- GetAssayData(seurat_rna,assay = "RNA_umi", slot = "counts")

# Retrieve cell metadata
meta <- seurat_rna@meta.data

#Define cell groups
groups_list <- list(
  CreN_LP = rownames(meta %>% filter(Group == "CreN" & annot_rna_joint_alv == "LP")),
  CreP_EARLY = rownames(meta %>% filter(Group == "CreP_EARLY")),
  CreP_LATE = rownames(meta %>% filter(Group == "CreP_LATE"))
)


# ------------------------------------------------------------
# Compute median cell–cell transcriptomic correlation
# ------------------------------------------------------------
calc_cell_corr <- function(cells, expr, label) {

# Subset expression matrix to selected cells
  mat <- expr[, cells, drop = FALSE]
  mat_dense <- as.matrix((mat))
  
  # Remove genes with zero variance
  keep <- apply(mat_dense, 1, sd) > 0
  mat_dense <- mat_dense[keep, ]

  #ship group with fewer than two cells
  if (ncol(mat_dense) < 2) {
    return(NULL) }

   # Compute Pearson correlation matrix between cells
  corr_mat <- cor(mat_dense, method = "pearson")

  # Compute the median correlation for each cell,
  # excluding self-correlation (diagonal)
    med_corr <- apply(corr_mat, 2, function(x) {
    median(x[x < 0.9999], na.rm = TRUE)
  })

  tibble(
    cell = names(med_corr),
    corr_score = med_corr,
    group = label
  )
}

# Compute correlation scores for each group
corr_df_solo <- bind_rows(
  calc_cell_corr(groups_list$CreN_LP, expr, "CreN_LP"),
  calc_cell_corr(groups_list$CreP_EARLY, expr, "CreP_EARLY"),
  calc_cell_corr(groups_list$CreP_LATE, expr, "CreP_LATE")
)


# ------------------------------------------------------------
# Visualize median cell–cell correlation
# ------------------------------------------------------------
ggplot(corr_df_solo, aes(group, corr_score, fill = group)) +
  geom_violin(trim = T, alpha = 1) +
  geom_boxplot(outliers = F, width = 0.2)+
  scale_fill_manual(values = cols_Group)+
  theme_classic() +
  labs(
    x = "",
    y = "Median cell–cell correlation"  ) +
  guides(fill = "none")

```

```{r}
# ------------------------------------------------------------
# Statistical analysis
# One-way ANOVA followed by Tukey's post hoc test
# ------------------------------------------------------------

anova_model <- aov(corr_score ~ group, 
                   data = corr_df_solo)
summary(anova_model)

tukey_res <- TukeyHSD(anova_model)
tukey_res
```


## Definition of discordant or transitory states 
```{r}
# ============================================================
# Definition of discordant or transitory cell states
# ============================================================

# Build annotation table combining lineage group, FACS identity, transcriptomic identity and epigenomic identity
dt <- data.frame(
                 Group_EL =seurat$Group,
                 Mark = seurat$mark,
                 FACS_ID =seurat$facs_phenotype,
                 RNA_ID = seurat$annot_rna_joint_alv,
                 Epi_ID = seurat$condition_k4_detailed)


dt <- dt %>%
  # Complete H3K4me1 annotations using H3K27me3-based annotations
  mutate(Epi_ID = case_when(is.na(Epi_ID) == T & 
                              is.na(seurat$condition_k27_broad)==F ~seurat$condition_k27_detailed,
                            TRUE ~Epi_ID )) %>%
    # Rename control group
  mutate(Group_EL = case_when(Group_EL == "Rosa" ~ "CreN",
                              TRUE ~ Group_EL))


# Simplify & homogeneise epigenomic annotation names
dt$Epi_ID <- gsub(".*_", "", dt$Epi_ID)
dt <- dt %>%
     mutate(Epi_ID = case_when(Epi_ID == "lp" ~ "LP",
                               Epi_ID == "hs" ~ "HS",
                               TRUE ~Epi_ID) ) 

# Remove cells without multiomic annotation
dt <- dt %>%
  filter(!Epi_ID == "noomics")%>%
  dplyr::filter(!is.na(FACS_ID), !is.na(RNA_ID), !is.na(Epi_ID))


head(dt) ; tail(dt)
```



### Fig. 4g > To be replace by Anna's version (FACS vs. RNA | RNA vs. Epi)
```{r}
# ============================================================
# Alluvial plot function
# ============================================================


# Define common ordering of cell identities across modalities
levels_x <- c("basal", "luminal", "Alv",  "LP","HS")
alpha <- 0.7

make_alluvial <- function(df, title = "") {
  
    # Compute frequency of each annotation combination
  df2 <- df %>%
    group_by(RNA_ID, FACS_ID, Epi_ID) %>%
    summarise(n = n(), .groups = "drop") %>%
    mutate(freq = n / sum(n) * 100) %>%
    dplyr::select( FACS_ID,  RNA_ID, Epi_ID, freq) %>%
    
      # Convert data into ggforce parallel sets format
    gather_set_data(1:3) %>%       
    arrange(x,  FACS_ID, RNA_ID, Epi_ID) %>%
    
     # Keep consistent ordering of cell states
    mutate(
      FACS_ID = factor(FACS_ID, levels = levels_x),
      RNA_ID  = factor(RNA_ID,  levels = levels_x),
      Epi_ID  = factor(Epi_ID,  levels = levels_x),
      y       = factor(y,       levels = levels_x)
    )
  
    # Generate alluvial representation
  ggplot(df2, aes(x = x, id = id, split = y, value = freq)) +
    geom_parallel_sets(aes(fill = RNA_ID), alpha = alpha, axis.width = 0.3) +
    geom_parallel_sets_axes(axis.width = 0.25, fill = "gray95",
                            color = "gray80", size = 0.3) +
    geom_parallel_sets_labels(colour = 'gray35', size = 4, angle = 0,
                              fontface = "bold") +
    theme_minimal() +
    labs(title = title) +
    theme(
      legend.position = "none",
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text.y = element_blank(),
      axis.text.x = element_text(size = 18, face = "bold"),
      axis.title.x = element_blank(),
      plot.title = element_text(size = 20, face = "bold")
    )
}



# ============================================================
# Generate alluvial plots for each biological condition
# ============================================================

# Control cells with luminal progenitor RNA identity
df_CreN_LP <- dt %>%
  dplyr::filter(Group_EL == "CreN", RNA_ID == "LP")

ggCreN <- make_alluvial(df_CreN_LP, title = "CreN — LP (RNA subset)")
ggCreN


# Early mutant cells
df_CreP_Early <- dt %>%
  dplyr::filter(Group_EL == "CreP_EARLY")

ggCreP_e <- make_alluvial(df_CreP_Early, title = "CreP — Early")
ggCreP_e


# Late mutant cells
df_CreP_Late <- dt %>%
  dplyr::filter(Group_EL == "CreP_LATE")

ggCreP_l <- make_alluvial(df_CreP_Late, title = "CreP — Late")
ggCreP_l
```


```{r}
# ============================================================
#  Pie chart definition of concordant, discordant and specific cell states
# ============================================================

## Function to generate pie charts from FACS, RNA and epigenomic annotations

make_pie_chart <- function(df, title = NULL, colors) {
  
  # Assign pie categories according to user-defined rules
  df <- df %>%
    mutate(pie_label = case_when(RNA_ID == "LP" & FACS_ID == "luminal" &  Epi_ID == "LP" ~ "Concordant",
                               RNA_ID == "basal" & FACS_ID == "basal" &  Epi_ID == "basal" ~ "basal" ,
                               RNA_ID == "Alv" &  FACS_ID == "luminal" &  Epi_ID == "LP" ~ "Alv" ,
                               RNA_ID == "basal" & (FACS_ID != "basal" |  Epi_ID != "basal" )~ "Discordant" ,
                               RNA_ID == "LP" & (FACS_ID != "luminal" |  Epi_ID != "LP" )~ "Discordant" ,
                               RNA_ID == "Alv" & (FACS_ID != "luminal" |  Epi_ID != "LP" )~ "Discordant" ,
                               TRUE ~ "Other"))
  
  # Compute fraction of cells in each category
  df_pie <- df %>%
    count(pie_label) %>%
    mutate(freq = n / sum(n),
           label = scales::percent(freq))
  
  # Generate donut plot
  p <- ggplot(df_pie, aes(x = 2, y = freq, fill = pie_label)) +
    geom_col(color = "white", width = 1) +
    geom_text(aes(label = label),
              position = position_stack(vjust = 0.5),
              size = 5,
              color = "white") +
    coord_polar(theta = "y") +
    xlim(0.5, 2.5) +
    scale_fill_manual(values = colors) +
    theme_void() +
    theme(
      legend.title = element_blank(),
      plot.title = element_text(size = 18, face = "bold")
    ) +
    labs(title = title)
  
  return(p)
}


# Define pie colors
pie_colors <- c(
  "Concordant" = "#a01858",
  "basal" = "#236ec4",
  "Alv" = "#F7DCAD",
  "HS" = "#f9c40c",
  "Discordant" = "black",
  "Other" = "grey80"
)

```

```{r}
p_CreN <- make_pie_chart(
  df_CreN_LP,
  title = "CreN LP",
  colors = pie_colors)

p_CreP_e <- make_pie_chart(
  df_CreP_Early,
  title = "CreP Early",
  colors = pie_colors)

p_CreP_l <- make_pie_chart(
  df_CreP_Late,
  title = "CreP Late",
  colors = pie_colors)


p_CreN
p_CreP_e
p_CreP_l
```










## Markers genes Basal Mutant 
Comparison of mutant and wild-type basal populations, restricted to cellsshowing concordant annotation between FACS phenotype and RNA-based cell identity


```{r}
# ============================================================
# Define coherent and discordant FACS/RNA annotations
# ============================================================

# Cells are considered coherent when FACS and transcriptomic annotations identify the same lineage compartment

seurat_rna@meta.data <- seurat_rna@meta.data %>%
  mutate(Annotation_Coherent_FACS_RNA = case_when(
    facs_phenotype == "basal" & annot_rna_joint_alv == "basal" ~ "CoherentFACSRNA_basal",
    facs_phenotype != "basal" & annot_rna_joint_alv != "basal" ~ "CoherentFACSRNA_Luminal",
    is.na(facs_phenotype) ~ paste0("CoherentFACSRNA_", annot_rna_joint_alv),
    TRUE ~ paste0("DicordandFACS_", facs_phenotype, "_RNA_", annot_rna_joint_alv))) 

# Rename control group for consistency 
seurat_rna@meta.data <- seurat_rna@meta.data %>%
  mutate(Group = case_when(Group == "Rosa" ~ "CreN",
                           TRUE ~ Group))

# Create combined identity for differential expression analysis comparing mutant and control populations with matched annotations
seurat_rna@meta.data$ID_DEGs <- paste0(
  seurat_rna@meta.data$Group, "_",
  seurat_rna@meta.data$Annotation_Coherent_FACS_RNA)

# Check distribution of coherent/discordant populations across groups
table(seurat_rna@meta.data $ID_DEGs)
```
```{r}
# Visualize coherent and discordant populations on the integrated RNA UMAP
DimPlot( seurat_rna , group.by = "Annotation_Coherent_FACS_RNA", reduction = "UMAP_spectral_rna_umi_harmony")

```

```{r}
# ============================================================
# Differential expression analysis between mutant and wild-type basal cells
# ============================================================
# restricted to cells with concordant FACS and RNA basal annotation

res <- FindMarkers(seurat_rna ,
                         group.by = "ID_DEGs", 
                         ident.1 =c("CreP_LATE_CoherentFACSRNA_basal", 
                                    "CreP_EARLY_CoherentFACSRNA_basal"), #104+12
                         ident.2 = c("CreN_CoherentFACSRNA_basal"), #636
                         test.use = 'wilcox_limma',
                         only.pos = F,
  recorrect_umi = FALSE)

# Select significant differentially expressed genes
signature <- subset(res, p_val_adj < 0.05)
signature
```





### Fig. 5g - Volcano
```{r}
# ============================================================
# Functional annotations of DEGs
# ============================================================

# Load published mammary epithelial cell signatures
Signature_ID_CreN_10x <- read.delim("C:/Users/Marthe Laisne/Desktop/BI_references_signatures_annotations/Mammarygland_Signatures_Custom/SignatureCreN_scRNA10x_Epithelial_poles.txt")

# Add alveolar signature from Chen et al., 2021
LAlv_Chen2021  <- read.delim("C:/Users/Marthe Laisne/Desktop/BI_references_signatures_annotations/Mammarygland_Signatures_cellTypes/LAlv_Chen2021.txt")
head(Signature_ID_CreN_10x) ; head(LAlv_Chen2021)

DEG <- res

# Define significant genes and direction of regulation
DEG$signif <- DEG$p_val_adj < 0.05
DEG$FC <- rep(NA, nrow(DEG))
DEG$FC[which(DEG$avg_log2FC > 0 & DEG$signif ==TRUE)] = "UP"
DEG$FC[which(DEG$avg_log2FC < 0 & DEG$signif ==TRUE)] = "DW"

DEG$gene <- rownames(res)

# Add biological annotations from reference signatures
DEG <- DEG %>% 
  left_join(Signature_ID_CreN_10x, by = c("gene" = "Gene")) %>%
  mutate(Type = case_when(gene %in% LAlv_Chen2021$Spp1 ~ "Alv",
                          is.na(Type) &  signif == TRUE ~ "Others",
                          TRUE ~ Type))

# Select top significant genes for volcano plot labeling
label_g <- DEG %>% 
    dplyr::filter(p_val_adj < 0.05) %>%
    arrange(p_val_adj) %>% 
    ungroup() %>%
    group_by(FC) %>%   
    slice_head(n = 10) %>%
    ungroup() %>%
    pull(gene)


# ============================================================
# Volcano plot highlighting differentially expressed signatures
# ============================================================
gg5 <- ggplot(DEG,
       aes(x = (avg_log2FC), y = -log10(p_val_adj), fill = Type , size = signif, alpha = signif) )+
  geom_point(shape = 21, colour = "black")+ 
  geom_point(data = subset(DEG, signif == TRUE), 
             shape = 21, colour = "black") +
  geom_label_repel( data = subset(DEG, gene %in% label_g),
                   aes(label = gene),  force = 10,  max.overlaps =100) +
  
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") + 
  geom_vline(xintercept = c(log2(0.5), log2(2)),linetype = "dashed") +
  ggtitle("Mutant vs. WT")+
  
  scale_fill_manual(values = c("#F7DCAD",  "#236ec4",  "#a01858", "#f9c40c" ,  "grey80" )) +
  scale_size_manual(values = c(2,2,1)) + 
  scale_alpha_manual(values = c(0.5, 1,1)) +
  scale_x_continuous(breaks = c(seq(-6, 6, 2)),       
                     limits = c(-6, 6))  +
  scale_y_continuous(limits = c(0,25))+
  theme_bw() +
  theme(panel.border = element_rect(colour = "black", fill = NA, size= 0.5),    
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank())
gg5
```
### Ext Fig. 13e - Violin

```{r}
# ============================================================
# Violins to visualize expression of selected basal and luminal marker genes
# ============================================================

VlnPlot(seurat_rna, group.by = "ID_DEGs",
        features = c("Krt14", "Krt5","Acta2", "Myl9", 
                     "Krt18", "Krt8","Csn3","Lgals3", "B2m"), pt.size = 0)&
  scale_x_discrete(limits = c("CreN_CoherentFACSRNA_basal","CreP_LATE_CoherentFACSRNA_basal", " ",
                              
                               "CreP_LATE_CoherentFACSRNA_Luminal"))&
  scale_fill_manual(values = c("#236ec4","#61980E", "#a01858",
                               rep("grey", 17)))&
  geom_pointrange(
    stat = "summary",
    fun.min = function(z) {quantile(z,0.25)},
    fun.max = function(z) {quantile(z,0.75)},
    fun = median )&
  theme(axis.text.x = element_text(size = 8) )
```


# sc Multi


##Generation of gene-level multi-omic feature space and UMAP visualization 
 Objective:
 Generate a gene-level multi-omic feature space by integrating:
 - transcriptional signal (RNA expression)
 - active chromatin signal (H3K4me1 gene activity)
 - repressive chromatin signal (H3K27me3 gene activity)

 Each observation corresponds to one gene.
 The feature space is generated from cell-level multi-omic measurements, and genes are projected using PCA/UMAP.

 Additional variables (not used for PCA/UMAP construction) are calculated to quantify gene activity/expression changes between CreN and CreP groups.


```{r}
# ============================================================
# CreN LP + Mutant cells for FC
# ============================================================

#Retrieve metadat
meta <- seurat@meta.data
meta$rownames <- rownames(meta)


# Extract multi-omic matrices
rna_mat <- GetAssayData( seurat,  assay = "SCT_umi", layer = "counts" )

h3k4me1_mat <- GetAssayData(  seurat, assay = "ga_h3k4me1")

h3k27me3_mat <- GetAssayData(  seurat, assay = "ga_h3k27me3")


# Select cells of interest:
# - CreN LP cells (control luminal population)
# - CreP early and late pre-tumoral populations
cells_keep <- meta %>%
  dplyr::filter(
    (Group == "CreN" & annot_rna_joint_alv == "LP") |
    (Group %in% c("CreP_EARLY", "CreP_LATE"))  ) 

# Keep only cells present in all three modalities
cells_use <- intersect(  colnames(rna_mat),  union(colnames(h3k4me1_mat),        colnames(h3k27me3_mat)))
cells_use <- Reduce(intersect, list(cells_keep$rownames, cells_use))

cells_keep <- cells_keep %>%
  dplyr::filter(rownames %in% cells_use)

# Keep only genes retrieve in all three modalities
genes_use <- Reduce(  intersect,  list(rownames(rna_mat), rownames(h3k4me1_mat), rownames(h3k27me3_mat)))

rna_mat       <- rna_mat[genes_use, cells_use]
h3k4me1_mat   <- h3k4me1_mat[genes_use, colnames(h3k4me1_mat) %in% cells_use]
h3k27me3_mat  <- h3k27me3_mat[genes_use, colnames(h3k27me3_mat) %in% cells_use]

dim(rna_mat) ; dim(h3k4me1_mat) ; dim(h3k27me3_mat)
```


```{r}
# ============================================================
# CreN LP only
# ============================================================
# Extract multi-omic matrices for CreN LP cells only
# This allows construction of a reference gene-level feature space based on the control population

rna_mat_CreN <- GetAssayData( seurat,  assay = "SCT_umi", layer = "counts" )

h3k4me1_mat_CreN <- GetAssayData(  seurat, assay = "ga_h3k4me1")

h3k27me3_mat_CreN  <- GetAssayData(  seurat, assay = "ga_h3k27me3")


cells_keep_CreN <- meta %>%
  dplyr::filter(Group == "CreN" & annot_rna_joint_alv == "LP") 

cells_use <- intersect(  colnames(rna_mat),  union(colnames(h3k4me1_mat),        colnames(h3k27me3_mat)))
cells_use <- Reduce(intersect, list(cells_keep_CreN$rownames, cells_use))

cells_keep_CreN <- cells_keep_CreN %>%
  dplyr::filter(rownames %in% cells_use)


genes_use <- Reduce(  intersect,  list(rownames(rna_mat), rownames(h3k4me1_mat), rownames(h3k27me3_mat)))

rna_mat_CreN        <- rna_mat_CreN [genes_use, cells_use]
h3k4me1_mat_CreN    <- h3k4me1_mat_CreN [genes_use, colnames(h3k4me1_mat) %in% cells_use]
h3k27me3_mat_CreN   <- h3k27me3_mat_CreN [genes_use, colnames(h3k27me3_mat) %in% cells_use]

dim(rna_mat_CreN ) ; dim(h3k4me1_mat_CreN ) ; dim(h3k27me3_mat_CreN )
```


```{r}
# ============================================================
# PCA and UMAP projection of genes
# ============================================================
# Build gene-level multi-omic feature matrix
# Rows = cells
# Columns = gene-associated features across modalities

gene_feature_mat <- cbind(rna_mat, h3k4me1_mat,h3k27me3_mat)

# Reference feature space generated from CreN LP cells only
gene_feature_mat_CreN <- cbind(rna_mat_CreN, h3k4me1_mat_CreN,h3k27me3_mat_CreN)


# Compute UMAP embedding where each point represents one gene
# Genes are positioned according to their multi-omic profiles
umap <- uwot::umap(data.frame(gene_feature_mat_CreN), n_neighbors = 10**2, metric = "correlation")
```


```{r}
# ============================================================
# Calculation of gene-level signal scores
# ============================================================

# Function to calculate mean signal per gene within a selected group
mean_by_group <- function(mat, groups, target){
  rowMeans(mat[ , groups %in% target], na.rm = T)
}

# Calculate average RNA expression signal
group_vec <- cells_keep$Group
rna_CreN <- mean_by_group(rna_mat, group_vec, "CreN")
rna_CreP <- mean_by_group(rna_mat, group_vec, c("CreP_EARLY", "CreP_LATE"))
rna_CreP_e <- mean_by_group(rna_mat, group_vec, c("CreP_EARLY"))
rna_CreP_l <- mean_by_group(rna_mat, group_vec, c("CreP_LATE"))

# Calculate average H3K4me1 gene activity signal
group_vec <- cells_keep$Group[cells_keep$mark == "h3k4me1"]
h3k4me1_CreN <- mean_by_group(h3k4me1_mat, group_vec, "CreN")
h3k4me1_CreP <- mean_by_group(h3k4me1_mat, group_vec, c("CreP_EARLY", "CreP_LATE"))
h3k4me1_CreP_e <- mean_by_group(h3k4me1_mat, group_vec, c("CreP_EARLY"))
h3k4me1_CreP_l <- mean_by_group(h3k4me1_mat, group_vec, c("CreP_LATE"))

# Calculate average H3K27me3 gene activity signal
group_vec <- cells_keep$Group[cells_keep$mark == "h3k27me3"]
h3k27me3_CreN <- mean_by_group(h3k27me3_mat, group_vec, "CreN")
h3k27me3_CreP <- mean_by_group(h3k27me3_mat, group_vec, c("CreP_EARLY", "CreP_LATE"))
h3k27me3_CreP_e <- mean_by_group(h3k27me3_mat, group_vec, c("CreP_EARLY"))
h3k27me3_CreP_l <- mean_by_group(h3k27me3_mat, group_vec, c("CreP_LATE"))



# Store gene-level metadata variables
# These variables are used for visualization and interpretation, but are not included in the dimensionality reduction
gene_meta <- data.frame(
  gene = rownames(gene_feature_mat),
  RNA_CreN = rna_CreN,
  RNA_CreP = rna_CreP,
  RNA_CreP_e = rna_CreP_e,
  RNA_CreP_l = rna_CreP_l,
  H3K4me1_CreN = h3k4me1_CreN,
  H3K4me1_CreP = h3k4me1_CreP,
  H3K4me1_CreP_e = h3k4me1_CreP_e,
  H3K4me1_CreP_l = h3k4me1_CreP_l,
  H3K27me3_CreN = h3k27me3_CreN,
  H3K27me3_CreP = h3k27me3_CreP,
  H3K27me3_CreP_e = h3k27me3_CreP_e,
  H3K27me3_CreP_l = h3k27me3_CreP_l)


# Calculate normalized fold changes between CreP late and CreN populations
gene_meta$FC_RNA <- (gene_meta$RNA_CreP_l - gene_meta$RNA_CreN)/(gene_meta$RNA_CreP_l + gene_meta$RNA_CreN)
gene_meta$FC_H3K4me1 <- (gene_meta$H3K4me1_CreP_l - gene_meta$H3K4me1_CreN)/ (gene_meta$H3K4me1_CreP_l + gene_meta$H3K4me1_CreN)
gene_meta$FC_H3K27me3 <- (gene_meta$H3K27me3_CreP_l - gene_meta$H3K27me3_CreN )/(gene_meta$H3K27me3_CreP_l + gene_meta$H3K27me3_CreN )
head(gene_meta)
```


```{r}
#Merge UMAP coordinates with gene-level annotations
gene_umap <- cbind(
  gene_meta,
  UMAP1 = umap[,1],
  UMAP2 = umap[,2])
head(gene_umap)
```




```{r}
# Add gene identifiers as a column for visualization
gene_umap$GeneID <- rownames(gene_umap)
```
### Fig. 4i

Visualization of the gene-level UMAP embedding generated from integrated RNA expression, H3K4me1 and H3K27me3 gene activity features.
 Each point represents one gene.
Gene coordinates are fixed from the multi-omic feature space, while colors represent independent gene-level signals.
```{r}
# ============================================================
# Gene-level multi-omic feature space visualization
# ============================================================

# Define color palettes for each  modality

pal_RNA   <- c("grey99", "#FCE9F2", "#F6C1DA", "#F099C1", "#EA71A9", "#E44991", 
               "#E03283", "#B61B63", "#8E154E", "#660F38", "#3E0922", "#16030C")  
pal_K4    <- c( "grey99" , "#FCF3E9","#F6DDC1", "#F0C799", "#EAB171", "#E49C49", "#E08F32", "#B66E1B", "#8E5615", "#663E0F", "#3E2509", "#160D03")  
pal_K27   <- c("grey99","#E9F2FC", "#C1DAF6",  "#99C1F0", "#71A9EA", "#4991E4", "#3283E0", "#1B63B6", "#154E8E", "#0F3866", "#09223E", "#030C16")  




# Function to visualize gene-level signals on the UMAP embedding
#
# Input:
# - data: gene-level UMAP table
# - var: gene-level variable used for coloring
# - title: plot title
# - cols: color gradient
#
# Highlighted genes are displayed with labels

plot_gene_umap <- function(data, var, title, cols,limits =c(0, 10), raster_dpi = 600,
                           genes_label = c("Krt14", "Col6a1","Mylk",
                                            "Cd81", "Lalba", "Krt18",
                                            "Cd82","Cdk19","Ly6d",
                                            "F3", "Csf1", "Il23a")){

  ggplot( data %>% 
            dplyr::filter(!is.na(.data[[var]])) %>%
            arrange(.data[[var]], na.rm = T),
          aes(UMAP1, UMAP2, 
              fill = (.data[[var]]  ),
              color = (.data[[var]]  )  )  ) +
    
        # Rasterized points to improve rendering of large gene sets
    ggrastr::geom_point_rast(  size = 0.6,   shape = 21,    alpha = 0.1,  raster.dpi = raster_dpi  ) +
    
        # Highlight selected marker genes
     geom_point(
      data = data %>% dplyr::filter(gene %in% genes_label),
      color = "black",size = 1, shape = 21) +
    
        # Add gene labels
    geom_text_repel(
      data = data %>% dplyr::filter(gene %in% genes_label),
      aes(label = gene),
      size = 2,  color = "black",  box.padding = 0.4,  point.padding = 0.3,   max.overlaps = Inf)+ 
    
    scale_fill_gradientn(  colours = cols , na.value = "white" , limits = limits) +
    scale_color_gradientn(  colours = cols , na.value = "white", limits = limits  ) +
    theme_classic() + theme(legend.key.size = unit(0.1, 'cm'), legend.title = element_blank())+
    ggtitle(title)
}

# Plot gene-level signals in the reference CreN-based UMAP space

p1 <- plot_gene_umap(gene_umap, "RNA_CreN",      "RNA – CreN",      pal_RNA, limits = c(0,3))
p2 <- plot_gene_umap(gene_umap, "H3K4me1_CreN",  "H3K4me1 – CreN",  pal_K4, limits = c(0,6))  
p3 <- plot_gene_umap(gene_umap, "H3K27me3_CreN", "H3K27me3 – CreN", pal_K27, limits = c(0,10))





(p1 | p2 | p3) 

```

```{r}
# Display genes with the strongest increase in RNA and H3K4me1 signals
# between CreP late and CreN populations

 gene_umap %>% arrange( -FC_H3K4me1 , -FC_RNA      ) %>% dplyr::select(FC_H3K27me3, FC_H3K4me1, FC_RNA ) %>% head(200)

```


Visualization of differential multi-omic signals (CreP late vs CreN)
```{r}
# Function to visualize normalized fold-change values
# on the gene-level UMAP embedding
#
# Positive values indicate increased signal in CreP late cells,
# whereas negative values indicate decreased signal.

 plot_gene_umap <- function(data, var, title, cols,
                           n_points = 11570,
                           raster_dpi = 600,
                            genes_label =c("Krt14", "Col6a1","Mylk",
                                            "Cd81", "Lalba", "Krt18", "Cd82","Cdk19",
                                            "Ly6d", "F3", "Csf1", "Il23a")){

    data <- data %>% dplyr::filter(!is.na(.data[[var]]))
    v <- data[[var]]

  # Define symmetric color scale centered on zero
    borne <- max(abs(v), na.rm = TRUE)
    
     # Subsample genes for background visualization
    data_bg <- data %>%
    dplyr::slice_sample(n = min(n_points, nrow(data)))
    
      # Always retain selected labeled genes
    data_lab <- data %>% dplyr::filter(gene %in% genes_label) 

  # Combine background and highlighted genes
    data_plot <- dplyr::bind_rows(data_bg, data_lab) %>%
    dplyr::distinct()

  ggplot( data_plot %>%  arrange(#abs
    (.data[[var]]), na.rm = T),
          aes(UMAP1, UMAP2, 
              fill = (.data[[var]]  ),
              color = (.data[[var]]  )  )  ) +
    
    ggrastr::geom_point_rast(  size = 0.6,   shape = 21,    alpha = 0.5,  raster.dpi = raster_dpi  ) +

     geom_point(data = data %>%  dplyr::filter(gene %in% genes_label), aes(fill = (.data[[var]]  )),
      color = "black",size = 2, shape = 21) +
    geom_text_repel(data = data %>%  dplyr::filter(gene %in% genes_label),
      aes(label = gene),
      size = 2,  color = "black",  box.padding = 0.4,  point.padding = 0.3,   max.overlaps = Inf)+ 
    scale_fill_gradientn(  colours = cols , na.value = "white", 
                           limits = c(-borne, borne) ) +
    scale_color_gradientn(  colours = cols , na.value = "white" , limits = c(-borne, borne) ) +
    theme_classic() + theme(legend.key.size = unit(0.1, 'cm'), legend.title = element_blank())+
    ggtitle(title)
}

# Diverging palette for fold-change visualization
pal_div <- c(  "grey90","grey80",  "#61980E"  )

grid.arrange(
  
  plot_gene_umap(gene_umap, "FC_RNA",      "RNA – FC",      pal_div),
  plot_gene_umap(gene_umap, "FC_H3K4me1",   "H3K4me1 – FC",      pal_div),
  plot_gene_umap(gene_umap, "FC_H3K27me3",  "H3K27me3 – FC",      pal_div),

ncol = 3)


```






### Ext Fig. 12 d-e
```{r}
# For Louvain's clustering:
# The input matrix contains the multi-omic feature space generated from CreN cells only.

seu <-  t(gene_feature_mat_CreN)
rownames(seu) <- 1:nrow(seu)
seu <- CreateSeuratObject(counts =seu)
seu <- NormalizeData(seu)
seu <- ScaleData(seu)
seu <- FindVariableFeatures(seu)
seu <- RunPCA(seu, npcs = 20)
seu <- FindNeighbors(seu, dims = 1:20, k.param = 100)
seu <- FindClusters(seu, resolution = 0.5)

DimPlot(seu)

gene_cluster <- seu$seurat_clusters
```
```{r}
# Add gene clusters informations
umap_df <- data.frame(
  umap_1 = umap[,1],
  umap_2 = umap[,2],
  cluster = gene_cluster
)

#Visualize
ggplot(umap_df, aes(umap_1, umap_2, color = cluster)) +
  geom_point(size = 1) +
  theme_classic()
```

```{r}
# ============================================================
# Functional enrichment analysis using Hallmark gene sets
# ============================================================

hallmarks <- msigdbr(
  species = "Homo sapiens",
  category = "H") %>%
  dplyr::select(gs_name, gene_symbol)

hallmark_list <- split(hallmarks$gene_symbol, hallmarks$gs_name)

# Perform pathway enrichment analysis independently for each gene cluster
enrich_per_cluster <- lapply(levels(gene_cluster), function(cl){

    # Extract genes belonging to the current cluster
  genes_cl <- rownames(gene_feature_mat_CreN)[gene_cluster == cl]

    # Identify enriched Hallmark pathways
  enrich <- enricher(
    gene          = toupper(genes_cl),
    TERM2GENE     = hallmarks,
    pvalueCutoff  = 0.05,
    pAdjustMethod = "fdr"
  )

  if (is.null(enrich)) return(NULL)

  enrich@result$cluster <- cl
  enrich@result
})

hall_df <- bind_rows(enrich_per_cluster)

# Convert GeneRatio into numeric values for visualization
hall_df <- hall_df %>%
  mutate(
    GeneRatio_num = as.numeric(sub("/.*", "", GeneRatio)) /
                    as.numeric(sub(".*/", "", GeneRatio))
  )

# Select the top enriched pathways per gene cluster
plot_df <- hall_df %>%
  group_by(cluster) %>%
  arrange(p.adjust) %>%
  slice_head(n = 5) %>%
  ungroup()

pval_limits <- range(plot_df$p.adjust, na.rm = TRUE)


# Plot Hallmark pathway enrichment across gene clusters
ggplot(data.frame(plot_df),
       aes(x = GeneRatio_num,
           y = reorder(Description, GeneRatio_num),
           fill = log10(p.adjust))) +
  geom_col(width = 0.8) +
  facet_wrap(~ cluster, scales = "free_y") +
  scale_fill_gradientn(  colours = rev(pal_RNA) , na.value = "white") +
  theme_classic() +
  labs(    x = "Gene ratio",    y = NULL,
    title = "Hallmark enrichment per gene cluster"  ) +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    axis.text.y = element_text(size = 9),
    axis.text.x = element_text(size = 10),
    plot.title = element_text(size = 14, face = "bold")  )
```


```{r}
# ============================================================
# Functional enrichment analysis using mammary epithelial signatures
# ============================================================

# Load previously defined mammary epithelial cell signatures
Signature_ID_CreN_10x <- read.delim("C:/Users/Marthe Laisne/Desktop/BI_references_signatures_annotations/Mammarygland_Signatures_Custom/SignatureCreN_scRNA10x_Epithelial_poles.txt")

# Add alveolar lineage signature from Chen et al. 2021
LAlv_Chen2021  <- read.delim("C:/Users/Marthe Laisne/Desktop/BI_references_signatures_annotations/Mammarygland_Signatures_cellTypes/LAlv_Chen2021.txt", col.names = F)

Sign <- data.frame(Type = c(Signature_ID_CreN_10x$Type,
                            rep("Alv", nrow(LAlv_Chen2021))),
                   Gene = c(Signature_ID_CreN_10x$Gene,
                            LAlv_Chen2021$FALSE.))

# Select the top 100 genes from each cell identity signature
Sign <- Sign %>%
  group_by(Type) %>%
  slice_head(n = 100) %>%  
  ungroup()

# Format custom gene sets for clusterProfiler enrichment analysis
custom_sets <- Sign %>%
  mutate(
    Gene = toupper(Gene),
    Type = as.character(Type)  ) %>%
  dplyr::select(
    gs_name = Type,
    gene_symbol = Gene  )

# Number of genes per custom signature
table(custom_sets$gs_name)
```

```{r}
# Perform enrichment analysis using custom mammary epithelial signatures

enrich_per_cluster <- lapply(levels(gene_cluster), function(cl){

  genes_cl <- rownames(gene_feature_mat_CreN)[gene_cluster == cl]

  enrich <- enricher(
    gene          = toupper(genes_cl),
    TERM2GENE     = custom_sets,
    pvalueCutoff  = 0.05,
    pAdjustMethod = "fdr"  )

  if (is.null(enrich) || nrow(enrich@result) == 0) return(NA)

  enrich@result$cluster <- cl
  enrich@result
})

custom_df <- bind_rows(enrich_per_cluster)

# Convert GeneRatio into numeric values
custom_df <- custom_df %>%
  mutate(
    GeneRatio_num =
      as.numeric(sub("/.*", "", GeneRatio)) /
      as.numeric(sub(".*/", "", GeneRatio))
  )

plot_df <- custom_df %>%
  dplyr::filter(cluster %in% c(0:5)) %>%
  group_by(cluster) %>%
  arrange(p.adjust) %>%
  ungroup()
```



```{r}
# Visualize enrichment of mammary epithelial signatures across gene clusters
ggplot(plot_df,
       aes( x = GeneRatio_num, y = reorder(Description, GeneRatio_num),   fill = log10(p.adjust)   )) +
  geom_col(width = 0.8) +
  facet_wrap(~ cluster, scales = "free_y") +
  scale_fill_gradientn(  colours = rev(pal_RNA),    na.value = "white",    name = "log10 FDR"  ) +
  theme_classic() +
  labs(    x = "Gene ratio",    y = NULL,    title = "Custom signature enrichment per gene cluster"  ) +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    axis.text.y = element_text(size = 9),
    axis.text.x = element_text(size = 10),
    plot.title = element_text(size = 14, face = "bold")  )
```































