---
title: "5.1 scMultiome preprocessing and annotation"
author: "Anna Schwager"
date: "`r Sys.Date()`"
output: html_document
---

```{r silence, include=FALSE, message=F, warning=F}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.show = "hold"
)
```


# Python setup
```{r reticulate, message=F, warning=F}

library(reticulate)
use_condaenv(
  "snap_py310",
  conda = "/Users/annaschwager/miniforge3/bin/conda",
  required = TRUE
)

```

```{python check-python, message=F, warning=F}
import sys
import numpy as np
import pandas as pd
from pathlib import Path
import scanpy as sc
import snapatac2 as snap
import skmisc

```

# R setup
```{r source, message=F, warning=F}

here::i_am("scripts/5.1_scMultiome_PreProcess_Annotation.Rmd")
mainDir = here::here()

set.seed(123)
```

```{r libraries_new, message=F, warning=F}
library(Signac)
library(Seurat)
library(harmony)
library(EnsDb.Mmusculus.v79)
library(org.Mm.eg.db)
library(tidyverse)
library(biovizBase)
library(ggsci)
library(scales)
library(pheatmap)
library(grid)
library(gridExtra)
library(rlang)
library(ggpointdensity)
library(tidytext)
library(patchwork)
library(GenomeInfoDb)
library(SingleCellExperiment)
library(ComplexHeatmap)
library(destiny)
library(ggalluvial)
library(ggrepel)
library(dplyr)
library(qs)
library(circlize)
library(GeneNMF)
library(UCell)
```

# Setting directories and input files
```{r dir}
mainDir = here::here()
inputDir =  file.path(mainDir,"input") 
outputDir = file.path(mainDir, "output")
annotDir = file.path(mainDir, "annotation")

outputDir_local = file.path(outputDir, "epigenomic_diffusion") ; if(!file.exists(outputDir_local)){dir.create(outputDir_local)}
outputDir_objects = file.path(outputDir_local, "objects") ; if(!file.exists(outputDir_objects)){dir.create(outputDir_objects)}
outputDir_plots = file.path(outputDir_local, "plots") ; if(!file.exists(outputDir_plots)){dir.create(outputDir_plots)}
outputDir_tables = file.path(outputDir_local, "tables") ; if(!file.exists(outputDir_tables)){dir.create(outputDir_tables)}

snap_dir <- file.path(outputDir_objects, "snap_export") ; if(!file.exists(snap_dir )){dir.create(snap_dir)}
```

# Set colors
```{r col}
mypal <- c("#4eb7b4","#F9C22EFF","#e4722c", "#c94c9d", "#5c4d7d", "#469D76FF", "#f7dcad", "#3C4B99FF", "#DF9ED4FF", "#C93F55FF", "#E59A52FF", "#EDCE79FF", "pink", "grey", "#92C051FF")

cell_cols <- c(
  basal   = "#236ec4",
  luminal = "#ac0000",
  LP      = "#a3215e",
  HS      = "#F9C22EFF",
  Alv     = "#f7dcad"
)

group_cols <- c(CreP_EARLY = "#c0d69f", 
                CreP_LATE = "#61980e",
                CreN = "#5c4d7d",
                Rosa = "#c94c9d")

```

# Load annotation
```{r load_annot, message=F, warning=F}
ref_inhouse <- read.delim(file.path(annotDir, "Landragin2025_CreN_mammary_gland_markers.txt"))
ref_visvader <- read.delim(file.path(annotDir, "Visvader2010_mammary_gland_markers.txt"))
```


# Load final clean object
Object are aready clean and quality-filtered
```{r load, message=F, warning=F, eval=F}
# Multi-modal object with all epi cells and a subset of RNA cells
seurat <- qread(file.path(outputDir_objects, "seurat_merged_final.qs"))
# RNA-only object with all RNA cells
seurat_rna <- qread(file.path(outputDir_objects, "seurat_merged_rna_final.qs"))
```


# Exporting data from Seurat to SnapATAC2
## RNA
```{r snap1, message=F, warning=F, eval=F}
seurat <- JoinLayers(seurat, assay = "RNA_umi")
rna_assay <- GetAssayData(seurat, assay = "RNA_umi", layer = "counts")

Matrix::writeMM(rna_assay, file.path(snap_dir, "matrix_rna_umi.mtx"))
write.table(colnames(rna_assay), file.path(snap_dir, "barcodes_rna_umi.tsv"),
            quote = FALSE, sep = "\t", row.names = FALSE, col.names = FALSE)

genes_df <- data.frame(
  name  = rownames(rna_assay),
  stringsAsFactors = FALSE
)
write.table(genes_df, file.path(snap_dir, "features_rna_umi.tsv"),
            quote = FALSE, sep = "\t", row.names = FALSE, col.names = TRUE)

```

RNA only object
```{r snap2, message=F, warning=F, eval=F}
seurat_rna <- JoinLayers(seurat_rna, assay = "RNA_umi")
rna_assay_rna <- GetAssayData(seurat_rna, assay = "RNA_umi", layer = "counts")

Matrix::writeMM(rna_assay_rna, file.path(snap_dir, "matrix_rna_umi_rna.mtx"))
write.table(colnames(rna_assay_rna), file.path(snap_dir, "barcodes_rna_umi_rna.tsv"),
            quote = FALSE, sep = "\t", row.names = FALSE, col.names = FALSE)

genes_df_rna <- data.frame(
  name  = rownames(rna_assay_rna),
  stringsAsFactors = FALSE
)
write.table(genes_df_rna, file.path(snap_dir, "features_rna_umi_rna.tsv"),
            quote = FALSE, sep = "\t", row.names = FALSE, col.names = TRUE)

```

## H3K4me1
```{r snap3, message=F, warning=F, eval=F}
k4_assay <- GetAssayData(seurat, assay = "bin_5k_h3k4me1", layer = "counts")

Matrix::writeMM(k4_assay, file.path(snap_dir, "matrix_k4.mtx"))
write.table(colnames(k4_assay), file.path(snap_dir, "barcodes_k4.tsv"),
            quote = FALSE, sep = "\t", row.names = FALSE, col.names = FALSE)


bins_df <- data.frame(
  name = rownames(k4_assay),
  stringsAsFactors = FALSE
)

bins_df$chrom <- sub("-.*", "", bins_df$name)
bins_df$start <- as.integer(sub("^chr[^-]+-(\\d+)-.*", "\\1", bins_df$name))
bins_df$end   <- as.integer(sub("^chr[^-]+-\\d+-(\\d+)$", "\\1", bins_df$name))

bins_df <- bins_df[, c("chrom", "start", "end", "name")]

write.table(
  bins_df,
  file.path(snap_dir, "features_k4.tsv"),
  quote = FALSE,
  sep = "\t",
  row.names = FALSE,
  col.names = TRUE
)
```

## H3K27me3
```{r snap4, message=F, warning=F, eval=F}

k27_assay <- GetAssayData(seurat, assay = "bin_5k_h3k27me3", layer = "counts")

Matrix::writeMM(k27_assay, file.path(snap_dir, "matrix_k27.mtx"))
write.table(colnames(k27_assay), file.path(snap_dir, "barcodes_k27.tsv"),
            quote = FALSE, sep = "\t", row.names = FALSE, col.names = FALSE)


bins_df <- data.frame(
  name  = rownames(k27_assay),
  stringsAsFactors = FALSE
)

bins_df$chrom <- sub("-.*", "", bins_df$name)
bins_df$start <- as.integer(sub("^chr[^-]+-(\\d+)-.*", "\\1", bins_df$name))
bins_df$end   <- as.integer(sub("^chr[^-]+-\\d+-(\\d+)$", "\\1", bins_df$name))

bins_df <- bins_df[, c("chrom", "start", "end", "name")]

write.table(bins_df, file.path(snap_dir, "features_k27.tsv"),
            quote = FALSE, sep = "\t", row.names = FALSE, col.names = TRUE)

```

# Spectral embeddings generation
## UMI RNA
Multi-modal object
```{python snap5, message=F, warning=F, eval=F}

snap_dir = Path(r.snap_dir) 
X = sc.read_mtx(snap_dir / "matrix_rna_umi.mtx").X.tocsr().transpose()
barcodes = pd.read_csv(snap_dir / "barcodes_rna_umi.tsv",
                       header=None, sep="\t")[0].values
feat = pd.read_csv(snap_dir / "features_rna_umi.tsv", sep="\t")

adata_rna = sc.AnnData(X=X)
adata_rna.obs_names = barcodes
adata_rna.var_names = feat["name"].values
adata_rna.X = adata_rna.X.tocsr().astype(np.float64)

adata_rna
```

```{python snap6, message=F, warning=F, eval=F}
sc.pp.highly_variable_genes(adata_rna, flavor='seurat_v3', n_top_genes=3000)
adata_rna = adata_rna[:, adata_rna.var.highly_variable]

sc.pp.normalize_total(adata_rna, target_sum=1e4)
sc.pp.log1p(adata_rna)

adata_rna
```

```{python snap7, message=F, warning=F, eval=F}
snap.tl.spectral(adata_rna, features=None)
```

RNA only object
```{python snap8, message=F, warning=F, eval=F}
snap_dir = Path(r.snap_dir) 
X_rna = sc.read_mtx(snap_dir / "matrix_rna_umi_rna.mtx").X.tocsr().transpose()
barcodes_rna = pd.read_csv(snap_dir / "barcodes_rna_umi_rna.tsv",
                       header=None, sep="\t")[0].values
feat_rna = pd.read_csv(snap_dir / "features_rna_umi_rna.tsv", sep="\t")

adata_rna_rna = sc.AnnData(X=X_rna)
adata_rna_rna.obs_names = barcodes_rna
adata_rna_rna.var_names = feat_rna["name"].values
adata_rna_rna.X = adata_rna_rna.X.tocsr().astype(np.float64)

adata_rna_rna
```

```{python snap9, message=F, warning=F, eval=F}
sc.pp.highly_variable_genes(adata_rna_rna, flavor='seurat_v3', n_top_genes=3000)
adata_rna_rna = adata_rna_rna[:, adata_rna_rna.var.highly_variable]

sc.pp.normalize_total(adata_rna_rna, target_sum=1e4)
sc.pp.log1p(adata_rna_rna)

adata_rna_rna
```

```{python snap10, message=F, warning=F, eval=F}
snap.tl.spectral(adata_rna_rna, features=None)
```

## H3K4me1
```{python snap11, message=F, warning=F, eval=F}
snap_dir = Path(r.snap_dir) 
X = sc.read_mtx(snap_dir / "matrix_k4.mtx").X.tocsr().transpose()  
barcodes = pd.read_csv(snap_dir / "barcodes_k4.tsv", header=None, sep="\t")[0].values
feat = pd.read_csv(snap_dir / "features_k4.tsv", sep="\t")

adata_k4 = sc.AnnData(X=X)
adata_k4.obs_names = barcodes
adata_k4.var_names = feat["name"].values
adata_k4.var["chrom"] = feat["chrom"].values
adata_k4.var["start"] = feat["start"].values
adata_k4.var["end"]   = feat["end"].values

adata_k4.X = adata_k4.X.tocsr().astype(np.float64)
adata_k4
```

```{python snap12, message=F, warning=F, eval=F}
snap.pp.select_features(adata_k4, n_features = 300000)
snap.tl.spectral(adata_k4, features = 'selected', distance_metric='cosine')
```

## H3K27me3
```{python snap13, message=F, warning=F, eval=F}
snap_dir = Path(r.snap_dir) 
X = sc.read_mtx(snap_dir / "matrix_k27.mtx").X.tocsr().transpose()  
barcodes = pd.read_csv(snap_dir / "barcodes_k27.tsv", header=None, sep="\t")[0].values
feat = pd.read_csv(snap_dir / "features_k27.tsv", sep="\t")

adata_k27 = sc.AnnData(X=X)
adata_k27.obs_names = barcodes
adata_k27.var_names = feat["name"].values
adata_k27.var["chrom"] = feat["chrom"].values
adata_k27.var["start"] = feat["start"].values
adata_k27.var["end"]   = feat["end"].values

adata_k27.X = adata_k27.X.tocsr().astype(np.float64)
adata_k27
```

```{python snap14, message=F, warning=F, eval=F}
snap.pp.select_features(adata_k27, n_features = 300000)
snap.tl.spectral(adata_k27, features = 'selected', distance_metric='cosine')
```

# Import embeddings to Seurat
```{r import1, message=F, warning=F, eval=F}
emb_rna <- as.matrix(py$adata_rna$obsm[["X_spectral"]])
rownames(emb_rna) <- py$adata_rna$obs_names$tolist()

emb_rna_rna <- as.matrix(py$adata_rna_rna$obsm[["X_spectral"]])
rownames(emb_rna_rna) <- py$adata_rna_rna$obs_names$tolist()

emb_k4 <- as.matrix(py$adata_k4$obsm[["X_spectral"]])
rownames(emb_k4) <- py$adata_k4$obs_names$tolist()

emb_k27 <- as.matrix(py$adata_k27$obsm[["X_spectral"]])
rownames(emb_k27) <- py$adata_k27$obs_names$tolist()

```

```{r import2, message=F, warning=F, eval=F}
seurat[["spectral_rna_umi"]] <- CreateDimReducObject(
  embeddings = emb_rna,
  key       = "specrnaumi_",
  assay     = "RNA_umi" 
)

seurat_rna[["spectral_rna_umi"]] <- CreateDimReducObject(
  embeddings = emb_rna_rna,
  key       = "specrnaumi_",
  assay     = "RNA_umi" 
)

seurat[["spectral_k4"]] <- CreateDimReducObject(
  embeddings = emb_k4,
  key       = "speck4_",
  assay     = "bin_5k_h3k4me1"
)

seurat[["spectral_k27"]] <- CreateDimReducObject(
  embeddings = emb_k27,
  key       = "speck27_",
  assay     = "bin_5k_h3k27me3"
)

```

# Reload step 1
```{r save_step1, message=F, warning=F, eval=F}
qsave(seurat, file.path(outputDir_objects, "seurat_step1.qs"))
qsave(seurat_rna, file.path(outputDir_objects, "seurat_rna_step1.qs"))
```

```{r load_step1, message=F, warning=F}
seurat <- qread(file.path(outputDir_objects, "seurat_step1.qs"))
seurat_rna <- qread(file.path(outputDir_objects, "seurat_rna_step1.qs"))

```

# RNA clustering
## SCT transform
```{r SCT, message=F, warning=F}
seurat <- SCTransform(seurat, assay = "RNA_umi", new.assay.name = "SCT_umi", vars.to.regress = c("nCount_RNA_umi"))

seurat_rna <- SCTransform(seurat_rna, assay = "RNA_umi", new.assay.name = "SCT_umi", vars.to.regress = c("nCount_RNA_umi"))

```

Multimodal object
```{r umpap1, message=F, warning=F}
seurat <- RunUMAP(seurat,
                  assay = "RNA_umi",
                  seed.use = 42,
                  dims = 1:length(colnames(seurat@reductions[["spectral_rna_umi"]])),
                  reduction = "spectral_rna_umi",
                  reduction.name = "UMAP_spectral_rna_umi")

seurat <- FindNeighbors(seurat,
                        reduction = "spectral_rna_umi",
                        cluster.name = "spectral_rna_umi_clusters",
                        graph.name = "spectral_rna_umi_snn",
                        k.param = 8)

seurat <- FindClusters(seurat,
                       graph.name   = "spectral_rna_umi_snn",
                       resolution = 0.05,
                       cluster.name = "spectral_rna_umi_clusters",
                       algorithm = 4)

```

```{r umap, message=F, warning=F}
DimPlot(seurat, cols = mypal,
        reduction = "UMAP_spectral_rna_umi",
        group.by = c("Exp",  "Group"),
        combine = TRUE)

DimPlot(seurat, cols = mypal,
        reduction = "UMAP_spectral_rna_umi",
        group.by = c("spectral_rna_umi_clusters", "facs_phenotype"),
        combine = TRUE)

FeaturePlot(seurat, reduction = "UMAP_spectral_rna_umi", feature = "percent.mt")

```


RNA only object
```{r umap2, message=F, warning=F}
seurat_rna <- RunUMAP(seurat_rna,
                  assay = "RNA_umi",
                  seed.use = 42,
                  dims = 1:length(colnames(seurat_rna@reductions[["spectral_rna_umi"]])),
                  reduction = "spectral_rna_umi",
                  reduction.name = "UMAP_spectral_rna_umi")

seurat_rna <- FindNeighbors(seurat_rna,
                        reduction = "spectral_rna_umi",
                        cluster.name = "spectral_rna_umi_clusters",
                        graph.name = "spectral_rna_umi_snn",
                        k.param = 15)

seurat_rna <- FindClusters(seurat_rna,
                       graph.name   = "spectral_rna_umi_snn",
                       resolution = 0.05,
                       cluster.name = "spectral_rna_umi_clusters",
                       algorithm = 4)

```

```{r umap3, message=F, warning=F}
DimPlot(seurat_rna, cols = mypal,
        reduction = "UMAP_spectral_rna_umi",
        group.by = c("Exp",  "Group"),
        combine = TRUE)

DimPlot(seurat_rna, cols = mypal,
        reduction = "UMAP_spectral_rna_umi",
        group.by = c("spectral_rna_umi_clusters", "facs_phenotype"),
        combine = TRUE)

```

## Batch correction
Multimodal object
```{r batch_rna, message=F, warning=F}
seurat <- RunHarmony(
  object        = seurat,
  group.by.vars = c("model"),
  reduction.use = "spectral_rna_umi",
  reduction.save = "spectral_rna_umi_harmony",
  project.dim    = FALSE,
  plot_convergence = TRUE,
  theta            = 1,
  lambda           = NULL,
  sigma = 0.06
)
```

RNA only object
```{r batch_rna2, message=F, warning=F}
seurat_rna <- RunHarmony(
  object        = seurat_rna,
  group.by.vars = c("model"),
  reduction.use = "spectral_rna_umi",
  reduction.save = "spectral_rna_umi_harmony",
  project.dim    = FALSE,
  plot_convergence = TRUE,
  theta            = 1,
  lambda           = NULL,
  sigma = 0.06
)

```

## Reclustering post batch-correction
Multimodal object
```{r cluster1, message=F, warning=F}
seurat <- RunUMAP(seurat,
                  assay = "RNA_umi",
                  seed.use = 172,
                  metric = "cosine",
                  n.neighbors = 60,
                  min.dist = 0.3,
                  spread = 1.5,
                  dims = 1:length(colnames(seurat@reductions[["spectral_rna_umi_harmony"]])),
                  reduction = "spectral_rna_umi_harmony",
                  reduction.name = "UMAP_spectral_rna_umi_harmony")

seurat <- FindNeighbors(seurat,
                        reduction = "spectral_rna_umi_harmony",
                        cluster.name = "spectral_rna_umi_harmony_clusters",
                        graph.name = "spectral_rna_umi_harmony_snn",
                        k.param = 8)

seurat <- FindClusters(seurat,
                       graph.name   = "spectral_rna_umi_harmony_snn",
                       resolution = 0.05,
                       cluster.name = "spectral_rna_umi_harmony_clusters",
                       algorithm = 4)

```


```{r cluster2, message=F, warning=F}
DimPlot(seurat, cols = mypal,
        reduction = "UMAP_spectral_rna_umi_harmony",
        group.by = c("Exp", "Group"),
        combine = TRUE)

DimPlot(seurat, cols = mypal,
        reduction = "UMAP_spectral_rna_umi_harmony",
        group.by = c("spectral_rna_umi_harmony_clusters","facs_phenotype"),
        combine = TRUE)

```

RNA only object
```{r cluster3, message=F, warning=F}
seurat_rna <- RunUMAP(seurat_rna,
                  assay = "RNA_umi",
                  seed.use = 42,
                  metric = "cosine",
                  n.neighbors = 60,
                  min.dist = 0.5,
                  spread = 1.1,
                  dims = 1:length(colnames(seurat_rna@reductions[["spectral_rna_umi_harmony"]])),
                  reduction = "spectral_rna_umi_harmony",
                  reduction.name = "UMAP_spectral_rna_umi_harmony")

seurat_rna <- FindNeighbors(seurat_rna,
                        reduction = "spectral_rna_umi_harmony",
                        cluster.name = "spectral_rna_umi_harmony_clusters",
                        graph.name = "spectral_rna_umi_harmony_snn",
                        k.param = 15)

seurat_rna <- FindClusters(seurat_rna,
                       graph.name   = "spectral_rna_umi_harmony_snn",
                       resolution = 0.05,
                       cluster.name = "spectral_rna_umi_harmony_clusters",
                       algorithm = 4)

```


```{r cluster4, message=F, warning=F}
DimPlot(seurat_rna, cols = mypal,
        reduction = "UMAP_spectral_rna_umi_harmony",
        group.by = c("Exp", "Group"),
        combine = TRUE)

#CellFateRNA - annotation from MutantAlv script
DimPlot(seurat_rna, cols = mypal,
        reduction = "UMAP_spectral_rna_umi_harmony",
        group.by = c("spectral_rna_umi_harmony_clusters","CellFate_RNA"),
        combine = TRUE)

```

## Cell type scores
```{r scores1, message=F, warning=F}
AddCellTypeScore <- function(seurat_obj,
                             marker_table,
                             cell_type,
                             assay = "SCT_umi",
                             slot = "data",
                             score_name = "cell_type_score") {
  marker_genes <- marker_table %>%
    filter(Type == cell_type) %>%
    pull(Gene) %>%
    unique()
  
  expr_matrix <- GetAssayData(seurat_obj, assay = assay, layer = slot)
  
  marker_genes_present <- intersect(marker_genes, rownames(expr_matrix))
  
  if (length(marker_genes_present) == 0) {
    warning(paste0("No marker genes found in assay for cell type: ", cell_type))
    seurat_obj[[score_name]] <- NA
    return(seurat_obj)
  }
  
  score_vector <- Matrix::colMeans(expr_matrix[marker_genes_present, , drop = FALSE])
  
  seurat_obj[[score_name]] <- score_vector
  
  return(seurat_obj)
}

```

Landragin
```{r scores2, message=F, warning=F}
seurat <- AddCellTypeScore(seurat_obj = seurat,
                 marker_table = ref_inhouse,
                 cell_type = "Basal",
                 assay = "SCT_umi",
                 score_name = "basal_score_Landragin")

seurat <- AddCellTypeScore(seurat_obj = seurat,
                 marker_table = ref_inhouse,
                 cell_type = "Luminal H-S",
                 assay = "SCT_umi",
                 score_name = "hs_score_Landragin")

seurat <- AddCellTypeScore(seurat_obj = seurat,
                 marker_table = ref_inhouse,
                 cell_type = "LP",
                 assay = "SCT_umi",
                 score_name = "lp_score_Landragin")

seurat_rna <- AddCellTypeScore(seurat_obj = seurat_rna,
                 marker_table = ref_inhouse,
                 cell_type = "Basal",
                 assay = "SCT_umi",
                 score_name = "basal_score_Landragin")

seurat_rna <- AddCellTypeScore(seurat_obj = seurat_rna,
                 marker_table = ref_inhouse,
                 cell_type = "Luminal H-S",
                 assay = "SCT_umi",
                 score_name = "hs_score_Landragin")

seurat_rna <- AddCellTypeScore(seurat_obj = seurat_rna,
                 marker_table = ref_inhouse,
                 cell_type = "LP",
                 assay = "SCT_umi",
                 score_name = "lp_score_Landragin")

```

Visvader
```{r scores3, message=F, warning=F}
seurat <- AddCellTypeScore(seurat_obj = seurat,
                 marker_table = ref_visvader,
                 cell_type = "Basal",
                 assay = "SCT_umi",
                 score_name = "basal_score_Visvader")

seurat <- AddCellTypeScore(seurat_obj = seurat,
                 marker_table = ref_visvader,
                 cell_type = "Luminal H-S",
                 assay = "SCT_umi",
                 score_name = "hs_score_Visvader")

seurat <- AddCellTypeScore(seurat_obj = seurat,
                 marker_table = ref_visvader,
                 cell_type = "LP",
                 assay = "SCT_umi",
                 score_name = "lp_score_Visvader")

seurat_rna <- AddCellTypeScore(seurat_obj = seurat_rna,
                 marker_table = ref_visvader,
                 cell_type = "Basal",
                 assay = "SCT_umi",
                 score_name = "basal_score_Visvader")

seurat_rna  <- AddCellTypeScore(seurat_obj = seurat_rna,
                 marker_table = ref_visvader,
                 cell_type = "Luminal H-S",
                 assay = "SCT_umi",
                 score_name = "hs_score_Visvader")

seurat_rna  <- AddCellTypeScore(seurat_obj = seurat_rna,
                 marker_table = ref_visvader,
                 cell_type = "LP",
                 assay = "SCT_umi",
                 score_name = "lp_score_Visvader")

```


```{r scores4, message=F, warning=F}
VlnPlot(seurat, c("basal_score_Landragin", "lp_score_Landragin", "hs_score_Landragin"),
                   group.by = c("spectral_rna_umi_harmony_clusters"), pt.size = 0.5,
                   cols = c("#a3215e", "#236ec4", "#F9C22EFF")) +
                   stat_summary(fun.y = median, geom='point', size = 2, colour = "black", position = position_dodge(width = 0.9))

VlnPlot(seurat, c("basal_score_Visvader", "lp_score_Visvader", "hs_score_Visvader"),
                   group.by = c("spectral_rna_umi_harmony_clusters"), pt.size = 0.5,
                   cols = c("#a3215e", "#236ec4", "#F9C22EFF")) +
                   stat_summary(fun.y = median, geom='point', size = 2, colour = "black", position = position_dodge(width = 0.9)) 

VlnPlot(seurat_rna, c("basal_score_Landragin", "lp_score_Landragin", "hs_score_Landragin"),
                   group.by = c("spectral_rna_umi_harmony_clusters"), pt.size = 0.5,
                   cols = c("#a3215e", "#236ec4", "#F9C22EFF")) +
                   stat_summary(fun.y = median, geom='point', size = 2, colour = "black", position = position_dodge(width = 0.9))

VlnPlot(seurat_rna, c("basal_score_Visvader", "lp_score_Visvader", "hs_score_Visvader"),
                   group.by = c("spectral_rna_umi_harmony_clusters"), pt.size = 0.5,
                   cols = c("#a3215e", "#236ec4", "#F9C22EFF")) +
                   stat_summary(fun.y = median, geom='point', size = 2, colour = "black", position = position_dodge(width = 0.9)) 

```

## RNA annotation
Multimodal object
```{r annot1, message=F, warning=F}
tmp_vector <- as.character(seurat$`spectral_rna_umi_harmony_clusters`)

seurat$annot_rna_joint <- dplyr::case_when(
  tmp_vector == "1" ~ "LP",
  tmp_vector == "2" ~ "basal",
  tmp_vector == "3" ~ "HS",
  TRUE ~ NA_character_
)

seurat$annot_rna_joint_alv <- dplyr::case_when(
  trimws(as.character(seurat$CellFate_RNA)) == "Alv" ~ "Alv",
  TRUE ~ as.character(seurat$annot_rna_joint)
)

```


```{r annot2, message=F, warning=F}
DimPlot(seurat, cols = cell_cols,
        reduction = "UMAP_spectral_rna_umi_harmony",
        group.by = c("annot_rna_joint", "annot_rna_joint_alv"),
        combine = TRUE)
```

RNA only object
```{r annot3, message=F, warning=F}
tmp_vector <- as.character(seurat_rna$`spectral_rna_umi_harmony_clusters`)

seurat_rna$annot_rna_joint <- dplyr::case_when(
  tmp_vector == "1" ~ "LP",
  tmp_vector == "2" ~ "basal",
  tmp_vector == "3" ~ "HS",
  TRUE ~ NA_character_
)

seurat_rna$annot_rna_joint_alv <- dplyr::case_when(
  trimws(as.character(seurat_rna$CellFate_RNA)) == "Alv" ~ "Alv",
  TRUE ~ as.character(seurat_rna$annot_rna_joint)
)

```


```{r annot4, message=F, warning=F}
DimPlot(seurat_rna, cols = cell_cols,
        reduction = "UMAP_spectral_rna_umi_harmony",
        group.by = c("annot_rna_joint", "annot_rna_joint_alv"),
        combine = TRUE)
```

# H3K4me1 clustering
```{r cluster_k4_1, message=F, warning=F}
seurat <- RunUMAP(seurat,
                  assay = "bin_5k_h3k4me1",
                  seed.use = 21,
                  dims = 1:length(colnames(seurat@reductions[["spectral_k4"]])),
                  n.neighbors = 35,
                  min.dist = 0.4,
                  spread = 2.2,
                  reduction = "spectral_k4",
                  reduction.name = "UMAP_spectral_k4")


seurat <- FindNeighbors(seurat,
                        reduction = "spectral_k4",
                        cluster.name = "spectral_k4_clusters",
                        graph.name = "spectral_k4_snn",
                        dims = 1:length(colnames(seurat@reductions[["spectral_k4"]])),
                        k.param = 20)

seurat <- FindClusters(seurat,
                       resolution = 0.05,
                       cluster.name = "spectral_k4_clusters",
                       graph.name = "spectral_k4_snn",
                       algorithm = 4)

```

```{r cluster_k4_11, message=F, warning=F}
DimPlot(seurat, cols = rep(mypal,2),
        reduction = "UMAP_spectral_k4",
        group.by = c("Exp", "spectral_k4_clusters", "annot_rna_joint_alv", "Group"),
        combine = TRUE)


FeaturePlot(seurat,
        reduction = "UMAP_spectral_k4",
        features = c("log10_fragments", "FRiP"),
        combine = TRUE)

```

## Cell type scores
Landragin
```{r scores_k4_1, message=F, warning=F}
seurat <- AddCellTypeScore(seurat_obj = seurat,
                 marker_table = ref_inhouse,
                 cell_type = "Basal",
                 assay = "ga_h3k4me1",
                 score_name = "basal_score_Landragin_k4")

seurat <- AddCellTypeScore(seurat_obj = seurat,
                 marker_table = ref_inhouse,
                 cell_type = "Luminal H-S",
                 assay = "ga_h3k4me1",
                 score_name = "hs_score_Landragin_k4")

seurat <- AddCellTypeScore(seurat_obj = seurat,
                 marker_table = ref_inhouse,
                 cell_type = "LP",
                 assay = "ga_h3k4me1",
                 score_name = "lp_score_Landragin_k4")

```

Visvader
```{r scores_k4_2, message=F, warning=F}
seurat <- AddCellTypeScore(seurat_obj = seurat,
                 marker_table = ref_visvader,
                 cell_type = "Basal",
                 assay = "ga_h3k4me1",
                 score_name = "basal_score_Visvader_k4")

seurat <- AddCellTypeScore(seurat_obj = seurat,
                 marker_table = ref_visvader,
                 cell_type = "Luminal H-S",
                 assay = "ga_h3k4me1",
                 score_name = "hs_score_Visvader_k4")

seurat <- AddCellTypeScore(seurat_obj = seurat,
                 marker_table = ref_visvader,
                 cell_type = "LP",
                 assay = "ga_h3k4me1",
                 score_name = "lp_score_Visvader_k4")

```


```{r scores_k4_3, message=F, warning=F}
VlnPlot(seurat, c("basal_score_Landragin_k4", "lp_score_Landragin_k4", "hs_score_Landragin_k4"),
                   group.by = c("spectral_k4_clusters"), pt.size = 0.2,
                   cols = c("#a3215e", "#236ec4", "#F9C22EFF")) +
                   stat_summary(fun.y = median, geom='point', size = 2, colour = "black", position = position_dodge(width = 0.9))

VlnPlot(seurat, c("basal_score_Visvader_k4", "lp_score_Visvader_k4", "hs_score_Visvader_k4"),
                   group.by = c("spectral_k4_clusters"), pt.size = 0.2,
                   cols = c("#a3215e", "#236ec4", "#F9C22EFF")) +
                   stat_summary(fun.y = median, geom='point', size = 2, colour = "black", position = position_dodge(width = 0.9)) 
```

## K4 annotation
```{r annot_k4_1, message=F, warning=F}
tmp_vector <- seurat@meta.data[["spectral_k4_clusters"]]

annot_k4_joint <- case_when(
  tmp_vector == 1 ~ "LP",
  tmp_vector == 2 ~ "basal",
  tmp_vector == 3 ~ "HS"
)

seurat@meta.data[["annot_k4_joint"]] <- annot_k4_joint 
```


```{r annot_k4_2, message=F, warning=F}
DimPlot(seurat, cols = cell_cols,
        reduction = "UMAP_spectral_k4",
        group.by = c("annot_k4_joint", "annot_rna_joint_alv"),
        combine = TRUE)
```

# H3K27me3 clustering
```{r cluster_k27_1, message=F, warning=F}
seurat <- RunUMAP(seurat,
                  assay = "bin_5k_h3k27me3",
                  seed.use = 42,
                  n.neighbors = 30,
                  min.dist = 0.1,
                  spread = 1.5,
                  dims = 1:length(colnames(seurat@reductions[["spectral_k27"]])),
                  reduction = "spectral_k27",
                  reduction.name = "UMAP_spectral_k27")
            
seurat <- FindNeighbors(seurat,
                        reduction = "spectral_k27",
                        cluster.name = "spectral_k27_clusters",
                        graph.name = "spectral_k27_snn",
                        dims = 1:length(colnames(seurat@reductions[["spectral_k27"]])),
                        k.param = 10)

seurat <- FindClusters(seurat,
                       resolution = 0.2,
                       cluster.name = "spectral_k27_clusters",
                       graph.name = "spectral_k27_snn",
                       algorithm = 4)


```


```{r cluster_k27_2, message=F, warning=F}
DimPlot(seurat, cols = mypal,
        reduction = "UMAP_spectral_k27",
        group.by = c("Exp", "spectral_k27_clusters", "annot_rna_joint_alv", "CellFate_CT_h3k27me3_50k"),
        combine = TRUE)

DimPlot(seurat, cols = mypal,
        reduction = "UMAP_spectral_k27",
        group.by = c("Exp", "spectral_k27_clusters", "annot_rna_joint_alv"),
        combine = TRUE)


FeaturePlot(seurat, 
        reduction = "UMAP_spectral_k27",
        features = c("log10_fragments", "FRiP"),
        combine = TRUE)


```

## Cell type scores
Landragin
```{r score_k27_1, message=F, warning=F}
seurat <- AddCellTypeScore(seurat_obj = seurat,
                 marker_table = ref_inhouse,
                 cell_type = "Basal",
                 assay = "ga_h3k27me3",
                 score_name = "basal_score_Landragin_k27")

seurat <- AddCellTypeScore(seurat_obj = seurat,
                 marker_table = ref_inhouse,
                 cell_type = "Luminal H-S",
                 assay = "ga_h3k27me3",
                 score_name = "hs_score_Landragin_k27")

seurat <- AddCellTypeScore(seurat_obj = seurat,
                 marker_table = ref_inhouse,
                 cell_type = "LP",
                 assay = "ga_h3k27me3",
                 score_name = "lp_score_Landragin_k27")

```

Visvader
```{r score_k27_2, message=F, warning=F}
seurat <- AddCellTypeScore(seurat_obj = seurat,
                 marker_table = ref_visvader,
                 cell_type = "Basal",
                 assay = "ga_h3k27me3",
                 score_name = "basal_score_Visvader_k27")

seurat <- AddCellTypeScore(seurat_obj = seurat,
                 marker_table = ref_visvader,
                 cell_type = "Luminal H-S",
                 assay = "ga_h3k27me3",
                 score_name = "hs_score_Visvader_k27")

seurat <- AddCellTypeScore(seurat_obj = seurat,
                 marker_table = ref_visvader,
                 cell_type = "LP",
                 assay = "ga_h3k27me3",
                 score_name = "lp_score_Visvader_k27")

```


```{r score_k27_3, message=F, warning=F}
DefaultAssay(seurat) <- "ga_h3k27me3"

VlnPlot(seurat,
        c("basal_score_Landragin_k27", "lp_score_Landragin_k27", "hs_score_Landragin_k27"),
        group.by = c("spectral_k27_clusters"),
        pt.size = 0.2, y.max = 3, 
        cols = c("#a3215e", "#236ec4", "#a3215e", "#F9C22EFF", "#236ec4")) +
         stat_summary(fun.y = median, geom='point', size = 2, colour = "black", position = position_dodge(width = 0.9))

VlnPlot(seurat, c("basal_score_Visvader_k27", "lp_score_Visvader_k27", "hs_score_Visvader_k27"),
                   group.by = c("spectral_k27_clusters"), pt.size = 0.2, y.max = 3,
                   cols = c("#a3215e", "#236ec4", "#a3215e", "#F9C22EFF", "#236ec4")) +
                   stat_summary(fun.y = median, geom='point', size = 2, colour = "black", position = position_dodge(width = 0.9))

```
1 - lp
2 - basal 
3 - lp
4 - hs
5 - basal 

## K27 annotation
```{r score_k27_4, message=F, warning=F}
tmp_vector <- seurat@meta.data[["spectral_k27_clusters"]]

annot_k27_joint <- case_when(
  tmp_vector == 1 ~ "LP",
  tmp_vector == 2 ~ "basal",
  tmp_vector == 3 ~ "LP",
  tmp_vector == 4 ~ "HS",
  tmp_vector == 5 ~ "basal",
)

seurat@meta.data[["annot_k27_joint"]] <- annot_k27_joint 
```


```{r score_k27_5, message=F, warning=F}
DimPlot(seurat, cols = cell_cols,
        reduction = "UMAP_spectral_k27",
        group.by = c("annot_k27_joint", "annot_rna_joint_alv"),
        combine = TRUE)

```

# Save common meta
```{save, message=F, warning=F, eval=F}
meta1 <- seurat@meta.data %>%
  rownames_to_column("cell")

meta2 <- seurat_rna@meta.data %>%
  rownames_to_column("cell")

meta_joint <- full_join(meta1, meta2, by = "cell", suffix = c(".seurat", ".rna"))

common_cols <- intersect(colnames(meta1), colnames(meta2))
common_cols <- setdiff(common_cols, "cell")

for (cn in common_cols) {
  col_seurat <- paste0(cn, ".seurat")
  col_rna    <- paste0(cn, ".rna")
  
  meta_joint[[cn]] <- dplyr::coalesce(meta_joint[[col_rna]], meta_joint[[col_seurat]])
}

meta_joint <- meta_joint %>%
  select(
    cell,
    all_of(common_cols),
    everything(),
    -ends_with(".seurat"),
    -ends_with(".rna")
  ) %>%
  column_to_rownames("cell")

write.csv(meta_joint, file.path(outputDir_tables, "metadata_joint_step2.csv"))
```


# Reload step 2
```{save2, message=F, warning=F, eval=F}
qsave(seurat, file.path(outputDir_objects, "seurat_step2.qs"))
qsave(seurat_rna, file.path(outputDir_objects, "seurat_rna_step2.qs"))

```

