---
title: "5.2 Diffusion and pseudotime analysis"
author: "Anna Schwager"
date: "`r Sys.Date()`"
output: html_document
---

```{r silence, include=FALSE, message=F, warning=F}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.show = "hold"
)
```

# Python setup
```{r reticulate, message=F, warning=F}

library(reticulate)
use_condaenv(
  "snap_py310",
  conda = "/Users/annaschwager/miniforge3/bin/conda",
  required = TRUE
)

```

```{python check-python, message=F, warning=F}

import sys
import numpy as np
import pandas as pd
from pathlib import Path
import scanpy as sc
import snapatac2 as snap
import skmisc

```

# R setup
```{r source, message=F, warning=F}

here::i_am("scripts/5.2_scMultiome_DiffusionPseudotime.Rmd")
mainDir = here::here()

set.seed(123)
```

```{r libraries_new, message=F, warning=F}
library(Signac)
library(Seurat)
library(harmony)
library(EnsDb.Mmusculus.v79)
library(org.Mm.eg.db)
library(tidyverse)
library(biovizBase)
library(ggsci)
library(scales)
library(pheatmap)
library(grid)
library(gridExtra)
library(rlang)
library(ggpointdensity)
library(tidytext)
library(patchwork)
library(GenomeInfoDb)
library(SingleCellExperiment)
library(ComplexHeatmap)
library(destiny)
library(ggalluvial)
library(ggrepel)
library(dplyr)
library(qs)
library(circlize)
library(GeneNMF)
library(UCell)
```

# Setting directories and input files
```{r dir, message=F, warning=F}
mainDir = here::here()
inputDir =  file.path(mainDir,"input") 
outputDir = file.path(mainDir, "output")
annotDir = file.path(mainDir, "annotation")

outputDir_local = file.path(outputDir, "epigenomic_diffusion") ; if(!file.exists(outputDir_local)){dir.create(outputDir_local)}
outputDir_objects = file.path(outputDir_local, "objects") ; if(!file.exists(outputDir_objects)){dir.create(outputDir_objects)}
outputDir_plots = file.path(outputDir_local, "plots") ; if(!file.exists(outputDir_plots)){dir.create(outputDir_plots)}
outputDir_tables = file.path(outputDir_local, "tables") ; if(!file.exists(outputDir_tables)){dir.create(outputDir_tables)}

snap_dir <- file.path(outputDir_objects, "snap_export") ; if(!file.exists(snap_dir )){dir.create(snap_dir)}
```

# Set colors
```{r col, message=F, warning=F}
mypal <- c("#4eb7b4","#F9C22EFF","#e4722c", "#c94c9d", "#5c4d7d", "#469D76FF", "#f7dcad", "#3C4B99FF", "#DF9ED4FF", "#C93F55FF", "#E59A52FF", "#EDCE79FF", "pink", "grey", "#92C051FF")

cell_cols <- c(
  basal   = "#236ec4",
  luminal = "#ac0000",
  LP      = "#a3215e",
  HS      = "#F9C22EFF",
  Alv     = "#f7dcad"
)

group_cols <- c(CreP_EARLY = "#c0d69f", 
                CreP_LATE = "#61980e",
                CreN = "#5c4d7d",
                Rosa = "#c94c9d")

```

# Load step 2 object
```{r load, message=F, warning=F}
seurat <- qread(file.path(outputDir_objects, "seurat_step2.qs"))
seurat_rna <- qread(file.path(outputDir_objects, "seurat_rna_step2.qs"))
```

# Downsampling the highly-covered exps
```{r downsample1, message=F, warning=F, eval=FALSE}
downsample_chrom_assay <- function(
  object,
  assay = "bin_5k_h3k4me1",
  sample_col = "Exp",
  samples_to_downsample,
  fractions,
  seed = 42,
  layer = "counts",
  new_meta_col = "nCount_downsampled",
  overwrite_nCount = FALSE,
  freq_col = "frequency_count",
  freq_threshold = NULL,
  freq_direction = ">",
  verbose = TRUE
) {
  if (is.null(names(fractions))) {
    stop("fractions must be a named vector")
  }
  
  if (!all(samples_to_downsample %in% names(fractions))) {
    stop(
      "Missing fractions for: ",
      paste(setdiff(samples_to_downsample, names(fractions)), collapse = ", ")
    )
  }
  
  if (!is.null(freq_threshold) && !freq_col %in% colnames(object@meta.data)) {
    stop("Metadata column not found: ", freq_col)
  }
  
  if (!freq_direction %in% c(">", ">=", "<", "<=")) {
    stop("freq_direction must be one of: '>', '>=', '<', '<='")
  }
  
  set.seed(seed)
  
  counts <- GetAssayData(object = object, assay = assay, layer = layer)
  counts <- as(counts, "dgCMatrix")
  
  assay_cells <- colnames(counts)
  meta_vec <- object[[sample_col, drop = TRUE]]
  
  if (length(meta_vec) != ncol(object)) {
    stop("Metadata column ", sample_col, " does not align with object cells")
  }
  
  for (s in samples_to_downsample) {
    p <- fractions[[s]]
    
    if (length(p) != 1 || is.na(p) || p <= 0 || p > 1) {
      stop("Invalid fraction for sample ", s, ": ", p)
    }
    
    cells_meta <- colnames(object)[meta_vec == s]
    
    if (!is.null(freq_threshold)) {
      freq_vals <- object[[freq_col, drop = TRUE]][cells_meta]
      
      keep <- switch(
        freq_direction,
        ">"  = freq_vals >  freq_threshold,
        ">=" = freq_vals >= freq_threshold,
        "<"  = freq_vals <  freq_threshold,
        "<=" = freq_vals <= freq_threshold
      )
      
      keep[is.na(keep)] <- FALSE
      cells_meta <- cells_meta[keep]
    }
    
    if (length(cells_meta) == 0) {
      warning("No cells found for sample ", s, " after applying filters")
      next
    }
    
    idx <- match(cells_meta, assay_cells)
    idx <- idx[!is.na(idx)]
    
    if (length(idx) == 0) {
      warning("No matching assay cells found for sample: ", s)
      next
    }
    
    if (verbose) {
      message(
        s, ": downsampling ", length(idx),
        " cells with p = ", p,
        if (!is.null(freq_threshold)) {
          paste0(" (", freq_col, " ", freq_direction, " ", freq_threshold, ")")
        } else {
          ""
        }
      )
    }
    
    mat <- counts[, idx, drop = FALSE]
    
    if (length(mat@x) > 0) {
      mat@x <- as.numeric(rbinom(n = length(mat@x), size = mat@x, prob = p))
      mat <- Matrix::drop0(mat)
    }
    
    counts[, idx] <- mat
  }
  
  object[[assay]] <- SetAssayData(
    object = object[[assay]],
    layer = layer,
    new.data = counts
  )
  
  ncount_ds <- Matrix::colSums(counts)
  names(ncount_ds) <- colnames(counts)
  
  md <- rep(NA_real_, ncol(object))
  names(md) <- colnames(object)
  md[names(ncount_ds)] <- ncount_ds
  
  object[[new_meta_col]] <- md
  
  if (overwrite_nCount) {
    object[[paste0("nCount_", assay)]] <- md
  }
  
  return(object)
}

```


```{r downsample2, message=F, warning=F, eval=F}
seurat_down <- downsample_chrom_assay(
  object = seurat,
  assay = "bin_5k_h3k4me1",
  sample_col = "Exp",
  samples_to_downsample = c("exp2", "exp3", "exp4", "exp5", "exp6", "exp7"),
  fractions = c(exp2 = 0.4, exp3 = 0.4, exp4 = 0.4, exp5 = 0.4, exp6 = 0.4, exp7 = 0.3),
  new_meta_col = "nCount_bin_5k_h3k4me1_downsampled",
  overwrite_nCount = FALSE,
  verbose = TRUE,
  freq_col = "frequency_count",
  freq_threshold = 4000,
  freq_direction = ">"
)

seurat_down$log10_frags_down_k4 <- log10(seurat_down$nCount_bin_5k_h3k4me1_downsampled)

seurat_down@assays[["bin_5k_h3k27me3"]] <- seurat@assays[["bin_5k_h3k27me3"]]

seurat_down <- downsample_chrom_assay(
  object = seurat_down,
  assay = "bin_5k_h3k27me3",
  sample_col = "Exp",
  samples_to_downsample = c("exp2", "exp3", "exp4", "exp5"),
  fractions = c(exp2 = 0.2, exp3 = 0.2, exp4 = 0.3, exp5 = 0.3),
  new_meta_col = "nCount_bin_5k_h3k27me3_downsampled",
  overwrite_nCount = FALSE,
  verbose = TRUE,
  freq_col = "frequency_count",
  freq_threshold = 3000,
  freq_direction = ">"
)

seurat_down$log10_frags_down_k27 <- log10(seurat_down$nCount_bin_5k_h3k27me3_downsampled)
```


```{r downsample3, message=F, warning=F, eval=F}
DefaultAssay(seurat_down) <- "bin_5k_h3k4me1"
VlnPlot(seurat_down, features = "log10_frags_down_k4", group.by = "Exp")
DefaultAssay(seurat_down) <- "bin_5k_h3k27me3"
VlnPlot(seurat_down, features = "log10_frags_down_k27", group.by = "Exp")

```

# Spectral embedding re-generation on downsampled samples
## H3K4me1
```{r spec1, message=F, warning=F, eval=F}
k4_assay <- GetAssayData(seurat_down, assay = "bin_5k_h3k4me1", layer = "counts")

Matrix::writeMM(k4_assay, file.path(snap_dir, "matrix_k4_down.mtx"))
write.table(colnames(k4_assay), file.path(snap_dir, "barcodes_k4_down.tsv"),
            quote = FALSE, sep = "\t", row.names = FALSE, col.names = FALSE)

bins_df <- data.frame(
  name  = rownames(k4_assay),
  stringsAsFactors = FALSE
)

write.table(bins_df, file.path(snap_dir, "features_k4_down.tsv"),
            quote = FALSE, sep = "\t", row.names = FALSE, col.names = TRUE)

```

```{python spec2, message=F, warning=F, eval=F}
snap_dir = Path(r.snap_dir) 
X = sc.read_mtx(snap_dir / "matrix_k4_down.mtx").X.tocsr().transpose()  
barcodes = pd.read_csv(snap_dir / "barcodes_k4_down.tsv", header=None, sep="\t")[0].values
feat = pd.read_csv(snap_dir / "features_k4_down.tsv", sep="\t")

adata_k4 = sc.AnnData(X=X)
adata_k4.obs_names = barcodes
adata_k4.var_names = feat["name"].values
adata_k4.var["chrom"] = feat["chrom"].values
adata_k4.var["start"] = feat["start"].values
adata_k4.var["end"]   = feat["end"].values

adata_k4.X = adata_k4.X.tocsr().astype(np.float64)
adata_k4
```

```{python spec3, message=F, warning=F, eval=F}
snap.pp.select_features(adata_k4, n_features = 300000)
snap.tl.spectral(adata_k4, features = 'selected', distance_metric='cosine')
```

```{r spec4, message=F, warning=F, eval=F}
emb_k4 <- as.matrix(py$adata_k4$obsm[["X_spectral"]])
rownames(emb_k4) <- py$adata_k4$obs_names$tolist()

seurat_down[["spectral_k4"]] <- CreateDimReducObject(
  embeddings = emb_k4,
  key       = "speck4_",
  assay     = "bin_5k_h3k4me1"
)

```

## H3K27me3
```{r spec5, message=F, warning=F, eval=F}
k27_assay <- GetAssayData(seurat_down, assay = "bin_5k_h3k27me3", layer = "counts")

Matrix::writeMM(k27_assay, file.path(snap_dir, "matrix_k27_down.mtx"))
write.table(colnames(k27_assay), file.path(snap_dir, "barcodes_k27_down.tsv"),
            quote = FALSE, sep = "\t", row.names = FALSE, col.names = FALSE)


bins_df <- data.frame(
  name  = rownames(k27_assay),
  stringsAsFactors = FALSE
)

write.table(bins_df, file.path(snap_dir, "features_k27_down.tsv"),
            quote = FALSE, sep = "\t", row.names = FALSE, col.names = TRUE)

```

```{python spec6, message=F, warning=F, eval=F}
snap_dir = Path(r.snap_dir) 
X = sc.read_mtx(snap_dir / "matrix_k27_down.mtx").X.tocsr().transpose()  
barcodes = pd.read_csv(snap_dir / "barcodes_k27_down.tsv", header=None, sep="\t")[0].values
feat = pd.read_csv(snap_dir / "features_k27_down.tsv", sep="\t")

adata_k27 = sc.AnnData(X=X)
adata_k27.obs_names = barcodes
adata_k27.var_names = feat["name"].values
adata_k27.var["chrom"] = feat["chrom"].values
adata_k27.var["start"] = feat["start"].values
adata_k27.var["end"]   = feat["end"].values

adata_k27.X = adata_k27.X.tocsr().astype(np.float64)
adata_k27
```

```{python spec7, message=F, warning=F, eval=F}
snap.pp.select_features(adata_k27, n_features = 300000)
snap.tl.spectral(adata_k27, features = 'selected', distance_metric='cosine')
```

```{r spec8, message=F, warning=F, eval=F}
emb_k27 <- as.matrix(py$adata_k27$obsm[["X_spectral"]])
rownames(emb_k27) <- py$adata_k27$obs_names$tolist()

seurat_down[["spectral_k27"]] <- CreateDimReducObject(
  embeddings = emb_k27,
  key       = "speck27_",
  assay     = "bin_5k_h3k27me3"
)

```


# Reload step 3 (down)
```{r save, message=F, warning=F, eval=F}
qsave(seurat_down, file.path(outputDir_objects, "seurat_down_step1.qs"))
```

```{r reload, message=F, warning=F}
seurat_down <- qread(file.path(outputDir_objects, "seurat_down_step1.qs"))
```


# H3K4me1 diffusion map and pseudotime
## Diffusion map - all cells
```{r diff1, message=F, warning=F}
X <- Seurat::Embeddings(seurat_down[["spectral_k4"]])[, 1:18]
dm <- DiffusionMap(X)

dc <- destiny::eigenvectors(dm)
dc <- as.matrix(dc)

seurat_down[["diffusion_spectral_k4"]] <- Seurat::CreateDimReducObject(
  embeddings = dc,
  key = "DIF_",
  assay = "bin_5k_h3k4me1"
)

```

```{r diff2, message=F, warning=F}

DimPlot(seurat_down, reduction = "diffusion_spectral_k4", group.by = c("Exp", "annot_rna_joint", "annot_k4_joint", "Group"), cols = mypal, dims = c(1,3))

```

## Global pseudotime 
```{r dpt1, message=F, warning=F}
dpt <- DPT(dm)
```


```{r dpt2, message=F, warning=F}
#plot(dpt, root = 1, paths_to = c(2, 3), col_by = 'branch', dcs = c(1,3))

tip_idx_branch1 <- which(dpt@tips[, 1])
tip_idx_branch1
```

```{r dpt3, message=F, warning=F}
plot(dpt, col_by = 'DPT106', dcs = c(1,3)) 
plot(dpt, col_by = 'DPT690', dcs = c(1,3)) 
plot(dpt, col_by = 'DPT9', dcs = c(1,3)) 
```

Add DPT and metadata to the object
```{r dpt4, message=F, warning=F}

dpt_global_from_lp <- dpt$DPT106
names(dpt_global_from_lp) <- rownames(destiny::eigenvectors(dm))

dpt_global_from_basal <- dpt$DPT9
names(dpt_global_from_basal) <- rownames(destiny::eigenvectors(dm))

dpt_global_from_hs <- dpt$DPT690
names(dpt_global_from_hs) <- rownames(destiny::eigenvectors(dm))


seurat_down <- Seurat::AddMetaData(
  seurat_down,
  metadata = dpt_global_from_lp,
  col.name = "DPT_K4_global_from_lp"
)

seurat_down <- Seurat::AddMetaData(
  seurat_down,
  metadata = dpt_global_from_basal,
  col.name = "DPT_K4_global_from_basal"
)

seurat_down <- Seurat::AddMetaData(
  seurat_down,
  metadata = dpt_global_from_hs,
  col.name = "DPT_K4_global_from_hs"
)

```


## Diffusion map - only LP and basal
```{r dpt5, message=F, warning=F}
s <- seurat_down
s@graphs <- list()
s@neighbors <- list()

DefaultAssay(s) <- "bin_5k_h3k4me1"

s <- subset(
  s,
  subset = !(annot_k4_joint %in% c("HS"))
)

X_BLP <- Seurat::Embeddings(s[["spectral_k4"]])[, 1:18]
dm_BLP <- DiffusionMap(X_BLP)

dc_BLP <- destiny::eigenvectors(dm_BLP)
dc_BLP <- as.matrix(dc_BLP)

seurat_down[["diffusion_spectral_k4_BLP"]] <- Seurat::CreateDimReducObject(
  embeddings = dc_BLP,
  key = "DIFBLP_",
  assay = "bin_5k_h3k4me1"
)

```

```{r dpt55, message=F, warning=F}
DimPlot(seurat_down, reduction = "diffusion_spectral_k4_BLP", group.by = c("Exp", "annot_rna_joint", "annot_k4_joint", "Group"), cols = mypal)

```

## LP-basal pseudotime
Calculate DPT
```{r dpt6, message=F, warning=F}
dpt_BLP  <- DPT(dm_BLP)
```


```{r dpt66, message=F, warning=F}
#plot(dpt_BLP, col_by = "branch", dcs = c(1, 2))
tip_idx_branch1 <- which(dpt_BLP@tips[, 1])
tip_idx_branch1
```

DPT from the tip of branch 1 (LP branch)
```{r dpt7, message=F, warning=F}
plot(dpt_BLP, col_by = 'DPT1473') # from LP intermediate
plot(dpt_BLP, col_by = 'DPT86') # from LP extreme
plot(dpt_BLP, col_by = 'DPT9') # from basal

```

Add DPT and metadata to the object
```{r dpt8, message=F, warning=F}

dpt_local_from_lp_ext <- dpt_BLP$DPT86
names(dpt_local_from_lp_ext) <- rownames(destiny::eigenvectors(dm_BLP))

dpt_local_from_lp_int <- dpt_BLP$DPT1473
names(dpt_local_from_lp_int) <- rownames(destiny::eigenvectors(dm_BLP))

dpt_local_from_basal <- dpt_BLP$DPT9
names(dpt_local_from_basal) <- rownames(destiny::eigenvectors(dm_BLP))

seurat_down <- Seurat::AddMetaData(
  seurat_down,
  metadata = dpt_local_from_lp_ext ,
  col.name = "DPT_K4_BLP_from_lp_ext"
)

seurat_down <- Seurat::AddMetaData(
  seurat_down,
  metadata = dpt_local_from_lp_int,
  col.name = "DPT_K4_BLP_from_lp_int"
)

seurat_down <- Seurat::AddMetaData(
  seurat_down,
  metadata = dpt_local_from_basal,
  col.name = "DPT_K4_BLP_from_basal"
)

```


# H3K27me3 diffusion map and pseudotime
## Diffusion map - all cells
```{r dpt9, message=F, warning=F}
X <- Seurat::Embeddings(seurat_down[["spectral_k27"]])[, 1:length(colnames(seurat@reductions[["spectral_k27"]]))]
dm <- DiffusionMap(X)

dc <- destiny::eigenvectors(dm)
dc <- as.matrix(dc)

seurat_down[["diffusion_spectral_k27"]] <- Seurat::CreateDimReducObject(
  embeddings = dc,
  key = "DIF_",
  assay = "bin_5k_h3k27me3"
)
```

```{r dpt10, message=F, warning=F}

DimPlot(seurat_down, reduction = "diffusion_spectral_k27", group.by = c("Exp", "annot_rna_joint", "annot_k27_joint", "Group"), cols = mypal, dims = c(1,3))

```

## Global pseudotime 
```{r dpt11, message=F, warning=F}
dpt <- DPT(dm)
```


```{r dpt12, message=F, warning=F}
#plot(dpt, root = 1, paths_to = c(2, 3), col_by = 'branch', dcs = c(1,3))
tip_idx_branch1 <- which(dpt@tips[, 1])
tip_idx_branch1
```

```{r dpt13, message=F, warning=F}
plot(dpt, col_by = 'DPT69', dcs = c(1,3)) 
plot(dpt, col_by = 'DPT164', dcs = c(1,3)) 
plot(dpt, col_by = 'DPT814', dcs = c(1,3)) 
```

Add DPT and metadata to the object
```{r dpt14, message=F, warning=F}
dpt_global_from_lp <- dpt$DPT814
names(dpt_global_from_lp) <- rownames(destiny::eigenvectors(dm))

dpt_global_from_hs <- dpt$DPT164
names(dpt_global_from_hs) <- rownames(destiny::eigenvectors(dm))

dpt_global_from_basal <- dpt$DPT69
names(dpt_global_from_basal) <- rownames(destiny::eigenvectors(dm))

seurat_down <- Seurat::AddMetaData(
  seurat_down,
  metadata = dpt_global_from_lp,
  col.name = "DPT_K27_global_from_lp"
)

seurat_down <- Seurat::AddMetaData(
  seurat_down,
  metadata = dpt_global_from_hs,
  col.name = "DPT_K27_global_from_hs"
)

seurat_down <- Seurat::AddMetaData(
  seurat_down,
  metadata = dpt_global_from_basal,
  col.name = "DPT_K27_global_from_basal"
)

```


## Diffusion map - only LP and basal
```{r diff_lp_b, message=F, warning=F}
s <- seurat_down
s@graphs <- list()
s@neighbors <- list()

s <- subset(
  s,
  subset = !(annot_k27_joint %in% c("HS"))
)

X_BLP <- Seurat::Embeddings(s[["spectral_k27"]])[, 1:length(colnames(seurat@reductions[["spectral_k27"]]))]
dm_BLP <- DiffusionMap(X_BLP)

dc_BLP <- destiny::eigenvectors(dm_BLP)
dc_BLP <- as.matrix(dc_BLP)

seurat_down[["diffusion_spectral_k27_BLP"]] <- Seurat::CreateDimReducObject(
  embeddings = dc_BLP,
  key = "DIFBLP_",
  assay = "bin_5k_h3k27me3"
)

```

```{r diff_lp_b2, message=F, warning=F}
DimPlot(seurat_down, reduction = "diffusion_spectral_k27_BLP", group.by = c("Exp", "annot_rna_joint", "annot_k27_joint", "Group"), cols = mypal)

```

## LP-basal pseudotime 
```{r diff_lp_b3, message=F, warning=F}
dpt_BLP  <- DPT(dm_BLP)
```


```{r diff_lp_b4, message=F, warning=F}
#plot(dpt_BLP, col_by = "branch", dcs = c(1, 2))
tip_idx_branch1 <- which(dpt_BLP@tips[, 1])
tip_idx_branch1

```

DPT from the tip of branch 1 (LP branch)
```{r diff_lp_b5, message=F, warning=F}
plot(dpt_BLP, col_by = 'DPT680') # from basal 1
plot(dpt_BLP, col_by = 'DPT879') # from LP 

```


Add DPT and metadata to the object
```{r diff_lp_b6, message=F, warning=F}
dpt_local_from_lp <- dpt_BLP$DPT879
names(dpt_local_from_lp) <- rownames(destiny::eigenvectors(dm_BLP))

dpt_local_from_basal_ext <- dpt_BLP$DPT146
names(dpt_local_from_basal) <- rownames(destiny::eigenvectors(dm_BLP))

seurat_down <- Seurat::AddMetaData(
  seurat_down,
  metadata = dpt_local_from_lp ,
  col.name = "DPT_K27_BLP_from_basal"
)

seurat_down <- Seurat::AddMetaData(
 seurat_down,
  metadata = dpt_local_from_lp ,
  col.name = "DPT_K27_BLP_from_lp"
)

```

# Reload step 4
```{r save4, message=F, warning=F, eval=F}
qsave(seurat_down, file.path(outputDir_objects, "seurat_down_step2.qs"))
```

```{r load4, message=F, warning=F}
seurat_down <- qread(file.path(outputDir_objects, "seurat_down_step2.qs"))
```

# RNA Diffusion map and pseudotime
## All cells
```{r rna_diff1, message=F, warning=F}
X <- Seurat::Embeddings(seurat_rna[["spectral_rna_umi_harmony"]])[, 1:length(colnames(seurat_rna@reductions[["spectral_rna_umi_harmony"]]))]
dm <- DiffusionMap(X)

dc <- destiny::eigenvectors(dm)
dc <- as.matrix(dc)

seurat_rna[["diffusion_spectral_rna"]] <- Seurat::CreateDimReducObject(
  embeddings = dc,
  key = "DIF_",
  assay = "RNA_umi"
)
```

```{r rna_diff2, message=F, warning=F}
DimPlot(seurat_rna, reduction = "diffusion_spectral_rna", group.by = c("annot_rna_joint", "Group"), cols = mypal)

```

```{r rna_diff22, message=F, warning=F}
dpt <- DPT(dm)
```

```{r rna_diff3, message=F, warning=F}
#plot(dpt, root = 1, paths_to = c(2, 3), col_by = 'branch')
tip_idx_branch1 <- which(dpt@tips[, 1])

tip_idx_branch1
```

```{r rna_diff4, message=F, warning=F}
plot(dpt, col_by = 'DPT169')
plot(dpt, col_by = 'DPT508') 
plot(dpt, col_by = 'DPT610') 

```

Add DPT and metadata to the object
```{r rna_diff5, message=F, warning=F}
dpt_global_lp <- dpt$DPT610
names(dpt_global_lp) <- rownames(destiny::eigenvectors(dm))

dpt_global_hs <- dpt$DPT169
names(dpt_global_hs) <- rownames(destiny::eigenvectors(dm))


seurat_rna <- Seurat::AddMetaData(
  seurat_rna,
  metadata = dpt_global_lp,
  col.name = "DPT_RNA_global_from_lp"
)

seurat_rna <- Seurat::AddMetaData(
  seurat_rna,
  metadata = dpt_global_hs,
  col.name = "DPT_RNA_global_from_hs"
)

```

## Only LP and basal

```{r rna_diff6, message=F, warning=F}

s <- seurat_rna
s@graphs <- list()
s@neighbors <- list()

s <- subset(
  s,
  subset = !(annot_rna_joint %in% c("HS"))
)

X_BLP <- Seurat::Embeddings(s[["spectral_rna_umi_harmony"]])[, 1:16]
dm_BLP <- DiffusionMap(X_BLP)

dc_BLP <- destiny::eigenvectors(dm_BLP)
dc_BLP <- as.matrix(dc_BLP)

seurat_rna[["diffusion_spectral_rna_BLP"]] <- Seurat::CreateDimReducObject(
  embeddings = dc_BLP,
  key = "DIFBLP_",
  assay = "RNA_umi"
)

```


```{r rna_diff7, message=F, warning=F}

DimPlot(seurat_rna, reduction = "diffusion_spectral_rna_BLP", group.by = c("annot_rna_joint", "Group"), cols = mypal)
DimPlot(seurat_rna, reduction = "diffusion_spectral_rna_BLP", group.by = c("annot_rna_joint", "Group"), cols = mypal, dims = c(1,3))

```


```{r rna_diff8, message=F, warning=F}
dpt_BLP  <- DPT(dm_BLP)
```


```{r rna_diff9, message=F, warning=F}
#plot(dpt_BLP, root = 1, paths_to = c(2, 3), col_by = 'branch')
tip_idx_branch1 <- which(dpt_BLP@tips[, 1])
tip_idx_branch1

```

```{r rna_diff10, message=F, warning=F}
plot(dpt_BLP, col_by = 'DPT64', dcs = c(1, 3))
plot(dpt_BLP, col_by = 'DPT1111', dcs = c(1, 3)) 
```



Add DPT and metadata to the object
```{r rna_diff11, message=F, warning=F}
dpt_local_from_basal <- dpt_BLP$DPT64
names(dpt_local_from_basal) <- rownames(destiny::eigenvectors(dm_BLP))

dpt_local_from_lp <- dpt_BLP$DPT1111
names(dpt_local_from_lp) <- rownames(destiny::eigenvectors(dm_BLP))

seurat_rna <- Seurat::AddMetaData(
  seurat_rna,
  metadata = dpt_local_from_basal,
  col.name = "DPT_RNA_BLP_from_basal"
)

seurat_rna <- Seurat::AddMetaData(
  seurat_rna,
  metadata = dpt_local_from_lp,
  col.name = "DPT_RNA_BLP_from_lp"
)

```

# Reload step 5
```{r save5, message=F, warning=F, eval=F}
qsave(seurat_rna, file.path(outputDir_objects, "seurat_rna_step3.qs"))
```

```{r reload5, message=F, warning=F, eval=F}
#| label: load_step4

seurat_rna <- qread(file.path(outputDir_objects, "seurat_rna_step3.qs"))
seurat <- qread(file.path(outputDir_objects, "seurat_step2.qs"))
seurat_down <- qread(file.path(outputDir_objects, "seurat_down_step2.qs"))
meta_marthe <- read.delim(file.path(inputDir, "meta_PT.txt")) # Marthe's metadata

```

# Add diffusion map embeddings
```{r add_emb, message=F, warning=F, eval=F}
seurat@reductions[["diffusion_spectral_k4"]] <- seurat_down@reductions[["diffusion_spectral_k4"]]
seurat@reductions[["diffusion_spectral_k27"]] <- seurat_down@reductions[["diffusion_spectral_k27"]]
seurat@reductions[["diffusion_spectral_rna"]] <- seurat_down@reductions[["diffusion_spectral_rna"]]

meta_new <- seurat_down@meta.data[colnames(seurat), , drop = FALSE]
stopifnot(identical(rownames(meta_new), colnames(seurat)))
seurat@meta.data <- meta_new

seurat_rna <- AddMetaData(seurat_rna, meta_new[, c("annot_k4_joint", "annot_k27_joint")])

```

