---
title: "5.3 scMultiome NMF"
author: "Anna Schwager"
date: "`r Sys.Date()`"
output: html_document
---

```{r silence, include=FALSE, message=F, warning=F}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.show = "hold"
)
```

# R setup

```{r source, message=F, warning=F}

here::i_am("scripts/5.3_scMultiome_NMF.Rmd")
mainDir = here::here()

set.seed(123)
```

```{r libraries_new, message=F, warning=F}
library(Signac)
library(Seurat)
library(tidyverse)
library(biovizBase)
library(scales)
library(pheatmap)
library(grid)
library(gridExtra)
library(rlang)
library(patchwork)
library(GenomeInfoDb)
library(SingleCellExperiment)
library(ComplexHeatmap)
library(destiny)
library(dplyr)
library(qs)
library(GeneNMF)
library(UCell)
library(ggpp)
library(tidytext)
library(ggpointdensity)
```

# Setting directories and input files

```{r dir}
mainDir = here::here()
inputDir =  file.path(mainDir,"input") 
outputDir = file.path(mainDir, "output")
annotDir = file.path(mainDir, "annotation")

outputDir_local = file.path(outputDir, "epigenomic_diffusion") ; if(!file.exists(outputDir_local)){dir.create(outputDir_local)}
outputDir_objects = file.path(outputDir_local, "objects") ; if(!file.exists(outputDir_objects)){dir.create(outputDir_objects)}
outputDir_plots = file.path(outputDir_local, "plots") ; if(!file.exists(outputDir_plots)){dir.create(outputDir_plots)}
outputDir_tables = file.path(outputDir_local, "tables") ; if(!file.exists(outputDir_tables)){dir.create(outputDir_tables)}

```

# Set colors

```{r col}
mypal <- c("#4eb7b4","#F9C22EFF","#e4722c", "#c94c9d", "#5c4d7d", "#469D76FF", "#f7dcad", "#3C4B99FF", "#DF9ED4FF", "#C93F55FF", "#E59A52FF", "#EDCE79FF", "pink", "grey", "#92C051FF")

cell_cols <- c(
  basal   = "#236ec4",
  luminal = "#ac0000",
  LP      = "#a3215e",
  HS      = "#F9C22EFF",
  Alv     = "#f7dcad"
)

group_cols <- c(CreP_EARLY = "#c0d69f", 
                CreP_LATE = "#61980e",
                CreN = "#5c4d7d",
                Rosa = "#c94c9d")

```

# Load objects

```{r load, message=F, warning=F}
seurat_rna <- qread(file.path(outputDir_objects, "seurat_rna_step4_new.qs"))
seurat <- qread(file.path(outputDir_objects, "seurat_step3_new.qs"))
seurat_down <- qread(file.path(outputDir_objects, "seurat_down_step3_new.qs"))

```

# NMF

Subset to CREpos only

```{r subset, message=F, warning=F}
seurat_rna$Cre <- dplyr::case_when(
  seurat_rna$Group %in% c("CreN", "Rosa") ~ "CreN",
  seurat_rna$Group %in% c("CreP_EARLY", "CreP_LATE") ~ "CreP",
  TRUE ~ NA_character_
)

seurat_crepos <- subset(seurat_rna, subset = Cre == "CreP")
```

## Program identification

Remove uncharachterised transcripts for interpretability

```{r nmf, message=F, warning=F, eval=F}
genes_keep <- rownames(seurat_crepos)

genes_keep <- genes_keep[
  !grepl("^Gm|Rik|^AC[0-9]|^AL[0-9]|^AP[0-9]|^AW[0-9]|^BC[0-9]|^CT[0-9]", genes_keep)
]
seurat_filtered <- subset(seurat_crepos, features = genes_keep)
seurat_filtered <- FindVariableFeatures(seurat_filtered, nfeatures = 3000, assay = "SCT_umi")

```

```{r nmf2, message=F, warning=F, eval=F}
seu.list <- SplitObject(seurat_filtered, split.by = "Exp")
```

```{r nmf3, message=F, warning=F, eval=F}
geneNMF.programs <- multiNMF(seu.list, assay="SCT_umi", slot="data", k=4:9, nfeatures = 900)
geneNMF.metaprograms <- getMetaPrograms(geneNMF.programs,
                                        nMP=16,
                                        weight.explained = 0.75,
                                        specificity.weight = 7,
                                        min.confidence = 0.6,
                                        max.genes= 80,
                                        hclust.method = "ward.D2")

geneNMF.metaprograms$metaprograms.metrics
```

```{r nmf4, message=F, warning=F, eval=F}
anno_colors <- c(mypal, "black", "grey")
names(anno_colors) <- names(geneNMF.metaprograms$metaprograms.genes)

ph <- plotMetaPrograms(geneNMF.metaprograms, annotation_colors = anno_colors )
```

Remove programs with bad internal consistency

```{r nmf44, message=F, warning=F, eval=F}
geneNMF.metaprograms <- dropMetaPrograms(
  geneNMF.metaprograms,
  dropMP = c("MP5","MP6","MP8", "MP11", "MP13")
)
```

## Cell MP scores

```{r nmf5, message=F, warning=F, eval=F}
mp.genes <- geneNMF.metaprograms$metaprograms.genes
seurat_filtered <- AddModuleScore_UCell(seurat_filtered, features = mp.genes, assay="SCT_umi", ncores=4, name = "")
```

## Dim reduction on NMF

```{r nmf6, message=F, warning=F, eval=F}
ndim <- 15

seurat_filtered <- runNMF(seurat_filtered, k = ndim, assay = "SCT_umi")
seurat_filtered <- RunUMAP(seurat_filtered,
                      reduction = "NMF",
                      dims=1:ndim, reduction.name = "NMF_UMAP",
                      reduction.key = "nmfUMAP_")
```

# Reload step 7

```{r nmf_save, message=F, warning=F, eval=F}
qsave(geneNMF.metaprograms, file.path(outputDir_objects, "nmf.qs"))
qsave(seurat_filtered, file.path(outputDir_objects, "seurat_rna_filtered_nmf.qs"))
```

```{r nmf_load, message=F, warning=F}
geneNMF.metaprograms <- qread(file.path(outputDir_objects, "nmf.qs"))
seurat_filtered <- qread(file.path(outputDir_objects, "seurat_rna_filtered_nmf.qs"))
mp.genes <- geneNMF.metaprograms$metaprograms.genes
```

## Final MP heatmap

```{r nmf_plot, message=F, warning=F}
program_cols <- c(
  MP1  = "#4eb7b4",
  MP2  = "#f7dcad",
  MP3  = "#faeb65",
  MP4  = "#afcf74",
  MP7  = "#d74e7b",
  MP9  =  "#d6e366", 
  MP10 =  "#a76a7f",
  MP12 = "#8c3a56",
  MP14 = "#390918", 
  MP15 = "#f9edd6",   
  MP16 = "#779c37"
)

ph <- plotMetaPrograms(geneNMF.metaprograms, annotation_colors = program_cols)
ph

geneNMF.metaprograms$metaprograms.metrics
```

```{r nmf_save1, message=F, warning=F, eval=F, include=F}
pdf(
  file.path(outputDir_plots, "nmf_heatmap_final.pdf"),
  width = 160 / 25.4,
  height = 150 / 25.4
)

ph
dev.off()
```

## Genes and weights per MP

```{r nmf_weights, message=F, warning=F}
df_weights <- purrr::imap_dfr(
  geneNMF.metaprograms$metaprograms.genes.weights,
  ~ tibble(
    gene = names(.x),
    weight = .x,
    MP = .y
  )
)

df_top <- df_weights %>%
  group_by(MP) %>%
  slice_max(weight, n = 10) %>%
  ungroup() %>%
  mutate(gene_plot = reorder_within(gene, weight, MP))

p_MP <- ggplot(df_top, aes(x = gene_plot, y = weight)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~MP, scales = "free_y") +
  scale_x_reordered() +
  theme_classic() +
  theme(
    axis.text.x = element_text(),
    axis.ticks.x = element_line()
  ) +
  labs(x = "Gene", y = "Weight")
#write.csv(df_weights, file.path(outputDir_tables, "MP_gene_weights.csv"))

p_MP 
```

```{r nmf_save2, message=F, warning=F, eval=F, include=F}
ggsave(file.path(outputDir_plots, "genes_nmf.pdf"),
       plot = p_MP  ,
       device = "pdf",
       units = "mm",
       width = 220,
       height = 150)
```

```{r nmf_vln, message=F, warning=F}
p_types <- VlnPlot(seurat_filtered, features=names(mp.genes),
                   group.by = c("annot_rna_joint_alv"), assay = "SCT_umi",
                   cols = cell_cols, pt.size = 0, ncol=6)

p_cre <- VlnPlot(seurat_filtered, features=names(mp.genes), assay = "SCT_umi",
                 group.by = c("Group"), pt.size = 0, ncol=6, cols = group_cols)

p_types
p_cre
```

```{r nmf_save3, message=F, warning=F, eval=F, include=F}
ggsave(file.path(outputDir_plots, "mps_cell_type.pdf"),
       plot = p_types  ,
       device = "pdf",
       units = "mm",
       width = 220,
       height = 150)

ggsave(file.path(outputDir_plots, "mps_cre.pdf"),
       plot = p_cre  ,
       device = "pdf",
       units = "mm",
       width = 220,
       height = 150)

```

# Functional enrichment per programm

## GSEA

```{r enrichments, message=F, warning=F}
top_p_HALLMARKS <- lapply(geneNMF.metaprograms$metaprograms.genes, function(program) {
  runGSEA(program, universe=rownames(seurat_crepos),
          category = "H",
          species = "Mus musculus",
          pval.thr = 0.05)
})


top_p_C2KEGG <- lapply(geneNMF.metaprograms$metaprograms.genes, function(program) {
  runGSEA(program, universe=rownames(seurat_crepos),
          category = "C2",
          subcategory = "CP:KEGG_LEGACY",
          species = "Mus musculus",
          pval.thr = 0.05)
})


top_p_C2WP <- lapply(geneNMF.metaprograms$metaprograms.genes, function(program) {
  runGSEA(program, universe=rownames(seurat_crepos),
          category = "C2",
          subcategory = "CP:WIKIPATHWAYS",
          species = "Mus musculus",
          pval.thr = 0.05)
})

```

```{r enrichments2, message=F, warning=F}
df_h <- purrr::imap_dfr(top_p_HALLMARKS, function(x, mp) {
  if (nrow(x) == 0) return(NULL)
  as.data.frame(x) %>%
    mutate(MP = mp)
}) %>%
  filter(padj < 0.05) %>%
  mutate(pathway_clean = gsub("^HALLMARK_", "", pathway)) %>%
  group_by(MP) %>%
  mutate(pathway_plot = reorder_within(pathway_clean, foldEnrichment, MP)) %>%
  ungroup()

df_h <- df_h %>% mutate(overlapGenes = sapply(overlapGenes, paste, collapse = ";"))
#write.csv(df_h, file.path(outputDir_tables, "MP_enrichment_H.csv"))

df_C2WP <- purrr::imap_dfr(top_p_C2WP, function(x, mp) {
  if (nrow(x) == 0) return(NULL)
  as.data.frame(x) %>%
    mutate(MP = mp)
}) %>%
  filter(padj < 0.05) %>%
  group_by(MP) %>%
  slice_min(padj, n = 10, with_ties = FALSE) %>%   # ← keep top 10
  mutate(pathway_plot = reorder_within(pathway, foldEnrichment, MP)) %>%
  ungroup()

df_C2WP <- df_C2WP %>% mutate(overlapGenes = sapply(overlapGenes, paste, collapse = ";"))
#write.csv(df_C2WP, file.path(outputDir_tables, "MP_enrichment_C2WP.csv"))

df_C2KEGG <- purrr::imap_dfr(top_p_C2KEGG, function(x, mp) {
  if (nrow(x) == 0) return(NULL)
  as.data.frame(x) %>%
    mutate(MP = mp)
}) %>%
  filter(padj < 0.05) %>%
  group_by(MP) %>%
  slice_min(padj, n = 10, with_ties = FALSE) %>%   # ← keep top 10
  mutate(pathway_plot = reorder_within(pathway, foldEnrichment, MP)) %>%
  ungroup()

df_C2KEGG <- df_C2KEGG %>% mutate(overlapGenes = sapply(overlapGenes, paste, collapse = ";"))
#write.csv(df_C2KEGG, file.path(outputDir_tables, "MP_enrichment_C2KEGG.csv"))
```

## Pathway heatmap

```{r enrichments3, message=F, warning=F}
bind_gsea_list <- function(x, source_name) {
  purrr::imap_dfr(x, function(res, mp) {
    if (nrow(res) == 0) return(NULL)
    as.data.frame(res) %>%
      mutate(
        MP = mp,
        source = source_name
      )
  })
}
```

```{r enrichments4, message=F, warning=F}
desired_order <- paste0("MP", 1:16)

df_enrich <- bind_rows(
  bind_gsea_list(top_p_HALLMARKS, "HALLMARK"),
  bind_gsea_list(top_p_C2WP, "WIKIPATHWAYS"),
  bind_gsea_list(top_p_C2KEGG, "KEGG")
) %>%
  mutate(
    padj = as.numeric(padj),
    score = -log10(padj)
  ) %>%
  filter(!is.na(padj), padj < 0.05)

df_top <- df_enrich %>%
  group_by(MP) %>%
  slice_min(padj, n = 10, with_ties = FALSE) %>%
  ungroup()

mat_df <- df_top %>%
  select(pathway, MP, score) %>%
  group_by(pathway, MP) %>%
  summarise(score = max(score), .groups = "drop") %>%
  pivot_wider(names_from = MP, values_from = score, values_fill = 0)

mat <- mat_df %>%
  column_to_rownames("pathway") %>%
  as.matrix()

row_order <- apply(mat, 1, which.max)
mat <- mat[order(row_order, rownames(mat)), , drop = FALSE]

annotation_col <- data.frame(
  metaprogram = colnames(mat),
  row.names = colnames(mat)
)

ann_colors <- list(
  metaprogram = program_cols
)

desired_order <- desired_order[desired_order %in% colnames(mat)]
mat <- mat[, desired_order]
annotation_col <- annotation_col[desired_order, , drop = FALSE]

p_all <- pheatmap(
  mat,
  cluster_rows =TRUE,
  cluster_cols = FALSE,
  color = colorRampPalette(c("white", "darkblue"))(100),
  annotation_col = annotation_col,
  annotation_colors = ann_colors,
  border_color = NA,
  fontsize_row = 8,
  fontsize_col = 10,
  main = "MP pathway enrichment"
)

p_all
```

```{r nmf_save4, message=F, warning=F, eval=F, include=F}
pdf(
  file.path(outputDir_plots, "nmf_all_enrich_padj.pdf"),
  width = 170 / 25.4,
  height = 150 / 25.4
)

p_all
dev.off()
```

# Dim reduction on NMF

```{r mimred, message=F, warning=F}
DimPlot(seurat_filtered, cols = mypal,
        reduction = "NMF_UMAP",
        group.by = c("annot_rna_joint_alv", "Group"),
        combine = TRUE)
```

# Harmonize meta's

```{r meta1, message=F, warning=F, eval=F}
meta_epi <- seurat@meta.data 
meta_rna <- seurat_rna@meta.data
meta_f <- seurat_filtered@meta.data


cols_to_add <- c("MP1","MP2","MP3","MP4","MP7","MP9","MP10","MP12","MP14","MP15","MP16")

meta_rna[, cols_to_add] <- meta_f[rownames(meta_rna), cols_to_add]
meta_epi[, cols_to_add] <- meta_f[rownames(meta_epi), cols_to_add]
```

```{r meta2, message=F, warning=F, eval=F}
seurat <- AddMetaData(seurat, meta_epi)
seurat_rna <- AddMetaData(seurat_rna, meta_rna)
```

# Reload step 8

```{r save8, message=F, warning=F, eval=F}
qsave(seurat, file.path(outputDir_objects, "seurat_step4_new.qs"))
qsave(seurat_rna, file.path(outputDir_objects, "seurat_rna_step6_new.qs"))
```

```{r load8, message=F, warning=F}
seurat <- qread(file.path(outputDir_objects, "seurat_step4_new.qs"))
seurat_rna <- qread(file.path(outputDir_objects, "seurat_rna_step6_new.qs"))
seurat_filtered <- qread(file.path(outputDir_objects, "seurat_rna_filtered_nmf.qs"))

meta_epi <- seurat@meta.data 
meta_rna <- seurat_rna@meta.data
meta_f <- seurat_filtered@meta.data

```

# Correlation of MPs with pseudotime and QC

```{r corr1, message=F, warning=F}
filter_outliers <- function(df, x, y, q = 0.99) {
  xq <- quantile(df[[x]], probs = c(1 - q, q), na.rm = TRUE)
  yq <- quantile(df[[y]], probs = c(1 - q, q), na.rm = TRUE)
  
  df %>%
    filter(
      .data[[x]] >= xq[1], .data[[x]] <= xq[2],
      .data[[y]] >= yq[1], .data[[y]] <= yq[2]
    )
}
```

```{r corr2, message=F, warning=F}
mp_vars <- c("MP1","MP2","MP3","MP4","MP7","MP9","MP10","MP12","MP14","MP15","MP16")
```

## Counts

```{r corr3, message=F, warning=F}
plots <- lapply(mp_vars, function(mp) {
  
  df_plot <- meta_f %>%
    dplyr::select(nCount_SCT_umi, !!rlang::sym(mp)) %>%
    dplyr::rename(MP = !!rlang::sym(mp)) %>%
    dplyr::filter(is.finite(nCount_SCT_umi), is.finite(MP)) %>%
    filter_outliers("nCount_SCT_umi", "MP", q = 0.99)
  
  lab <- "n < 3"
  if (nrow(df_plot) >= 3) {
    cor_res <- suppressWarnings(cor.test(df_plot$nCount_SCT_umi, df_plot$MP, method = "pearson"))
    lab <- paste0(
      "R = ", round(unname(cor_res$estimate), 3),
      "\np-val = ", format.pval(cor_res$p.value, digits = 3, eps = 1e-300)
    )
  }
  
  ggplot(df_plot, aes(x = nCount_SCT_umi, y = MP)) +
    geom_pointdensity(adjust = 1) +
    geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.8) +
    scale_color_viridis_c(option = "magma") +
    
    theme_classic() +
    theme(
      legend.position = "none",
      axis.title.x = element_text(size = 9),
      axis.title.y = element_text(size = 9),
      axis.text.x  = element_text(size = 8),
      axis.text.y  = element_text(size = 8)
    ) +
    labs(
      x = "nCount_SCT_umi",
      y = mp,
      title = paste("nCount_SCT_umi vs", mp)
    ) +
    annotate(
      "text_npc",
      npcx = 0.02, npcy = 0.98,
      label = lab,
      hjust = 0, vjust = 1,
      size = 3.2
    )
})

names(plots) <- mp_vars
patchwork::wrap_plots(plots, ncol = 4)
```

## Mito reads

```{r corr4, message=F, warning=F}
plots <- lapply(mp_vars, function(mp) {
  
  df_plot <- meta_f %>%
    dplyr::select(percent.mt, !!rlang::sym(mp)) %>%
    dplyr::rename(MP = !!rlang::sym(mp)) %>%
    dplyr::filter(is.finite(percent.mt), is.finite(MP)) %>%
    filter_outliers("percent.mt", "MP", q = 0.99)
  
  lab <- "n < 3"
  if (nrow(df_plot) >= 3) {
    cor_res <- suppressWarnings(cor.test(df_plot$percent.mt, df_plot$MP, method = "pearson"))
    lab <- paste0(
      "R = ", round(unname(cor_res$estimate), 3),
      "\np-val = ", format.pval(cor_res$p.value, digits = 3, eps = 1e-300)
    )
  }
  
  ggplot(df_plot, aes(x = percent.mt, y = MP)) +
    geom_pointdensity(adjust = 1) +
    geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.8) +
    scale_color_viridis_c(option = "magma") +
    
    theme_classic() +
    theme(
      legend.position = "none",
      axis.title.x = element_text(size = 9),
      axis.title.y = element_text(size = 9),
      axis.text.x  = element_text(size = 8),
      axis.text.y  = element_text(size = 8)
    ) +
    labs(
      x = "percent.mt",
      y = mp,
      title = paste("percent.mt vs", mp)
    ) +
    annotate(
      "text_npc",
      npcx = 0.02, npcy = 0.98,
      label = lab,
      hjust = 0, vjust = 1,
      size = 3.2
    )
})

names(plots) <- mp_vars

patchwork::wrap_plots(plots, ncol = 4)
```

## Pretumoral pseudotime

```{r corr44, message=F, warning=F}
plots <- lapply(mp_vars, function(mp) {
  
  df_plot <- meta_f %>%
    dplyr::select(pseudotime, !!rlang::sym(mp)) %>%
    dplyr::rename(MP = !!rlang::sym(mp)) %>%
    dplyr::filter(is.finite(pseudotime), is.finite(MP)) %>%
    filter_outliers("pseudotime", "MP", q = 0.999)
  
  lab <- "n < 3"
  if (nrow(df_plot) >= 3) {
    cor_res <- suppressWarnings(cor.test(df_plot$pseudotime, df_plot$MP, method = "pearson"))
    lab <- paste0(
      "R = ", round(unname(cor_res$estimate), 3),
      "\np-val = ", format.pval(cor_res$p.value, digits = 3, eps = 1e-300)
    )
  }
  
  ggplot(df_plot, aes(x = pseudotime, y = MP)) +
    geom_pointdensity(adjust = 1) +
    geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.8) +
    scale_color_viridis_c(option = "magma") +
    
    theme_classic() +
    theme(
      legend.position = "none",
      axis.title.x = element_text(size = 9),
      axis.title.y = element_text(size = 9),
      axis.text.x  = element_text(size = 8),
      axis.text.y  = element_text(size = 8)
    ) +
    labs(
      x = "Pseudotime",
      y = mp,
      title = paste("Pseudotime vs", mp)
    ) +
    annotate(
      "text_npc",
      npcx = 0.02, npcy = 0.98,
      label = lab,
      hjust = 0, vjust = 1,
      size = 3.2
    )
})

names(plots) <- mp_vars

patchwork::wrap_plots(plots, ncol = 4)
```

```{r nmf_save5, message=F, warning=F, eval=F, include=F}
pdf(
  file.path(outputDir_plots, "MPs_vs_pseudotime.pdf"),
  width = 175 / 25.4,
  height = 150 / 25.4
)

patchwork::wrap_plots(plots, ncol = 4)
dev.off()
```
